### 2024-06-14 paper/paper.tex
updated figures and added paper folder

%  LaTeX support: latex@mdpi.com 
%  For support, please attach all files needed for compiling as well as the log file, and specify your operating system, LaTeX version, and LaTeX editor.

%=================================================================
\documentclass[journal,article,submit,pdftex,moreauthors]{Definitions/mdpi}

%--------------------
% Class Options:
%--------------------
%----------
% journal
%----------
% Choose between the following MDPI journals:
% acoustics, actuators, addictions, admsci, adolescents, aerobiology, aerospace, agriculture, agriengineering, agrochemicals, agronomy, ai, air, algorithms, allergies, alloys, analytica, analytics, anatomia, animals, antibiotics, antibodies, antioxidants, applbiosci, appliedchem, appliedmath, applmech, applmicrobiol, applnano, applsci, aquacj, architecture, arm, arthropoda, arts, asc, asi, astronomy, atmosphere, atoms, audiolres, automation, axioms, bacteria, batteries, bdcc, behavsci, beverages, biochem, bioengineering, biologics, biology, biomass, biomechanics, biomed, biomedicines, biomedinformatics, biomimetics, biomolecules, biophysica, biosensors, biotech, birds, bloods, blsf, brainsci, breath, buildings, businesses, cancers, carbon, cardiogenetics, catalysts, cells, ceramics, challenges, chemengineering, chemistry, chemosensors, chemproc, children, chips, cimb, civileng, cleantechnol, climate, clinpract, clockssleep, cmd, coasts, coatings, colloids, colorants, commodities, compounds, computation, computers, condensedmatter, conservation, constrmater, cosmetics, covid, crops, cryptography, crystals, csmf, ctn, curroncol, cyber, dairy, data, ddc, dentistry, dermato, dermatopathology, designs, devices, diabetology, diagnostics, dietetics, digital, disabilities, diseases, diversity, dna, drones, dynamics, earth, ebj, ecologies, econometrics, economies, education, ejihpe, electricity, electrochem, electronicmat, electronics, encyclopedia, endocrines, energies, eng, engproc, entomology, entropy, environments, environsciproc, epidemiologia, epigenomes, est, fermentation, fibers, fintech, fire, fishes, fluids, foods, forecasting, forensicsci, forests, foundations, fractalfract, fuels, future, futureinternet, futurepharmacol, futurephys, futuretransp, galaxies, games, gases, gastroent, gastrointestdisord, gels, genealogy, genes, geographies, geohazards, geomatics, geosciences, geotechnics, geriatrics, grasses, gucdd, hazardousmatters, healthcare, hearts, hemato, hematolrep, heritage, higheredu, highthroughput, histories, horticulturae, hospitals, humanities, humans, hydrobiology, hydrogen, hydrology, hygiene, idr, ijerph, ijfs, ijgi, ijms, ijns, ijpb, ijtm, ijtpp, ime, immuno, informatics, information, infrastructures, inorganics, insects, instruments, inventions, iot, j, jal, jcdd, jcm, jcp, jcs, jcto, jdb, jeta, jfb, jfmk, jimaging, jintelligence, jlpea, jmmp, jmp, jmse, jne, jnt, jof, joitmc, jor, journalmedia, jox, jpm, jrfm, jsan, jtaer, jvd, jzbg, kidneydial, kinasesphosphatases, knowledge, land, languages, laws, life, liquids, literature, livers, logics, logistics, lubricants, lymphatics, machines, macromol, magnetism, magnetochemistry, make, marinedrugs, materials, materproc, mathematics, mca, measurements, medicina, medicines, medsci, membranes, merits, metabolites, metals, meteorology, methane, metrology, micro, microarrays, microbiolres, micromachines, microorganisms, microplastics, minerals, mining, modelling, molbank, molecules, mps, msf, mti, muscles, nanoenergyadv, nanomanufacturing,\gdef\@continuouspages{yes}} nanomaterials, ncrna, ndt, network, neuroglia, neurolint, neurosci, nitrogen, notspecified, %%nri, nursrep, nutraceuticals, nutrients, obesities, oceans, ohbm, onco, %oncopathology, optics, oral, organics, organoids, osteology, oxygen, parasites, parasitologia, particles, pathogens, pathophysiology, pediatrrep, pharmaceuticals, pharmaceutics, pharmacoepidemiology,\gdef\@ISSN{2813-0618}\gdef\@continuous pharmacy, philosophies, photochem, photonics, phycology, physchem, physics, physiologia, plants, plasma, platforms, pollutants, polymers, polysaccharides, poultry, powders, preprints, proceedings, processes, prosthesis, proteomes, psf, psych, psychiatryint, psychoactives, publications, quantumrep, quaternary, qubs, radiation, reactions, receptors, recycling, regeneration, religions, remotesensing, reports, reprodmed, resources, rheumato, risks, robotics, ruminants, safety, sci, scipharm, sclerosis, seeds, sensors, separations, sexes, signals, sinusitis, skins, smartcities, sna, societies, socsci, software, soilsystems, solar, solids, spectroscj, sports, standards, stats, std, stresses, surfaces, surgeries, suschem, sustainability, symmetry, synbio, systems, targets, taxonomy, technologies, telecom, test, textiles, thalassrep, thermo, tomography, tourismhosp, toxics, toxins, transplantology, transportation, traumacare, traumas, tropicalmed, universe, urbansci, uro, vaccines, vehicles, venereology, vetsci, vibration, virtualworlds, viruses, vision, waste, water, wem, wevj, wind, women, world, youth, zoonoticdis 
% For posting an early version of this manuscript as a preprint, you may use "preprints" as the journal. Changing "submit" to "accept" before posting will remove line numbers.

%---------
% article
%---------
% The default type of manuscript is "article", but can be replaced by: 
% abstract, addendum, article, book, bookreview, briefreport, casereport, comment, commentary, communication, conferenceproceedings, correction, conferencereport, entry, expressionofconcern, extendedabstract, datadescriptor, editorial, essay, erratum, hypothesis, interestingimage, obituary, opinion, projectreport, reply, retraction, review, perspective, protocol, shortnote, studyprotocol, systematicreview, supfile, technicalnote, viewpoint, guidelines, registeredreport, tutorial
% supfile = supplementary materials

%----------
% submit
%----------
% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted. This will only make changes to the frontpage (e.g., the logo of the journal will get visible), the headings, and the copyright information. Also, line numbering will be removed. Journal info and pagination for accepted papers will also be assigned by the Editorial Office.

%------------------
% moreauthors
%------------------
% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.

%---------
% pdftex
%---------
% The option pdftex is for use with pdfLaTeX. Remove "pdftex" for (1) compiling with LaTeX & dvi2pdf (if eps figures are used) or for (2) compiling with XeLaTeX.

%=================================================================
% MDPI internal commands - do not modify
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{2024}
\copyrightyear{2024}
%\externaleditor{Academic Editor: Firstname Lastname}
\datereceived{ } 
\daterevised{ } % Comment out if no revised date
\dateaccepted{ } 
\datepublished{ } 
%\datecorrected{} % For corrected papers: "Corrected: XXX" date in the original paper.
%\dateretracted{} % For corrected papers: "Retracted: XXX" date in the original paper.
\hreflink{https://doi.org/} % If needed use \linebreak
%\doinum{}
%\pdfoutput=1 % Uncommented for upload to arXiv.org
%\CorrStatement{yes}  % For updates

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, inputenc, calc, indentfirst, fancyhdr, graphicx, epstopdf, lastpage, ifthen, float, amsmath, amssymb, lineno, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, tabto, xcolor, colortbl, soul, multirow, microtype, tikz, totcount, changepage, attrib, upgreek, array, tabularx, pbox, ragged2e, tocloft, marginnote, marginfix, enotez, amsthm, natbib, hyperref, cleveref, scrextend, url, geometry, newfloat, caption, draftwatermark, seqsplit
% cleveref: load \crefname definitions after \begin{document}

%=================================================================
% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition, Notation, Assumption
%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).

%=================================================================
% Full title of the paper (Capitalized)
\Title{Generative Simplex Mapping: Nonlinear Endmember Extraction and Unmixing for Hyperspectral Imagery}

% MDPI internal command: Title for citation in the left column
\TitleCitation{Generative Simplex Mapping: Nonlinear Endmember Extraction and Unmixing for Hyperspectral Imagery}

% Author Orchid ID: enter ID or remove command
\newcommand{\orcidauthorA}{0000-0000-0000-000X} % Add \orcidA{} behind the author's name
%\newcommand{\orcidauthorB}{0000-0000-0000-000X} % Add \orcidB{} behind the author's name

% Authors, for the paper (add full first names)
\Author{John Waczak and David J. Lary}

%\longauthorlist{yes}

% MDPI internal command: Authors, for metadata in PDF
\AuthorNames{John Waczak and David J. Lary}

% MDPI internal command: Authors, for citation in the left column
\AuthorCitation{Waczak, J.; Lary, D.J.}
% If this is a Chicago style journal: Lastname, Firstname, Firstname Lastname, and Firstname Lastname.

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Affiliation 1; john.waczak@utdallas.edu\\
$^{2}$ \quad Affiliation 2; david.lary@utdallas.edu}

% Contact information of the corresponding author
\corres{Correspondence: e-mail@e-mail.com; Tel.: (optional; include country code; if there are multiple corresponding authors, add author initials) +xx-xxxx-xxx-xxxx (F.L.)}

% Current address and/or shared authorship
\firstnote{Current address: Affiliation.}  % Current address should not be the same as any items in the Affiliation section.
\secondnote{These authors contributed equally to this work.}
% The commands \thirdnote{} till \eighthnote{} are available for further notes

%\simplesumm{} % Simple summary

%\conference{} % An extended version of a conference paper

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{A single paragraph of about 200 words maximum. For research articles, abstracts should give a pertinent overview of the work. We strongly encourage authors to use the following style of structured abstracts, but without headings: (1) Background: place the question addressed in a broad context and highlight the purpose of the study; (2) Methods: describe briefly the main methods or treatments applied; (3) Results: summarize the article's main findings; (4) Conclusions: indicate the main conclusions or interpretations. The abstract should be an objective representation of the article, it must not contain results which are not presented and substantiated in the main text and should not exaggerate the main conclusions.}

% Keywords
\keyword{keyword 1; keyword 2; keyword 3 (List three to ten pertinent keywords specific to the article; yet reasonably common within the subject discipline.)}

% The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
%\PACS{J0101}
%\MSC{}
%\JEL{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Diversity
%\LSID{\url{http://}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Applied Sciences
%\featuredapplication{Authors are encouraged to provide a concise description of the specific application or a potential application of the work. This section is not mandatory.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Data
%\dataset{DOI number or link to the deposited data set if the data set is published separately. If the data set shall be published as a supplement to this paper, this field will be filled by the journal editors. In this case, please submit the data set as a supplement.}
%\datasetlicense{License under which the data set is made available (CC0, CC-BY, CC-BY-SA, CC-BY-NC, etc.)}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Toxins
%\keycontribution{The breakthroughs or highlights of the manuscript. Authors can write one or two sentences to describe the most important part of the paper.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Encyclopedia
%\encyclopediadef{For entry manuscripts only: please provide a brief overview of the entry title instead of an abstract.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Advances in Respiratory Medicine
%\addhighlights{yes}
%\renewcommand{\addhighlights}{%

%\noindent This is an obligatory section in “Advances in Respiratory Medicine”, whose goal is to increase the discoverability and readability of the article via search engines and other scholars. Highlights should not be a copy of the abstract, but a simple text allowing the reader to quickly and simplified find out what the article is about and what can be cited from it. Each of these parts should be devoted up to 2~bullet points.\vspace{3pt}\\
%\textbf{What are the main findings?}
% \begin{itemize}[labelsep=2.5mm,topsep=-3pt]
% \item First bullet.
% \item Second bullet.
% \end{itemize}\vspace{3pt}
%\textbf{What is the implication of the main finding?}
% \begin{itemize}[labelsep=2.5mm,topsep=-3pt]
% \item First bullet.
% \item Second bullet.
% \end{itemize}
%}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setcounter{section}{-1} %% Remove this when starting to work on the template.
\section{How to Use this Template}

The template details the sections that can be used in a manuscript. Note that the order and names of article sections may differ from the requirements of the journal (e.g., the positioning of the Materials and Methods section). Please check the instructions on the authors' page of the journal to verify the correct order and names. For any questions, please contact the editorial office of the journal or support@mdpi.com. For LaTeX-related questions please contact latex@mdpi.com.%\endnote{This is an endnote.} % To use endnotes, please un-comment \printendnotes below (before References). Only journal Laws uses \footnote.

% The order of the section titles is different for some journals. Please refer to the "Instructions for Authors” on the journal homepage.

\section{Introduction}

The introduction should briefly place the study in a broad context and highlight why it is important. It should define the purpose of the work and its significance. The current state of the research field should be reviewed carefully and key publications cited. Please highlight controversial and diverging hypotheses when necessary. Finally, briefly mention the main aim of the work and highlight the principal conclusions. As far as possible, please keep the introduction comprehensible to scientists outside your particular field of research. Citing a journal paper \cite{ref-journal}. Now citing a book reference \cite{ref-book1,ref-book2} or other reference types \cite{ref-unpublish,ref-communication,ref-proceeding}. Please use the command \citep{ref-thesis,ref-url} for the following MDPI journals, which use author--date citation: Administrative Sciences, Arts, Econometrics, Economies, Genealogy, Humanities, IJFS, Journal of Intelligence, Journalism and Media, JRFM, Languages, Laws, Religions, Risks, Social Sciences, Literature.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}

Materials and Methods should be described with sufficient details to allow others to replicate and build on published results. Please note that publication of your manuscript implicates that you must make all materials, data, computer code, and protocols associated with the publication available to readers. Please disclose at the submission stage any restrictions on the availability of materials or information. New methods and protocols should be described in detail while well-established methods can be briefly described and appropriately cited.

Research manuscripts reporting large datasets that are deposited in a publicly avail-able database should specify where the data have been deposited and provide the relevant accession numbers. If the accession numbers have not yet been obtained at the time of submission, please state that they will be provided during review. They must be provided prior to publication.

Interventionary studies involving animals or humans, and other studies require ethical approval must list the authority that provided approval and the corresponding ethical approval code.
\begin{quote}
This is an example of a quote.
\end{quote}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

This section may be divided by subheadings. It should provide a concise and precise description of the experimental results, their interpretation as well as the experimental conclusions that can be drawn.
\subsection{Subsection}
\subsubsection{Subsubsection}

Bulleted lists look like this:
\begin{itemize}
\item	First bullet;
\item	Second bullet;
\item	Third bullet.
\end{itemize}

Numbered lists can be added as follows:
\begin{enumerate}
\item	First item; 
\item	Second item;
\item	Third item.
\end{enumerate}

The text continues here.

\subsection{Figures, Tables and Schemes}

All figures and tables should be cited in the main text as Figure~\ref{fig1}, Table~\ref{tab1}, etc.

\begin{figure}[H]
\includegraphics[width=10.5 cm]{Definitions/logo-mdpi}
\caption{This is a figure. Schemes follow the same formatting. If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel. (\textbf{b}) Description of what is contained in the second panel. Figures should be placed in the main text near to the first time they are cited. A caption on a single line should be centered.\label{fig1}}
\end{figure}   
\unskip

\begin{table}[H] 
\caption{This is a table caption. Tables should be placed in the main text near to the first time they are~cited.\label{tab1}}
\begin{tabularx}{\textwidth}{CCC}
\toprule
\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
\midrule
Entry 1		& Data			& Data\\
Entry 2		& Data			& Data \textsuperscript{1}\\
\bottomrule
\end{tabularx}
\noindent{\footnotesize{\textsuperscript{1} Tables may have a footer.}}
\end{table}

The text continues here (Figure~\ref{fig2} and Table~\ref{tab2}).

% Example of a figure that spans the whole page width. The same concept works for tables, too.
\begin{figure}[H]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=15.5cm]{Definitions/logo-mdpi}
\end{adjustwidth}
\caption{This is a wide figure.\label{fig2}}
\end{figure}

\begin{table}[H]
\caption{This is a wide table.\label{tab2}}
	\begin{adjustwidth}{-\extralength}{0cm}
		\begin{tabularx}{\fulllength}{CCCC}
			\toprule
			\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}     & \textbf{Title 4}\\
			\midrule
\multirow[m]{3}{*}{Entry 1 *}	& Data			& Data			& Data\\
			  	                   & Data			& Data			& Data\\
			             	      & Data			& Data			& Data\\
                   \midrule
\multirow[m]{3}{*}{Entry 2}    & Data			& Data			& Data\\
			  	                  & Data			& Data			& Data\\
			             	     & Data			& Data			& Data\\
                   \midrule
\multirow[m]{3}{*}{Entry 3}    & Data			& Data			& Data\\
			  	                 & Data			& Data			& Data\\
			             	    & Data			& Data			& Data\\
                  \midrule
\multirow[m]{3}{*}{Entry 4}   & Data			& Data			& Data\\
			  	                 & Data			& Data			& Data\\
			             	    & Data			& Data			& Data\\
			\bottomrule
		\end{tabularx}
	\end{adjustwidth}
	\noindent{\footnotesize{* Tables may have a footer.}}
\end{table}

%\begin{listing}[H]
%\caption{Title of the listing}
%\rule{\columnwidth}{1pt}
%\raggedright Text of the listing. In font size footnotesize, small, or normalsize. Preferred format: left aligned and single spaced. Preferred border format: top border line and bottom border line.
%\rule{\columnwidth}{1pt}
%\end{listing}

Text.

Text.

\subsection{Formatting of Mathematical Components}

This is the example 1 of equation:
\begin{linenomath}
\begin{equation}
a = 1,
\end{equation}
\end{linenomath}
the text following an equation need not be a new paragraph. Please punctuate equations as regular text.
%% If the documentclass option "submit" is chosen, please insert a blank line before and after any math environment (equation and eqnarray environments). This ensures correct linenumbering. The blank line should be removed when the documentclass option is changed to "accept" because the text following an equation should not be a new paragraph.

This is the example 2 of equation:
\begin{adjustwidth}{-\extralength}{0cm}
\begin{equation}
a = b + c + d + e + f + g + h + i + j + k + l + m + n + o + p + q + r + s + t + u + v + w + x + y + z
\end{equation}
\end{adjustwidth}

% Example of a page in landscape format (with table and table footnote).
%\startlandscape
%\begin{table}[H] %% Table in wide page
%\caption{This is a very wide table.\label{tab3}}
%	\begin{tabularx}{\textwidth}{CCCC}
%		\toprule
%		\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}	& \textbf{Title 4}\\
%		\midrule
%		Entry 1		& Data			& Data			& This cell has some longer content that runs over two lines.\\
%		Entry 2		& Data			& Data			& Data\textsuperscript{1}\\
%		\bottomrule
%	\end{tabularx}
%	\begin{adjustwidth}{+\extralength}{0cm}
%		\noindent\footnotesize{\textsuperscript{1} This is a table footnote.}
%	\end{adjustwidth}
%\end{table}
%\finishlandscape

Please punctuate equations as regular text. Theorem-type environments (including propositions, lemmas, corollaries etc.) can be formatted as follows:
%% Example of a theorem:
\begin{Theorem}
Example text of a theorem.
\end{Theorem}

The text continues here. Proofs must be formatted as follows:

%% Example of a proof:
\begin{proof}[Proof of Theorem 1]
Text of the proof. Note that the phrase ``of Theorem 1'' is optional if it is clear which theorem is being referred to.
\end{proof}
The text continues here.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

Authors should discuss the results and how they can be interpreted from the perspective of previous studies and of the working hypotheses. The findings and their implications should be discussed in the broadest context possible. Future research directions may also be highlighted.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}

This section is not mandatory, but can be added to the manuscript if the discussion is unusually long or complex.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Patents}

This section is not mandatory, but may be added if there are patents resulting from the work reported in this manuscript.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{6pt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
%\supplementary{The following supporting information can be downloaded at:  \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title.}

% Only for journal Methods and Protocols:
% If you wish to submit a video article, please do so with any other supplementary material.
% \supplementary{The following supporting information can be downloaded at: \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title. A supporting video article is available at doi: link.}

% Only for journal Hardware:
% If you wish to submit a video article, please do so with any other supplementary material.
% \supplementary{The following supporting information can be downloaded at: \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title.\vspace{6pt}\\
%\begin{tabularx}{\textwidth}{lll}
%\toprule
%\textbf{Name} & \textbf{Type} & \textbf{Description} \\
%\midrule
%S1 & Python script (.py) & Script of python source code used in XX \\
%S2 & Text (.txt) & Script of modelling code used to make Figure X \\
%S3 & Text (.txt) & Raw data from experiment X \\
%S4 & Video (.mp4) & Video demonstrating the hardware in use \\
%... & ... & ... \\
%\bottomrule
%\end{tabularx}
%}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{For research articles with several authors, a short paragraph specifying their individual contributions must be provided. The following statements should be used ``Conceptualization, X.X. and Y.Y.; methodology, X.X.; software, X.X.; validation, X.X., Y.Y. and Z.Z.; formal analysis, X.X.; investigation, X.X.; resources, X.X.; data curation, X.X.; writing---original draft preparation, X.X.; writing---review and editing, X.X.; visualization, X.X.; supervision, X.X.; project administration, X.X.; funding acquisition, Y.Y. All authors have read and agreed to the published version of the manuscript.'', please turn to the  \href{http://img.mdpi.org/data/contributor-role-instruction.pdf}{CRediT taxonomy} for the term explanation. Authorship must be limited to those who have contributed substantially to the work~reported.}

\funding{Please add: ``This research received no external funding'' or ``This research was funded by NAME OF FUNDER grant number XXX.'' and  and ``The APC was funded by XXX''. Check carefully that the details given are accurate and use the standard spelling of funding agency names at \url{https://search.crossref.org/funding}, any errors may affect your future funding.}

\institutionalreview{In this section, you should add the Institutional Review Board Statement and approval number, if relevant to your study. You might choose to exclude this statement if the study did not require ethical approval. Please note that the Editorial Office might ask you for further information. Please add “The study was conducted in accordance with the Declaration of Helsinki, and approved by the Institutional Review Board (or Ethics Committee) of NAME OF INSTITUTE (protocol code XXX and date of approval).” for studies involving humans. OR “The animal study protocol was approved by the Institutional Review Board (or Ethics Committee) of NAME OF INSTITUTE (protocol code XXX and date of approval).” for studies involving animals. OR “Ethical review and approval were waived for this study due to REASON (please provide a detailed justification).” OR “Not applicable” for studies not involving humans or animals.}

\informedconsent{Any research article describing a study involving humans should contain this statement. Please add ``Informed consent was obtained from all subjects involved in the study.'' OR ``Patient consent was waived due to REASON (please provide a detailed justification).'' OR ``Not applicable'' for studies not involving humans. You might also choose to exclude this statement if the study did not involve humans.

Written informed consent for publication must be obtained from participating patients who can be identified (including by the patients themselves). Please state ``Written informed consent has been obtained from the patient(s) to publish this paper'' if applicable.}

\dataavailability{We encourage all authors of articles published in MDPI journals to share their research data. In this section, please provide details regarding where data supporting reported results can be found, including links to publicly archived datasets analyzed or generated during the study. Where no new data were created, or where data is unavailable due to privacy or ethical restrictions, a statement is still required. Suggested Data Availability Statements are available in section ``MDPI Research Data Policies'' at \url{https://www.mdpi.com/ethics}.}

% Only for journal Nursing Reports
%\publicinvolvement{Please describe how the public (patients, consumers, carers) were involved in the research. Consider reporting against the GRIPP2 (Guidance for Reporting Involvement of Patients and the Public) checklist. If the public were not involved in any aspect of the research add: ``No public involvement in any aspect of this research''.}

% Only for journal Nursing Reports
%\guidelinesstandards{Please add a statement indicating which reporting guideline was used when drafting the report. For example, ``This manuscript was drafted against the XXX (the full name of reporting guidelines and citation) for XXX (type of research) research''. A complete list of reporting guidelines can be accessed via the equator network: \url{https://www.equator-network.org/}.}

% Only for journal Nursing Reports
%\useofartificialintelligence{Please describe in detail any and all uses of artificial intelligence (AI) or AI-assisted tools used in the preparation of the manuscript. This may include, but is not limited to, language translation, language editing and grammar, or generating text. Alternatively, please state that “AI or AI-assisted tools were not used in drafting any aspect of this manuscript”.}

\acknowledgments{In this section you can acknowledge any support given which is not covered by the author contribution or funding sections. This may include administrative and technical support, or donations in kind (e.g., materials used for experiments).}

\conflictsofinterest{Declare conflicts of interest or state ``The authors declare no conflicts of interest.'' Authors must identify and declare any personal circumstances or interest that may be perceived as inappropriately influencing the representation or interpretation of reported research results. Any role of the funders in the design of the study; in the collection, analyses or interpretation of data; in the writing of the manuscript; or in the decision to publish the results must be declared in this section. If there is no role, please state ``The funders had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript; or in the decision to publish the results''.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Optional

%% Only for journal Encyclopedia
%\entrylink{The Link to this entry published on the encyclopedia platform.}

\abbreviations{Abbreviations}{
The following abbreviations are used in this manuscript:\\

\noindent 
\begin{tabular}{@{}ll}
MDPI & Multidisciplinary Digital Publishing Institute\\
DOAJ & Directory of open access journals\\
TLA & Three letter acronym\\
LD & Linear dichroism
\end{tabular}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Optional
\appendixtitles{no} % Leave argument "no" if all appendix headings stay EMPTY (then no dot is printed after "Appendix A"). If the appendix sections contain a heading then change the argument to "yes".
\appendixstart
\appendix
\section[\appendixname~\thesection]{}
\subsection[\appendixname~\thesubsection]{}
The appendix is an optional section that can contain details and data supplemental to the main text---for example, explanations of experimental details that would disrupt the flow of the main text but nonetheless remain crucial to understanding and reproducing the research shown; figures of replicates for experiments of which representative data are shown in the main text can be added here if brief, or as Supplementary Data. Mathematical proofs of results not central to the paper can be added as an appendix.

\begin{table}[H] 
\caption{This is a table caption.\label{tab5}}
\begin{tabularx}{\textwidth}{CCC}
\toprule
\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
\midrule
Entry 1		& Data			& Data\\
Entry 2		& Data			& Data\\
\bottomrule
\end{tabularx}
\end{table}

\section[\appendixname~\thesection]{}
All appendix sections must be cited in the main text. In the appendices, Figures, Tables, etc. should be labeled, starting with ``A''---e.g., Figure A1, Figure A2, etc.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{adjustwidth}{-\extralength}{0cm}
%\printendnotes[custom] % Un-comment to print a list of endnotes

\reftitle{References}

% Please provide either the correct journal abbreviation (e.g. according to the “List of Title Word Abbreviations” http://www.issn.org/services/online-services/access-to-the-ltwa/) or the full name of the journal.
% Citations and References in Supplementary files are permitted provided that they also appear in the reference list here.

%=====================================
% References, variant A: external bibliography
%=====================================
%\bibliography{your_external_BibTeX_file}

%=====================================
% References, variant B: internal bibliography
%=====================================
\begin{thebibliography}{999}
% Reference 1
\bibitem[Author1(year)]{ref-journal}
Author~1, T. The title of the cited article. {\em Journal Abbreviation} {\bf 2008}, {\em 10}, 142--149.
% Reference 2
\bibitem[Author2(year)]{ref-book1}
Author~2, L. The title of the cited contribution. In {\em The Book Title}; Editor 1, F., Editor 2, A., Eds.; Publishing House: City, Country, 2007; pp. 32--58.
% Reference 3
\bibitem[Author3(year)]{ref-book2}
Author 1, A.; Author 2, B. \textit{Book Title}, 3rd ed.; Publisher: Publisher Location, Country, 2008; pp. 154--196.
% Reference 4
\bibitem[Author4(year)]{ref-unpublish}
Author 1, A.B.; Author 2, C. Title of Unpublished Work. \textit{Abbreviated Journal Name} year, \textit{phrase indicating stage of publication (submitted; accepted; in press)}.
% Reference 5
\bibitem[Author5(year)]{ref-communication}
Author 1, A.B. (University, City, State, Country); Author 2, C. (Institute, City, State, Country). Personal communication, 2012.
% Reference 6
\bibitem[Author6(year)]{ref-proceeding}
Author 1, A.B.; Author 2, C.D.; Author 3, E.F. Title of presentation. In Proceedings of the Name of the Conference, Location of Conference, Country, Date of Conference (Day Month Year); Abstract Number (optional), Pagination (optional).
% Reference 7
\bibitem[Author7(year)]{ref-thesis}
Author 1, A.B. Title of Thesis. Level of Thesis, Degree-Granting University, Location of University, Date of Completion.
% Reference 8
\bibitem[Author8(year)]{ref-url}
Title of Site. Available online: URL (accessed on Day Month Year).
\end{thebibliography}

% If authors have biography, please use the format below
%\section*{Short Biography of Authors}
%\bio
%{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/author1.pdf}}}
%{\textbf{Firstname Lastname} Biography of first author}
%
%\bio
%{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/author2.jpg}}}
%{\textbf{Firstname Lastname} Biography of second author}

% For the MDPI journals use author-date citation, please follow the formatting guidelines on http://www.mdpi.com/authors/references
% To cite two works by the same author: \citeauthor{ref-journal-1a} (\citeyear{ref-journal-1a}, \citeyear{ref-journal-1b}). This produces: Whittaker (1967, 1975)
% To cite two works by the same author with specific pages: \citeauthor{ref-journal-3a} (\citeyear{ref-journal-3a}, p. 328; \citeyear{ref-journal-3b}, p.475). This produces: Wong (1999, p. 328; 2000, p. 475)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% for journal Sci
%\reviewreports{\\
%Reviewer 1 comments and authors’ response\\
%Reviewer 2 comments and authors’ response\\
%Reviewer 3 comments and authors’ response
%}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\PublishersNote{}
\end{adjustwidth}
\end{document}

### 2024-06-14 paper/template.tex
updated figures and added paper folder

--- 2024-06-14 paper/paper.tex
@@
 % Full title of the paper (Capitalized)
-\Title{Generative Simplex Mapping: Nonlinear Endmember Extraction and Unmixing for Hyperspectral Imagery}
+\Title{Title}
 
 % MDPI internal command: Title for citation in the left column
-\TitleCitation{Generative Simplex Mapping: Nonlinear Endmember Extraction and Unmixing for Hyperspectral Imagery}
+\TitleCitation{Title}
 
@@
 % Authors, for the paper (add full first names)
-\Author{John Waczak and David J. Lary}
+\Author{Firstname Lastname $^{1,\dagger,\ddagger}$\orcidA{}, Firstname Lastname $^{2,\ddagger}$ and Firstname Lastname $^{2,}$*}
 
@@
 % MDPI internal command: Authors, for metadata in PDF
-\AuthorNames{John Waczak and David J. Lary}
+\AuthorNames{Firstname Lastname, Firstname Lastname and Firstname Lastname}
 
 % MDPI internal command: Authors, for citation in the left column
-\AuthorCitation{Waczak, J.; Lary, D.J.}
+\AuthorCitation{Lastname, F.; Lastname, F.; Lastname, F.}
 % If this is a Chicago style journal: Lastname, Firstname, Firstname Lastname, and Firstname Lastname.
@@
 \address{%
-$^{1}$ \quad Affiliation 1; john.waczak@utdallas.edu\\
-$^{2}$ \quad Affiliation 2; david.lary@utdallas.edu}
+$^{1}$ \quad Affiliation 1; e-mail@e-mail.com\\
+$^{2}$ \quad Affiliation 2; e-mail@e-mail.com}
 

### 2024-06-14 1_synthetic-dataset.jl

@@ -816,3 +816,128 @@ fig
 save(joinpath(figpath, "abundance-fit.png"), fig)
 save(joinpath(figpath, "abundance-fit.pdf"), fig)
 
+
+
+
+
+# ---------------------------------------------
+# ------------- LNEAR MODEL w/ GSMBig ---------
+# ---------------------------------------------
+
+figpath = joinpath(savepath, "linear-uniform")
+
+
+k = 75
+m = 10
+λ = 0.001
+Nᵥ = 3
+
+n_nodes = binomial(k + Nᵥ - 2, Nᵥ - 1)
+n_rbfs = binomial(m + Nᵥ - 2, Nᵥ - 1)
+
+gsm_l = GSMBig(n_nodes=n_nodes, n_rbfs=n_rbfs, Nv=Nᵥ, λ=λ,  nonlinear=false, linear=true, bias=false, make_positive=true, tol=1e-9, nepochs=100, rand_init=false, rng=StableRNG(42))
+mach_l = machine(gsm_l, X)
+
+fit!(mach_l, verbosity=1)
+
+
+abund_l = DataFrame(MLJ.transform(mach_l, X));
+model = fitted_params(mach_l)[:gsm]
+
+rpt = report(mach_l)
+node_means = rpt[:node_data_means]
+Q = rpt[:Q]
+llhs = rpt[:llhs]
+idx_vertices = rpt[:idx_vertices]
+stdev = sqrt(rpt[:β⁻¹])
+
+
+model = fitted_params(mach_l)[:gsm]
+πk = model.πk
+Z = model.Z
+
+# plot log likelihoods
+fig = Figure();
+ax = Axis(fig[1,1], xlabel="Iteration", ylabel="Log-likelihood", yscale=log10)
+lines!(ax, 2:length(llhs), llhs[2:end], linewidth=3)
+fig
+
+save(joinpath(figpath, "llh__BIG.png"), fig)
+save(joinpath(figpath, "llh__BIG.pdf"), fig)
+
+fig = Figure();
+ax = Axis(fig[1,1], xlabel="Iteration", ylabel="Q (a.u.)", yscale=log10)
+lines!(ax, 2:length(Q), Q[2:end], linewidth=3)
+fig
+
+save(joinpath(figpath, "Q-fit__BIG.png"), fig)
+save(joinpath(figpath, "Q-fit__BIG.pdf"), fig)
+
+
+# plot endmembers for linear data
+fig = Figure();
[…]

### 2024-06-14 2_process-hysu.jl

@@ -35,7 +35,7 @@ X_all = pyconvert(Array, X_all);
 X_full = pyconvert(Array, X_full);
 
 # rescale to reflectance ∈ [0,1]
-scale_fac = 10_000.0000
+scale_fac = 10_000.0
 
 X_large = X_large/scale_fac ;
 X_small = X_small/scale_fac ;

### 2024-06-14 3_hysu-gsm.jl

@@ -38,8 +38,6 @@ end
 
 
 df = CSV.read(datapath, DataFrame)
-colnames = names(df)
-df = DataFrame(Array(df) ./ 10, colnames)
 X = Array(df)
 
 # 1. Visualize the dataset

### 2024-06-17 2_process-hysu.jl
minor update with more in step 3

@@ -61,10 +61,9 @@ fig0 = cmk.heatmap(masks["materials"][:,:,1])
 fig1 = cmk.heatmap(masks["targetsROI"])
 fig2 = cmk.heatmap(X_full[:,:,100])
 
-wavelengths = [0.417400, 0.421020, 0.424640, 0.428270, 0.431890, 0.435510, 0.439130, 0.442760, 0.446380, 0.450000, 0.453620, 0.457250, 0.460870, 0.464490, 0.468110, 0.471730, 0.475360, 0.478980, 0.482600, 0.486220, 0.489850, 0.493470, 0.497090, 0.500710, 0.504340, 0.507960, 0.511580, 0.515200, 0.518830, 0.522450, 0.526070, 0.529690, 0.533310, 0.536940, 0.540560, 0.544180, 0.547800, 0.551430, 0.555050, 0.558670, 0.562290, 0.565920, 0.569540, 0.573160, 0.576780, 0.580400, 0.584030, 0.587650, 0.591270, 0.594890, 0.598520, 0.602140, 0.605760, 0.609380, 0.613010, 0.616630, 0.620250, 0.623870, 0.627490, 0.631120, 0.634740, 0.638360, 0.641980, 0.645610, 0.649230, 0.652850, 0.656470, 0.660100, 0.663720, 0.667340, 0.670960, 0.674590, 0.678210, 0.681830, 0.685450, 0.689070, 0.692700, 0.696320, 0.699940, 0.703560, 0.707190, 0.710810, 0.714430, 0.718050, 0.721680, 0.725300, 0.728920, 0.732540, 0.736160, 0.739790, 0.743410, 0.747030, 0.750650, 0.754280, 0.757900, 0.761520, 0.765140, 0.768770, 0.772390, 0.776010, 0.779630, 0.783260, 0.786880, 0.790500, 0.794120, 0.797740, 0.801370, 0.804990, 0.808610, 0.812230, 0.815860, 0.819480, 0.823100, 0.826720, 0.830350, 0.833970, 0.837590, 0.841210, 0.844830, 0.848460, 0.852080, 0.855700, 0.859320, 0.862950, 0.866570, 0.870190, 0.873810, 0.877440, 0.881060, 0.884680, 0.888300, 0.891920, 0.895550, 0.899170, 0.902790]
[…]

### 2024-06-17 3_hysu-gsm.jl

@@ -5,6 +5,7 @@ using Distributions
 using Statistics
 using LinearAlgebra, Random
 using StableRNGs
+using JSON
 
 include("./utils/hysu.jl")
 
@@ -28,41 +29,106 @@ update_theme!(
 )
 
 
-datapath = "./data/hysu/all.csv"
-@assert ispath(datapath)
+datapath_all = "./data/hysu/all.csv"
+datapath_large = "./data/hysu/large.csv"
+datapath_small = "./data/hysu/small.csv"
+@assert ispath(datapath_all)
+@assert ispath(datapath_large)
+@assert ispath(datapath_small)
 
 figpath = "./figures/hysu"
 if !isdir(figpath)
     mkdir(figpath)
+    mkdir(joinpath(figpath, "all"))
+    mkdir(joinpath(figpath, "large"))
+    mkdir(joinpath(figpath, "small"))
 end
 
 
-df = CSV.read(datapath, DataFrame)
-X = Array(df)
+
+df_ref = CSV.read("./data/hysu/reference-spectra.csv", DataFrame)
+df_svc = CSV.read("./data/hysu/svc-spectra.csv", DataFrame)
+
+
+
+df_all = CSV.read(datapath_all, DataFrame)
+df_large = CSV.read(datapath_large, DataFrame)
+df_small = CSV.read(datapath_small, DataFrame)
+
+X_all = Array(df_all)
+X_large = Array(df_large)
+X_small = Array(df_small)
 
 # 1. Visualize the dataset
 fig = Figure();
 ax = Axis(fig[1,1], xlabel="λ", ylabel="Reflectance")
 let
-    for i ∈ axes(X, 1)
-        lines!(ax, λs, X[i,:], color=(mints_colors[3], 0.5), linewidth=1)
+    for i ∈ axes(X_all, 1)
+        lines!(ax, λs, X_all[i,:], color=(mints_colors[3], 0.5), linewidth=1)
     end
 
-    lines!(ax, λs, mean(X, dims=1)[:], color=mints_colors[2], linewidth=3)
+    lines!(ax, λs, mean(X_all, dims=1)[:], color=mints_colors[2], linewidth=3)
 end
 xlims!(ax, λs[1], λs[end])
 ylims!(ax, 0, 1)
 fig
-save(joinpath(figpath, "spectra-all.png"), fig)
[…]

### 2024-06-17 utils/hysu.jl

@@ -1,4 +1,4 @@
-λs = [0.417400, 0.421020, 0.424640, 0.428270, 0.431890, 0.435510, 0.439130, 0.442760, 0.446380, 0.450000, 0.453620, 0.457250, 0.460870, 0.464490, 0.468110, 0.471730, 0.475360, 0.478980, 0.482600, 0.486220, 0.489850, 0.493470, 0.497090, 0.500710, 0.504340, 0.507960, 0.511580, 0.515200, 0.518830, 0.522450, 0.526070, 0.529690, 0.533310, 0.536940, 0.540560, 0.544180, 0.547800, 0.551430, 0.555050, 0.558670, 0.562290, 0.565920, 0.569540, 0.573160, 0.576780, 0.580400, 0.584030, 0.587650, 0.591270, 0.594890, 0.598520, 0.602140, 0.605760, 0.609380, 0.613010, 0.616630, 0.620250, 0.623870, 0.627490, 0.631120, 0.634740, 0.638360, 0.641980, 0.645610, 0.649230, 0.652850, 0.656470, 0.660100, 0.663720, 0.667340, 0.670960, 0.674590, 0.678210, 0.681830, 0.685450, 0.689070, 0.692700, 0.696320, 0.699940, 0.703560, 0.707190, 0.710810, 0.714430, 0.718050, 0.721680, 0.725300, 0.728920, 0.732540, 0.736160, 0.739790, 0.743410, 0.747030, 0.750650, 0.754280, 0.757900, 0.761520, 0.765140, 0.768770, 0.772390, 0.776010, 0.779630, 0.783260, 0.786880, 0.790500, 0.794120, 0.797740, 0.801370, 0.804990, 0.808610, 0.812230, 0.815860, 0.819480, 0.823100, 0.826720, 0.830350, 0.833970, 0.837590, 0.841210, 0.844830, 0.848460, 0.852080, 0.855700, 0.859320, 0.862950, 0.866570, 0.870190, 0.873810, 0.877440, 0.881060, 0.884680, 0.888300, 0.891920, 0.895550, 0.899170, 0.902790]
[…]

### 2024-07-11 1_toy-rgb-dataset.jl
updating so we can get started on the paper

@@ -1,928 +1,0 @@
-using CairoMakie
-using MLJ, GenerativeTopographicMapping
-using DataFrames
-using TernaryDiagrams
-using Distributions
-using StableRNGs
-using Statistics
-using Random
-
-
-
-rng = StableRNG(42)
-
-include("./utils/makie-defaults.jl")
-set_theme!(mints_theme)
-update_theme!(
-    figure_padding=30,
-    Axis=(
-        xticklabelsize=20,
-        yticklabelsize=20,
-        xlabelsize=22,
-        ylabelsize=22,
-        titlesize=25,
-    ),
-    Colorbar=(
-        ticklabelsize=20,
-        labelsize=22
-    )
-)
-
-
-savepath = "./figures/synthetic-rgb"
-if !ispath(savepath)
-    mkpath(savepath)
-end
-
-
-
-
-# Generate toy dataset of "RGB" spectra
-λs = range(350, stop=750, length=1000)
-
-Rb = exp.(-(λs .- 460.0).^2 ./(2*(22)^2))
-Rg = exp.(-(λs .- 525.0).^2 ./(2*(28)^2))
-Rr = exp.(-(λs .- 625.0).^2 ./(2*(20)^2))
-
-Rb .= Rb./maximum(Rb)
-Rg .= Rg./maximum(Rg)
-Rr .= Rr./maximum(Rr)
-
-fig = Figure();
-ax = Axis(fig[2,1], xlabel="λ", ylabel="Reflectance");
-lr = lines!(ax, λs, Rr, color=mints_colors[2], linewidth=2)
-lg = lines!(ax, λs, Rg, color=mints_colors[1], linewidth=2)
-lb = lines!(ax, λs, Rb, color=mints_colors[3], linewidth=2)
-fig[1,1] = Legend(fig, [lr, lg, lb], ["Red Band", "Green Band", "Blue Band"], framevisible=false, orientation=:horizontal, padding=(0,0,0,0), labelsize=13, height=-5)
-xlims!(ax, λs[1], λs[end])
-
-fig
-
-save(joinpath(savepath, "rgb-orig.png"), fig)
-save(joinpath(savepath, "rgb-orig.pdf"), fig)
-
-
-
-# generate dataframe of reflectance values from combined dataset
-
-# Npoints = 10_000
-Npoints = 25_000
-α_true = 0.05 * ones(3)
[…]

### 2024-07-11 1a_usgs-setup.jl

@@ -1,0 +1,220 @@
+using DataFrames, CSV
+using Distributions
+using Random, Statistics, LinearAlgebra
+using StableRNGs
+
+
+using CairoMakie, TernaryDiagrams
+
+include("./utils/makie-defaults.jl")
+set_theme!(mints_theme)
+update_theme!(
+    figure_padding=30,
+    Axis=(
+        xticklabelsize=20,
+        yticklabelsize=20,
+        xlabelsize=22,
+        ylabelsize=22,
+        titlesize=25,
+    ),
+    Colorbar=(
+        ticklabelsize=20,
+        labelsize=22
+    )
+)
+
+
+
+rng = StableRNG(42)
+
+
+# set up paths
+
+datapath = "./data/1_usgs"
+if !ispath(datapath)
+    mkpath(datapath)
+end
+
+if !ispath(joinpath(datapath, "linear"))
+    mkpath(joinpath(datapath, "linear"))
+end
+
+if !ispath(joinpath(datapath, "bilinear"))
+    mkpath(joinpath(datapath, "bilinear"))
+end
+
+
+figpath = "./figures/1_usgs"
+if !ispath(figpath)
+    mkpath(figpath)
+end
+
+if !ispath(joinpath(figpath, "linear"))
+    mkpath(joinpath(figpath, "linear"))
+end
+
+if !ispath(joinpath(figpath, "bilinear"))
+    mkpath(joinpath(figpath, "bilinear"))
+end
+
+
+outpath = "./output/1_usgs"
+if !ispath(outpath)
+    mkpath(outpath)
+end
+
+if !ispath(joinpath(outpath, "linear"))
+    mkpath(joinpath(outpath, "linear"))
+end
+
+if !ispath(joinpath(outpath, "bilinear"))
+    mkpath(joinpath(outpath, "bilinear"))
+end
+
+
+
+
+# ---------------------------------------------------
+#
+#
+#
+# we have 2 tests to do for the USG data
+# 1. Linear Mixing
+# 2. Linear Mixing + Bilinear Mixing
+#
+#
+#
+# ---------------------------------------------------
+
+
+SNR = [0, 5, 10, 15, 20, 25, 30, 35, Inf]
+
+
+# load in source spectra
+df = CSV.read("./data/src/usgs/usgs.csv", DataFrame)
+
+λs = df.λ
[…]

### 2024-07-11 1b_usgs-fit-gsm.jl

@@ -1,0 +1,430 @@
+using MLJ, GenerativeTopographicMapping, MLJNonnegativeMatrixFactorization
+using DataFrames, CSV
+using StableRNGs
+using Random
+using JSON
+
+
+
+
+datapath = "./data/1_usgs"
+outpath = "./output/1_usgs"
+@assert ispath(datapath)
+@assert ispath(outpath)
+
+include("./utils/fit-metrics.jl")
+
+
+# load in source spectra
+df = CSV.read("./data/src/usgs/usgs.csv", DataFrame)
+λs = df.λ
+min_to_use = ["Carnallite", "Ammonium Illite", "Biotite"]
+R1 = df[:, min_to_use[1]]
+R2 = df[:, min_to_use[2]]
+R3 = df[:, min_to_use[3]]
+
+abund = CSV.read("./data/1_usgs/df_abund.csv", DataFrame)
+
+
+# ------------------------
+# 1. LINEAR MIXING
+# ------------------------
+
+dpath = joinpath(datapath, "linear")
+
+# loop through all SNR and fit GSM model
+csvs = [f for f ∈ readdir(dpath) if endswith(f, ".csv")]
+
+snrs = [split(split(splitext(csv)[1], "_")[2], "-")[2] for csv ∈ csvs]
+stds = [parse(Float64, split(split(splitext(csv)[1], "_")[3], "-")[2]) for csv ∈ csvs]
+
+
+k = 75
+λs = [0.00001, 0.0001, 0.001, 0.01, 0.1]
+Nᵥ = 3
+
+
+
+for i ∈ 1:length(csvs)
+    csv = csvs[i]
+    snr = snrs[i]
+    σnoise = stds[i]
+
+    println("\nWorking on SNR = $(snr)")
+
+    # create new folder for each SNR
+    opath = joinpath(outpath, "linear", snr)
+    if !ispath(opath)
+        mkpath(opath)
+    end
+
+    # load in data
+    df = CSV.read(joinpath(dpath, csv), DataFrame)
+
+    @assert all(Array(df) .≥ 0.0)
+
+    # DO NMF - Euclidean, no regularization
+    println("\n------- Euclidean -----\n")
+    nmf = NMF(k=Nᵥ, cost=:Euclidean, normalize_abundance=true, tol=1e-9, maxiters=1000, rng=StableRNG(42))
+    mach = machine(nmf, df)
+    fit!(mach, verbosity=1)
+
[…]

### 2024-07-11 1c_usgs-plots.jl

@@ -1,0 +1,437 @@
+using CairoMakie
+using MLJ, GenerativeTopographicMapping
+using DataFrames, CSV
+using TernaryDiagrams
+using JSON
+using Distances
+
+
+
+include("./utils/makie-defaults.jl")
+set_theme!(mints_theme)
+update_theme!(
+    figure_padding=30,
+    Axis=(
+        xticklabelsize=20,
+        yticklabelsize=20,
+        xlabelsize=22,
+        ylabelsize=22,
+        titlesize=25,
+    ),
+    Colorbar=(
+        ticklabelsize=20,
+        labelsize=22
+    )
+)
+
+datapath = "./data/1_usgs"
+outpath = "./output/1_usgs"
+figpath = "./figures/1_usgs"
+@assert ispath(datapath)
+@assert ispath(outpath)
+@assert ispath(figpath)
+
+
+# load in source spectra
+df = CSV.read("./data/src/usgs/usgs.csv", DataFrame)
+λs = df.λ
+min_to_use = ["Carnallite", "Ammonium Illite", "Biotite"]
+R1 = df[:, min_to_use[1]]
+R2 = df[:, min_to_use[2]]
+R3 = df[:, min_to_use[3]]
+
+abund = CSV.read("./data/1_usgs/df_abund.csv", DataFrame)
+
+
+# ------------------------
+# 1. LINEAR MIXING
+# ------------------------
+
+respath = joinpath(outpath, "linear")
+
+results_gsm = []
+results_gsm_big = []
+results_nmf_euc = []
+results_nmf_kl = []
+results_nmf_L21 = []
+
+for (root, dirs, files) ∈ walkdir(respath)
+    for f ∈ files
+        if endswith(f, ".json")
+            if occursin("gsm-big", f)
+                push!(results_gsm_big, JSON.parsefile(joinpath(root, f)))
+            elseif occursin("nmf-euclidean", f)
+                push!(results_nmf_euc, JSON.parsefile(joinpath(root, f)))
+            elseif occursin("nmf-kl", f)
+                push!(results_nmf_kl, JSON.parsefile(joinpath(root, f)))
[…]

### 2024-07-11 2_process-hysu.jl

@@ -1,398 +1,0 @@
-using CondaPkg
-CondaPkg.add("numpy")
-CondaPkg.add("tifffile")
-using PythonCall
-using HDF5
-using MAT
-using Images, ImageIO
-import CairoMakie as cmk
-using DataFrames, CSV
-
-tifffile = pyimport("tifffile")
-
-path_large_targets = "./data/hysu/original/HySpex/large_targets"
-path_small_targets = "./data/hysu/original/HySpex/small_targets"
-path_all_targets = "./data/hysu/original/HySpex/all_targets"
-path_full = "./data/hysu/original/HySpex/full"
-
-masks_path = joinpath(path_all_targets, "TargetsROIs", "Matlab", "masks.mat")
-
-# all pixels in all_targets that include on the 6 classes
-pixel_mask = .!(Bool.(load(joinpath(path_all_targets, "DLR_HySU_HS_all_targets_mask.png"))))
-
-size(pixel_mask)
-
-# load data from tiff files
-X_large = tifffile.tifffile.imread(joinpath(path_large_targets, "DLR_HySU_HS_large_targets.tif"))
-X_small = tifffile.tifffile.imread(joinpath(path_small_targets, "DLR_HySU_HS_small_targets.tif"))
-X_all= tifffile.tifffile.imread(joinpath(path_all_targets, "DLR_HySU_HS_all_targets.tif"))
-X_full = tifffile.tifffile.imread(joinpath(path_full, "DLR_HySU_HS_full.tif"))
-
-# convert to Julia Array
-X_large = pyconvert(Array, X_large);
-X_small = pyconvert(Array, X_small);
-X_all = pyconvert(Array, X_all);
-X_full = pyconvert(Array, X_full);
-
-# rescale to reflectance ∈ [0,1]
-scale_fac = 10_000.0
-
-X_large = X_large/scale_fac ;
-X_small = X_small/scale_fac ;
-X_all = X_all/scale_fac ;
-X_full = X_full/scale_fac ;
-
-
-# load in area masks for "all_targets"
-masks = matread(masks_path)
-masks
-
-size(masks["materials"])
-size(masks["targetsROI"])
-
-
-size(X_large)
-size(X_small)
-size(X_all)
-
-masks["materials"];
-
-fig0 = cmk.heatmap(masks["materials"][:,:,1])
[…]

### 2024-07-11 2a_robotteam-data-setup.jl

@@ -1,0 +1,174 @@
+using HDF5
+using CSV, DataFrames
+using ProgressMeter
+
+include("./utils/robot-team-config.jl")
+
+
+datapath = "./data/robot-team"
+dye_path = joinpath(datapath, "dye-test")
+sup_path = joinpath(datapath, "supervised")
+
+mkpath(datapath)
+mkpath(dye_path)
+mkpath(sup_path)
+
+
+h5_basepath = "/Users/johnwaczak/data/robot-team/processed/hsi"
+@assert ispath(h5_basepath)
+
+
+idx_900 = findfirst(wavelengths .≥ 900)
+
+
+# generate file lists
+files_dict = Dict(
+    "11-23" => Dict(
+        "no-dye" => [
+            joinpath(h5_basepath, "11-23", "Scotty_1", "Scotty_1-2.h5"),
+            joinpath(h5_basepath, "11-23", "Scotty_1", "Scotty_1-7.h5"),
+            joinpath(h5_basepath, "11-23", "Scotty_1", "Scotty_1-13.h5"),
+            joinpath(h5_basepath, "11-23", "Scotty_1", "Scotty_1-14.h5"),
+            joinpath(h5_basepath, "11-23", "Scotty_2", "Scotty_2-1.h5"),
+        ],
+        "dye" => [
+            joinpath(h5_basepath, "11-23", "Scotty_4", "Scotty_4-1.h5"),
+            joinpath(h5_basepath, "11-23", "Scotty_5", "Scotty_5-1.h5"),
+        ],
+    ),
+    "12-09" => Dict(
+        "no-dye" => [
+            joinpath(h5_basepath, "12-09", "NoDye_1", "NoDye_1-4.h5"),
+            joinpath(h5_basepath, "12-09", "NoDye_2", "NoDye_2-2.h5"),
+        ],
+        "dye" => [
+            joinpath(h5_basepath, "12-09", "Dye_1", "Dye_1-6.h5"),
+            joinpath(h5_basepath, "12-09", "Dye_2", "Dye_2-5.h5"),
+        ],
+    ),
+    "12-10" => Dict(
+        "no-dye" => [
+            joinpath(h5_basepath, "12-10", "NoDye_1", "NoDye_1-1.h5"),
+            joinpath(h5_basepath, "12-10", "NoDye_2", "NoDye_2-20.h5"),
+        ],
+        "dye" => [
+            joinpath(h5_basepath, "12-10", "Dye_1", "Dye_1-6.h5"),
[…]

### 2024-07-11 2b-robotteam-fit-gsm.jl

@@ -1,0 +1,77 @@
+using MLJ, GenerativeTopographicMapping
+using DataFrames, CSV
+using StableRNGs
+using Random
+using JSON
+
+using CairoMakie
+include("./utils/makie-defaults.jl")
+set_theme!(mints_theme)
+update_theme!(
+    figure_padding=30,
+    Axis=(
+        xticklabelsize=20,
+        yticklabelsize=20,
+        xlabelsize=22,
+        ylabelsize=22,
+        titlesize=25,
+    ),
+    Colorbar=(
+        ticklabelsize=20,
+        labelsize=22
+    )
+)
+
+
+outpath = "./output/robot-team"
+figpath = "./figures/robot-team"
+mkpath(outpath)
+mkpath(figpath)
+
+
+include("./utils/fit-metrics.jl")
+
+
+# LOAD IN  COMBINED DATASET
+readdir("./data/robot-team/dye-test")
+datapath = "./data/robot-team/df_features.csv"
+df = CSV.read(datapath, DataFrame);
+
+
+# Explore data and infer intrinsic dimensionality (i.e. # of endmembers)
+using LinearAlgebra
+
+size(Array(df))
+
+Svd = svd(Array(df) .- mean(Array(df), dims=1));
+
+σs = abs2.(Svd.S) ./ (nrow(df) - 1)
+
+variances =  σs ./ sum(σs)
+cum_var = cumsum(σs ./ sum(σs))
+
+fig = Figure();
+ax = Axis(fig[1,1], xlabel="component", ylabel="Explained Variance", xticks=(1:length(σs)));
+lines!(ax, 1:length(σs), variances, linewidth=2, linestyle=:dash)
+scatter!(ax, 1:length(σs), variances, markersize=25)
+xlims!(ax, 0.85, 10)
+ylims!(ax, -0.01, nothing)
+fig
+
+save(joinpath(figpath, "pca-variance.png"), fig)
+
+
+# Based on this, between 4-8 endmembers should be fine
+# We can try to justify this 2 different ways:
+#  1. BIC/AIC
+#  2. Distribution(s) of abundances.
+
+GSMMultUpLinear()
+GSMMultUpNonlinear()
+
+
+
+
+
+
+

### 2024-07-11 3_hysu-gsm.jl

@@ -1,476 +1,0 @@
-using CairoMakie
-using MLJ, GenerativeTopographicMapping
-using DataFrames, CSV
-using Distributions
-using Statistics
-using LinearAlgebra, Random
-using StableRNGs
-using JSON
-
-
-
-include("./utils/hysu.jl")
-rng = StableRNG(42)
-
-include("./utils/makie-defaults.jl")
-set_theme!(mints_theme)
-update_theme!(
-    figure_padding=30,
-    Axis=(
-        xticklabelsize=20,
-        yticklabelsize=20,
-        xlabelsize=22,
-        ylabelsize=22,
-        titlesize=25,
-    ),
-    Colorbar=(
-        ticklabelsize=20,
-        labelsize=22
-    )
-)
-
-
-datapath_all = "./data/hysu/all.csv"
-datapath_large = "./data/hysu/large.csv"
-datapath_small = "./data/hysu/small.csv"
-@assert ispath(datapath_all)
-@assert ispath(datapath_large)
-@assert ispath(datapath_small)
-
-figpath = "./figures/hysu"
-mkdir(figpath)
-mkdir(joinpath(figpath, "all"))
-mkdir(joinpath(figpath, "large"))
-mkdir(joinpath(figpath, "small"))
-
-
-df_ref = CSV.read("./data/hysu/reference-spectra.csv", DataFrame)
-df_svc = CSV.read("./data/hysu/svc-spectra.csv", DataFrame)
-
-
-
-df_all = CSV.read(datapath_all, DataFrame)
-df_large = CSV.read(datapath_large, DataFrame)
-df_small = CSV.read(datapath_small, DataFrame)
-
-X_all = Array(df_all)
-X_large = Array(df_large)
-X_small = Array(df_small)
-
-# 1. Visualize the dataset
-fig = Figure();
-ax = Axis(fig[1,1], xlabel="λ", ylabel="Reflectance")
-let
-    for i ∈ axes(X_all, 1)
-        lines!(ax, λs, X_all[i,:], color=(mints_colors[3], 0.5), linewidth=1)
-    end
-
-    lines!(ax, λs, mean(X_all, dims=1)[:], color=mints_colors[2], linewidth=3)
-end
-xlims!(ax, λs[1], λs[end])
-ylims!(ax, 0, 1)
[…]

### 2024-07-11 4_prepare-usgs.jl

@@ -1,133 +1,0 @@
-using CairoMakie
-using DataFrames, CSV
-using Distributions
-using Statistics
-using LinearAlgebra, Random
-using JSON
-using DataInterpolations
-
-include("./utils/makie-defaults.jl")
-set_theme!(mints_theme)
-update_theme!(
-    figure_padding=30,
-    Axis=(
-        xticklabelsize=20,
-        yticklabelsize=20,
-        xlabelsize=22,
-        ylabelsize=22,
-        titlesize=25,
-    ),
-    Colorbar=(
-        ticklabelsize=20,
-        labelsize=22
-    )
-)
-
-
-datapath = joinpath("./data/usgs/original")
-
-
-spec_dict = Dict()
-
-
-name_map = Dict(
-    "almandine" => "Almandine",
-    "ammonio-illite" => "Ammonium Illite",
-    "biotite" => "Biotite",
-    "carnallite" => "Carnallite",
-    "ilmenite" => "Ilmenite",
-    "magnetite" => "Magnetite",
-    "quartz" => "Quartz",
-    "rutile" => "Rutile",
-    "zircon" => "Zircon"
-)
-
-@assert all(ispath.(joinpath.(datapath, keys(name_map) .* "_wl.txt")))
-@assert all(ispath.(joinpath.(datapath, keys(name_map) .* "_ref.txt")))
-
-
-ref_dict = Dict()
-
-for (fname, truename) ∈ name_map
-    println(fname)
-    λ = parse.(Float64, readlines(joinpath(datapath, fname*"_wl.txt"))[2:end])
-    ref = parse.(Float64, readlines(joinpath(datapath, fname*"_ref.txt"))[2:end])
-
-    idx_good = findall(ref .≥ 0.0)
-
-    # interpolate data to desired values
-    λs = 0.5:0.005:2.5
-    itp = LinearInterpolation(ref[idx_good], λ[idx_good])
-
-    ref_dict[truename] = Dict(
-        "λ" => λs,
-        "R" => itp(λs)
-    )
-end
-
-
-
-min_names = keys(ref_dict)
-
-fig = Figure();
-ax = Axis(fig[1,1], xlabel="λ (μm)", ylabel="Reflectance");
-ls = []
-i = 1
-
-for min_name ∈ min_names
[…]

### 2024-07-11 5_usgs-gtm.jl

@@ -1,435 +1,0 @@
-using CairoMakie
-using MLJ, GenerativeTopographicMapping
-using DataFrames, CSV
-using TernaryDiagrams
-using Distributions
-using StableRNGs
-using Statistics, LinearAlgebra
-using Random
-using JSON
-
-
-rng = StableRNG(42)
-
-include("./utils/makie-defaults.jl")
-set_theme!(mints_theme)
-update_theme!(
-    figure_padding=30,
-    Axis=(
-        xticklabelsize=20,
-        yticklabelsize=20,
-        xlabelsize=22,
-        ylabelsize=22,
-        titlesize=25,
-    ),
-    Colorbar=(
-        ticklabelsize=20,
-        labelsize=22
-    )
-)
-
-
-savepath = "./figures/synthetic-usgs"
-if !ispath(savepath)
-    mkpath(savepath)
-end
-
-
-# Generate synthetic datasets
-df = CSV.read("./data/usgs/usgs.csv", DataFrame)
-
-names(df)
-
-λs = df.λ
-names(df)
-min_to_use = ["Carnallite", "Ammonium Illite", "Biotite"]  # based on original VCA paper
-# min_to_use = ["Carnallite", "Ammonium Illite", "Rutile"]  # based on original VCA paper
-R1 = df[:, min_to_use[1]]
-R2 = df[:, min_to_use[2]]
-R3 = df[:, min_to_use[3]]
-
-
-
-Npoints = 1_000
-α_true = [1/3, 1/3, 1/3]
-f_dir = Dirichlet(α_true)
-abund = rand(rng, f_dir, Npoints)
-
-
-X = zeros(Npoints, length(λs));
-X = zeros(Npoints, length(λs));
-SNR = 20.0
-
-# create dataset
-for i ∈ axes(X, 1)
-    X[i,:] .= (abund[1,i] .* R1) .+ (abund[2,i] .* R2) .+ (abund[3,i] .* R3)
-end
-
-# compute std needed to acheive desired SNR
-σnoise = sqrt(mean(X.^2)/(10^(SNR/10)))
-
-for i ∈ axes(X, 1)
-    X[i,:] .= X[i,:] .+ rand(rng, Normal(0, σnoise), length(λs))
-end
-
-
-colnames = Symbol.(["λ_" * lpad(i, 3, "0") for i ∈ 1:length(λs)])
[…]

### 2024-07-11 utils/fit-metrics.jl

@@ -1,0 +1,18 @@
+using LinearAlgebra
+using Distances
+
+# evaluate performance using spectral angle
+function spectral_angle(r1, r2)
+    return acosd(dot(r1, r2)/(norm(r1) * norm(r2)))  # return angle in degrees
+end
+
+function rmse(r1, r2)
+    # return sqrt(sum((r1 .- r2).^2)/length(r1))
+    return rmsd(r1, r2)
+end
+
+function spectral_information_divergence(r1, r2)
+    # add 10 to deal with any negative values resulting from added noise.
+    return kl_divergence(r1 .+ 10, r2 .+ 10) + kl_divergence(r2 .+ 10, r1 .+ 10)
+end
+

### 2024-07-11 utils/robot-team-config.jl

@@ -1,0 +1,103 @@
+using LaTeXStrings
+
+# details about the sensors:
+# https://www.waterprobes.com/sensor-parameters-water-quality-monitoring
+
+# dictionary of targets
+# symbol => [units, long-name, min-val]
+targets_dict = Dict(
+    :Br => ["mg/l", "Br⁻", 0.0],
+    :CDOM => ["ppb", "CDOM", 0.0],
+    :CO => ["ppb", "Crude Oil", 0.0],
+    :Ca => ["mg/l", "Ca⁺⁺", 0.0],
+    :Chl => ["μg/l", "Chlorophyll A", 0.0],
+    :ChlRed => ["μg/l", "Chlorophyll A\nwith Red Excitation", 0.0],
+    :Cl => ["mg/l", "Cl⁻", 0.0],
+    :HDO => ["mg/l", "Dissolved Oxygen", 0.0],
+    :HDO_percent => ["% Sat", "Dissolved Oxygen", 0.0],
+    :NH4 => ["mg/l-N", "NH₄⁺", 0.0],
+    :NO3 => ["mg/l-N", "NO₃⁻", 0.0],
+    :Na => ["mg/l", "Na⁺", 0.0],
+    :OB => ["ppb", "Optical Brighteners", 0.0],
+    :RefFuel => ["ppb", "Refined Fuels", 0.0],
+    :Salinity3488 => ["PSU", "Salinity"],
+    #:Salinity3489 => ["PSU", "Salinity (3489)"],
+    :Salinity3490 => ["PSU", "Salinity"],
+    :SpCond => ["μS/cm", "Conductivity"],
+    :TDS => ["mg/l", "Total Dissolved Solids", 0.0],
+    :TRYP => ["ppb", "Tryptophan", 0.0],
+    :Temp3488 => ["°C", "Temperature"],
+    :Temp3489 => ["°C", "Temperature"],
+    :Temp3490 => ["°C", "Temperature"],
+    :Turb3489 => ["FNU", "Turbidity"],
+    :Turb3488 => ["FNU", "Turbidity"],
+    :Turb3490 => ["FNU", "Turbidity"],
+    :bg => ["ppb", "Blue-Green Algae\n (Phycocyanin)", 0.0],
+    :bgm => ["ppb", "Blue-Green Algae\n (Phycoerythrin)", 0.0],
+    :pH => ["0-14", "pH", 0.0],
+    :pH_mV => ["mV", "pH"],
[…]

### 2024-07-11 1a_usgs-setup.jl
updated scripts

@@ -4,7 +4,7 @@ using Random, Statistics, LinearAlgebra
 using StableRNGs
 
 
-using CairoMakie, TernaryDiagrams
+using CairoMakie, # TernaryDiagrams
 
 include("./utils/makie-defaults.jl")
 set_theme!(mints_theme)
@@ -167,40 +167,42 @@ CSV.write(joinpath(datapath, "df_abund.csv"), df_abund)
 
 
 # Plot abundance distributions
-fig = Figure();
-ax = Axis(fig[1, 1], aspect=1);
+# fig = Figure();
+# ax = Axis(fig[1, 1], aspect=1);
 
-ternaryaxis!(
-    ax,
-    hide_triangle_labels=false,
-    hide_vertex_labels=false,
-    labelx_arrow = "Red",
-    label_fontsize=20,
-    tick_fontsize=15,
-)
+# ternaryaxis!(
+#     ax,
+#     hide_triangle_labels=false,
+#     hide_vertex_labels=false,
+#     labelx_arrow = "Red",
+#     label_fontsize=20,
+#     tick_fontsize=15,
+# )
 
-ternaryscatter!(
-    ax,
-    abund[1,:],
-    abund[2,:],
-    abund[3,:],
-    color=[CairoMakie.RGBf(abund[:,i]...) for i ∈ 1:Npoints],
-    marker=:circle,
-    markersize = 8,
-)
+# ternaryscatter!(
+#     ax,
+#     abund[1,:],
+#     abund[2,:],
+#     abund[3,:],
+#     color=[CairoMakie.RGBf(abund[:,i]...) for i ∈ 1:Npoints],
+#     marker=:circle,
+#     markersize = 8,
+# )
 
-# the triangle is drawn from (0,0) to (0.5, sqrt(3)/2) to (1,0).
-xlims!(ax, -0.2, 1.2) # to center the triangle and allow space for the labels
-ylims!(ax, -0.3, 1.1)
-hidedecorations!(ax) # to hide the axis decorations
+# # the triangle is drawn from (0,0) to (0.5, sqrt(3)/2) to (1,0).
+# xlims!(ax, -0.2, 1.2) # to center the triangle and allow space for the labels
+# ylims!(ax, -0.3, 1.1)
+# hidedecorations!(ax) # to hide the axis decorations
 
-text!(ax, Point2f(-0.01, 0.5), text=min_to_use[3], fontsize=22)
[…]

### 2024-07-11 1c_usgs-plots.jl

@@ -1,7 +1,7 @@
 using CairoMakie
 using MLJ, GenerativeTopographicMapping
 using DataFrames, CSV
-using TernaryDiagrams
+# using TernaryDiagrams
 using JSON
 using Distances
 
@@ -199,8 +199,8 @@ sort!(df_nmf_L21, :SNR, rev=true)
 
 # df_gsm_plot[:, [:SNR, :θ]]
 
-extrema(df_linear.reconst_rmse)
-extrema(df_linear.θ)
+# extrema(df_linear.reconst_rmse)
+# extrema(df_linear.θ)
 
 
 # figure out the lambda with the best θ for each SNR
@@ -393,44 +393,44 @@ save(joinpath(figpath, "linear", "extracted-endmembers.png"), fig)
 
 
 
-# visualize the distribution of abundances
-Npoints = length(A1_fit)
+# # visualize the distribution of abundances
+# Npoints = length(A1_fit)
 
-fig = Figure();
-ax = Axis(fig[1, 1], aspect=1);
+# fig = Figure();
+# ax = Axis(fig[1, 1], aspect=1);
 
-ternaryaxis!(
-    ax,
-    hide_triangle_labels=false,
-    hide_vertex_labels=false,
-    labelx_arrow = "Red",
-    label_fontsize=20,
-    tick_fontsize=15,
-)
+# ternaryaxis!(
+#     ax,
+#     hide_triangle_labels=false,
+#     hide_vertex_labels=false,
+#     labelx_arrow = "Red",
+#     label_fontsize=20,
+#     tick_fontsize=15,
+# )
 
-ternaryscatter!(
-    ax,
-    A3_fit[:],
-    A1_fit[:],
-    A2_fit[:],
-    color=[CairoMakie.RGBf(A3_fit[i], A1_fit[i], A2_fit[i]) for i ∈ 1:Npoints],
-    marker=:circle,
-    markersize = 8,
-)
+# ternaryscatter!(
+#     ax,
+#     A3_fit[:],
+#     A1_fit[:],
+#     A2_fit[:],
+#     color=[CairoMakie.RGBf(A3_fit[i], A1_fit[i], A2_fit[i]) for i ∈ 1:Npoints],
+#     marker=:circle,
+#     markersize = 8,
+# )
 
-size(abund)
+# size(abund)
 
-xlims!(ax, -0.2, 1.2) # to center the triangle and allow space for the labels
-ylims!(ax, -0.3, 1.1)
[…]

### 2024-07-11 paper/paper.tex
Updates from Overleaf

@@
 %=================================================================
-\documentclass[journal,article,submit,pdftex,moreauthors]{Definitions/mdpi}
+\documentclass[remotesensing,article,submit,pdftex,moreauthors]{Definitions/mdpi}
 
@@
 % Author Orchid ID: enter ID or remove command
-\newcommand{\orcidauthorA}{0000-0000-0000-000X} % Add \orcidA{} behind the author's name
-%\newcommand{\orcidauthorB}{0000-0000-0000-000X} % Add \orcidB{} behind the author's name
+\newcommand{\orcidauthorA}{0000-0002-5910-0183} % John
+\newcommand{\orcidauthorB}{0000-0003-4265-9543} % David
 
 % Authors, for the paper (add full first names)
-\Author{John Waczak and David J. Lary}
+\Author{John Waczak \orcidA{} and David J. Lary *\orcidB{}}
 
@@
 % Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
-\address{%
-$^{1}$ \quad Affiliation 1; john.waczak@utdallas.edu\\
-$^{2}$ \quad Affiliation 2; david.lary@utdallas.edu}
+% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
+\address[1]{%
+Hanson Center for Space Sciences, University of Texas at Dallas, Richardson, TX 75080, USA;  john.waczak@utdallas.edu (J.W.)}
 
 % Contact information of the corresponding author
-\corres{Correspondence: e-mail@e-mail.com; Tel.: (optional; include country code; if there are multiple corresponding authors, add author initials) +xx-xxxx-xxx-xxxx (F.L.)}
-
-% Current address and/or shared authorship
-\firstnote{Current address: Affiliation.}  % Current address should not be the same as any items in the Affiliation section.
-\secondnote{These authors contributed equally to this work.}
-% The commands \thirdnote{} till \eighthnote{} are available for further notes
+\corres{\hangafter=1 \hangindent=1.05em \hspace{-0.82em}Correspondence: david.lary@utdallas.edu} %; Tel.: (optional; include country code; if there are multiple corresponding authors, add author initials) +xx-xxxx-xxx-xxxx (F.L.)}
 
@@
 % Abstract (Do not insert blank lines, i.e. \\) 
-\abstract{A single paragraph of about 200 words maximum.
-For research articles, abstracts should give a pertinent overview of the work.
-We strongly encourage authors to use the following style of structured abstracts, but without headings: (1) Background: place the question addressed in a broad context and highlight the purpose of the study; (2) Methods: describe briefly the main methods or treatments applied; (3) Results: summarize the article's main findings; (4) Conclusions: indicate the main conclusions or interpretations.
-The abstract should be an objective representation of the article, it must not contain results which are not presented and substantiated in the main text and should not exaggerate the main conclusions.}
+\abstract{\textcolor{red}{UPDATE ME!}}
 
 % Keywords
-\keyword{keyword 1; keyword 2; keyword 3 (List three to ten pertinent keywords specific to the article; yet reasonably common within the subject discipline.)}
-
-% The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
-%\PACS{J0101}
-%\MSC{}
-%\JEL{}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Only for the journal Diversity
-%\LSID{\url{http://}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Only for the journal Applied Sciences
-%\featuredapplication{Authors are encouraged to provide a concise description of the specific application or a potential application of the work. This section is not mandatory.}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Only for the journal Data
-%\dataset{DOI number or link to the deposited data set if the data set is published separately.
-If the data set shall be published as a supplement to this paper, this field will be filled by the journal editors.
-In this case, please submit the data set as a supplement.}
-%\datasetlicense{License under which the data set is made available (CC0, CC-BY, CC-BY-SA, CC-BY-NC, etc.)}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Only for the journal Toxins
-%\keycontribution{The breakthroughs or highlights of the manuscript. Authors can write one or two sentences to describe the most important part of the paper.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Only for the journal Encyclopedia
-%\encyclopediadef{For entry manuscripts only: please provide a brief overview of the entry title instead of an abstract.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Only for the journal Advances in Respiratory Medicine
-%\addhighlights{yes}
-%\renewcommand{\addhighlights}{%
-
-%\noindent This is an obligatory section in “Advances in Respiratory Medicine”, whose goal is to increase the discoverability and readability of the article via search engines and other scholars.
-Highlights should not be a copy of the abstract, but a simple text allowing the reader to quickly and simplified find out what the article is about and what can be cited from it.
-Each of these parts should be devoted up to 2~bullet points.\vspace{3pt}\\
-%\textbf{What are the main findings?}
-% \begin{itemize}[labelsep=2.5mm,topsep=-3pt]
-% \item First bullet.
-% \item Second bullet.
-% \end{itemize}\vspace{3pt}
-%\textbf{What is the implication of the main finding?}
-% \begin{itemize}[labelsep=2.5mm,topsep=-3pt]
-% \item First bullet.
-% \item Second bullet.
-% \end{itemize}
-%}
+\keyword{Endmember Extraction; Spectral Unmixing; Hyperspectral Imaging; Unsupervised Machine Learning}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\setcounter{section}{-1} %% Remove this when starting to work on the template.
-\section{How to Use this Template}
-
-The template details the sections that can be used in a manuscript.
-Note that the order and names of article sections may differ from the requirements of the journal (e.g., the positioning of the Materials and Methods section).
-Please check the instructions on the authors' page of the journal to verify the correct order and names.
-For any questions, please contact the editorial office of the journal or support@mdpi.com.
-For LaTeX-related questions please contact latex@mdpi.com.%\endnote{This is an endnote.} % To use endnotes, please un-comment \printendnotes below (before References).
-Only journal Laws uses \footnote.
-
-% The order of the section titles is different for some journals. Please refer to the "Instructions for Authors” on the journal homepage.
 
@@
 
-The introduction should briefly place the study in a broad context and highlight why it is important.
-It should define the purpose of the work and its significance.
-The current state of the research field should be reviewed carefully and key publications cited.
-Please highlight controversial and diverging hypotheses when necessary.
-Finally, briefly mention the main aim of the work and highlight the principal conclusions.
-As far as possible, please keep the introduction comprehensible to scientists outside your particular field of research.
-Citing a journal paper \cite{ref-journal}.
-Now citing a book reference \cite{ref-book1,ref-book2} or other reference types \cite{ref-unpublish,ref-communication,ref-proceeding}.
-Please use the command \citep{ref-thesis,ref-url} for the following MDPI journals, which use author--date citation: Administrative Sciences, Arts, Econometrics, Economies, Genealogy, Humanities, IJFS, Journal of Intelligence, Journalism and Media, JRFM, Languages, Laws, Religions, Risks, Social Sciences, Literature.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Materials and Methods}
+1. Overview of problem
 
-Materials and Methods should be described with sufficient details to allow others to replicate and build on published results.
-Please note that publication of your manuscript implicates that you must make all materials, data, computer code, and protocols associated with the publication available to readers.
-Please disclose at the submission stage any restrictions on the availability of materials or information.
-New methods and protocols should be described in detail while well-established methods can be briefly described and appropriately cited.
+2. Traditional approaches for endmember extraction and spectral unmixing
 
-Research manuscripts reporting large datasets that are deposited in a publicly avail-able database should specify where the data have been deposited and provide the relevant accession numbers.
-If the accession numbers have not yet been obtained at the time of submission, please state that they will be provided during review.
-They must be provided prior to publication.
+3. Generative topographic mapping
 
-Interventionary studies involving animals or humans, and other studies require ethical approval must list the authority that provided approval and the corresponding ethical approval code.
-\begin{quote}
-This is an example of a quote.
-\end{quote}
+4. Proposed method: generative simplex mapping
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+5. Summary of paper structure
+
+\section{Generative Simplex Mapping}
+\subsection{Method Derivation}
+
+- multiplicative updated to guarantee non-negativity (Generalized EM procedure)
+- sum-to-one constraint is guaranteed via representation of latent space with barycentric coordinates.
+
+\subsection{Implementation Details}
+
+- Linear Method
+
+- Nonlinear Method
+
+- Mixed Method
+
+\section{Experiments}
+\subsection{Linear Mixing: Comparison to NMF}
+- describe USGS data
+- describe NMF method (for comparisson)
+- describe evaluation criteria
+
+\subsection{Nonlinear Mixing: Water Contaminant Identification}
+- describe robot team data collection & processing
+- describe
+
 \section{Results}
+\subsection{Linear Mixing}
+\subsection{Nonlinear Mixing: Rhodamine Dye Plume}
+\subsection{Nonlinear Mixing: Semi-supervised Analysis}
 
-This section may be divided by subheadings.
-It should provide a concise and precise description of the experimental results, their interpretation as well as the experimental conclusions that can be drawn.
-\subsection{Subsection}
-\subsubsection{Subsubsection}
-
-Bulleted lists look like this:
-\begin{itemize}
-\item	First bullet;
-\item	Second bullet;
-\item	Third bullet.
-\end{itemize}
-
-Numbered lists can be added as follows:
-\begin{enumerate}
-\item	First item; 
-\item	Second item;
-\item	Third item.
-\end{enumerate}
-
-The text continues here.
-
-\subsection{Figures, Tables and Schemes}
-
-All figures and tables should be cited in the main text as Figure~\ref{fig1}, Table~\ref{tab1}, etc.
-
-\begin{figure}[H]
-\includegraphics[width=10.5 cm]{Definitions/logo-mdpi}
-\caption{This is a figure.
-Schemes follow the same formatting.
-If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel.
-(\textbf{b}) Description of what is contained in the second panel.
-Figures should be placed in the main text near to the first time they are cited.
-A caption on a single line should be centered.\label{fig1}}
-\end{figure}   
-\unskip
-
-\begin{table}[H] 
-\caption{This is a table caption. Tables should be placed in the main text near to the first time they are~cited.\label{tab1}}
-\begin{tabularx}{\textwidth}{CCC}
-\toprule
-\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
-\midrule
-Entry 1		& Data			& Data\\
-Entry 2		& Data			& Data \textsuperscript{1}\\
-\bottomrule
-\end{tabularx}
-\noindent{\footnotesize{\textsuperscript{1} Tables may have a footer.}}
-\end{table}
-
-The text continues here (Figure~\ref{fig2} and Table~\ref{tab2}).
-
-% Example of a figure that spans the whole page width. The same concept works for tables, too.
-\begin{figure}[H]
-\begin{adjustwidth}{-\extralength}{0cm}
-\centering
-\includegraphics[width=15.5cm]{Definitions/logo-mdpi}
-\end{adjustwidth}
-\caption{This is a wide figure.\label{fig2}}
-\end{figure}
-
-\begin{table}[H]
-\caption{This is a wide table.\label{tab2}}
-	\begin{adjustwidth}{-\extralength}{0cm}
-		\begin{tabularx}{\fulllength}{CCCC}
-			\toprule
-			\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}     & \textbf{Title 4}\\
-			\midrule
-\multirow[m]{3}{*}{Entry 1 *}	& Data			& Data			& Data\\
-			  	                   & Data			& Data			& Data\\
-			             	      & Data			& Data			& Data\\
-                   \midrule
-\multirow[m]{3}{*}{Entry 2}    & Data			& Data			& Data\\
-			  	                  & Data			& Data			& Data\\
-			             	     & Data			& Data			& Data\\
-                   \midrule
-\multirow[m]{3}{*}{Entry 3}    & Data			& Data			& Data\\
-			  	                 & Data			& Data			& Data\\
-			             	    & Data			& Data			& Data\\
-                  \midrule
-\multirow[m]{3}{*}{Entry 4}   & Data			& Data			& Data\\
-			  	                 & Data			& Data			& Data\\
-			             	    & Data			& Data			& Data\\
-			\bottomrule
-		\end{tabularx}
-	\end{adjustwidth}
-	\noindent{\footnotesize{* Tables may have a footer.}}
-\end{table}
-
-%\begin{listing}[H]
-%\caption{Title of the listing}
-%\rule{\columnwidth}{1pt}
-%\raggedright Text of the listing. In font size footnotesize, small, or normalsize. Preferred format: left aligned and single spaced. Preferred border format: top border line and bottom border line.
-%\rule{\columnwidth}{1pt}
-%\end{listing}
-
-Text.
-
-Text.
-
-\subsection{Formatting of Mathematical Components}
-
-This is the example 1 of equation:
-\begin{linenomath}
-\begin{equation}
-a = 1,
-\end{equation}
-\end{linenomath}
-the text following an equation need not be a new paragraph. Please punctuate equations as regular text.
-%% If the documentclass option "submit" is chosen, please insert a blank line before and after any math environment (equation and eqnarray environments).
-This ensures correct linenumbering.
-The blank line should be removed when the documentclass option is changed to "accept" because the text following an equation should not be a new paragraph.
-
-This is the example 2 of equation:
-\begin{adjustwidth}{-\extralength}{0cm}
-\begin{equation}
-a = b + c + d + e + f + g + h + i + j + k + l + m + n + o + p + q + r + s + t + u + v + w + x + y + z
-\end{equation}
-\end{adjustwidth}
-
-% Example of a page in landscape format (with table and table footnote).
-%\startlandscape
-%\begin{table}[H] %% Table in wide page
-%\caption{This is a very wide table.\label{tab3}}
-%	\begin{tabularx}{\textwidth}{CCCC}
-%		\toprule
-%		\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}	& \textbf{Title 4}\\
-%		\midrule
-%		Entry 1		& Data			& Data			& This cell has some longer content that runs over two lines.\\
-%		Entry 2		& Data			& Data			& Data\textsuperscript{1}\\
-%		\bottomrule
-%	\end{tabularx}
-%	\begin{adjustwidth}{+\extralength}{0cm}
-%		\noindent\footnotesize{\textsuperscript{1} This is a table footnote.}
-%	\end{adjustwidth}
-%\end{table}
-%\finishlandscape
-
-Please punctuate equations as regular text. Theorem-type environments (including propositions, lemmas, corollaries etc.) can be formatted as follows:
-%% Example of a theorem:
-\begin{Theorem}
-Example text of a theorem.
-\end{Theorem}
-
-The text continues here. Proofs must be formatted as follows:
-
-%% Example of a proof:
-\begin{proof}[Proof of Theorem 1]
-Text of the proof. Note that the phrase ``of Theorem 1'' is optional if it is clear which theorem is being referred to.
-\end{proof}
-The text continues here.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Discussion}
 
-Authors should discuss the results and how they can be interpreted from the perspective of previous studies and of the working hypotheses.
-The findings and their implications should be discussed in the broadest context possible.
-Future research directions may also be highlighted.
+- Comparison to NMF on linear mixing problem
+    - note that there are many versions of NMF with different constraints/regularization. The purpose here was to show that GSM is at least \textit{comparable} to NMF for linear unmixing problems
+    - Further note that both NMF and GSM do not rely on assumption of pure pixels (unlike VCA and other methods) 
+    - GSM also allows estimation of endmember variability via the precision parameter $\beta$ i.e. nodes in data space follow Gaussian distribution with $\sigma = \sqrt{\beta^{-1}}$
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+- Nonlinear GSM for robot team 
+    - discuss identification of rhodamine plume and easy abundance mapping via the embedding coordinate
+    - discuss intrinsic dimensionality identification via abundance sparsity and BIC
+    - Discuss identification of non-linearity \textit{strenghth} via
+
+- Discuss  extensions to the GSM
+    - batch training as outlined in Bishop paper, extensions to noise model to allow adaptive uncertainty for each wavelength bin
+
+- Discuss other applications of the GSM
+    - Optimal route planning for in-situ data collection based on prize-collecting travelling salesman problem, i.e.
+construct a route which maximizies area traversed in GSM latent space (e.g.
+the n-simplex).
+    - Applications to source apportionment, e.g. linear GSM can be used for standard source apportionment problems
+
 \section{Conclusions}
 
-This section is not mandatory, but can be added to the manuscript if the discussion is unusually long or complex.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Patents}
 
-This section is not mandatory, but may be added if there are patents resulting from the work reported in this manuscript.
+\textcolor{red}{UPDATE ME!}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% optional
-%\supplementary{The following supporting information can be downloaded at:  \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title.}
-
-% Only for journal Methods and Protocols:
-% If you wish to submit a video article, please do so with any other supplementary material.
-% \supplementary{The following supporting information can be downloaded at: \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title.
-A supporting video article is available at doi: link.}
-
-% Only for journal Hardware:
-% If you wish to submit a video article, please do so with any other supplementary material.
-% \supplementary{The following supporting information can be downloaded at: \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title.\vspace{6pt}\\
-%\begin{tabularx}{\textwidth}{lll}
-%\toprule
-%\textbf{Name} & \textbf{Type} & \textbf{Description} \\
-%\midrule
-%S1 & Python script (.py) & Script of python source code used in XX \\
-%S2 & Text (.txt) & Script of modelling code used to make Figure X \\
-%S3 & Text (.txt) & Raw data from experiment X \\
-%S4 & Video (.mp4) & Video demonstrating the hardware in use \\
-%... & ... & ... \\
-%\bottomrule
-%\end{tabularx}
-%}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\authorcontributions{For research articles with several authors, a short paragraph specifying their individual contributions must be provided.
-The following statements should be used ``Conceptualization, X.X.
-and Y.Y.; methodology, X.X.; software, X.X.; validation, X.X., Y.Y.
-and Z.Z.; formal analysis, X.X.; investigation, X.X.; resources, X.X.; data curation, X.X.; writing---original draft preparation, X.X.; writing---review and editing, X.X.; visualization, X.X.; supervision, X.X.; project administration, X.X.; funding acquisition, Y.Y.
-All authors have read and agreed to the published version of the manuscript.'', please turn to the  \href{http://img.mdpi.org/data/contributor-role-instruction.pdf}{CRediT taxonomy} for the term explanation.
-Authorship must be limited to those who have contributed substantially to the work~reported.}
+\authorcontributions{Methodology, J.W.; conceptualization, J.W.; software, J.W.; validation, J.W.; formal analysis, J.W.; investigation J.W.; resources, D.J.L.; writing---original draft preparation, J.W.; writing---review and editing, J.W.
+and D.J.L.; visualization, J.W.; supervision, D.J.L.; project administration, D.J.L.; funding acquisition, D.J.L.
+All authors have read and agreed to the published version of the manuscript.
+}
 
@@
 
-\institutionalreview{In this section, you should add the Institutional Review Board Statement and approval number, if relevant to your study.
-You might choose to exclude this statement if the study did not require ethical approval.
-Please note that the Editorial Office might ask you for further information.
-Please add “The study was conducted in accordance with the Declaration of Helsinki, and approved by the Institutional Review Board (or Ethics Committee) of NAME OF INSTITUTE (protocol code XXX and date of approval).” for studies involving humans.
-OR “The animal study protocol was approved by the Institutional Review Board (or Ethics Committee) of NAME OF INSTITUTE (protocol code XXX and date of approval).” for studies involving animals.
-OR “Ethical review and approval were waived for this study due to REASON (please provide a detailed justification).” OR “Not applicable” for studies not involving humans or animals.}
+\institutionalreview{Not applicable.}
 
-\informedconsent{Any research article describing a study involving humans should contain this statement.
-Please add ``Informed consent was obtained from all subjects involved in the study.'' OR ``Patient consent was waived due to REASON (please provide a detailed justification).'' OR ``Not applicable'' for studies not involving humans.
-You might also choose to exclude this statement if the study did not involve humans.
-
-Written informed consent for publication must be obtained from participating patients who can be identified (including by the patients themselves).
-Please state ``Written informed consent has been obtained from the patient(s) to publish this paper'' if applicable.}
+\informedconsent{Not applicable.}
 
@@
 
-% Only for journal Nursing Reports
-%\publicinvolvement{Please describe how the public (patients, consumers, carers) were involved in the research.
-Consider reporting against the GRIPP2 (Guidance for Reporting Involvement of Patients and the Public) checklist.
-If the public were not involved in any aspect of the research add: ``No public involvement in any aspect of this research''.}
-
-% Only for journal Nursing Reports
-%\guidelinesstandards{Please add a statement indicating which reporting guideline was used when drafting the report.
-For example, ``This manuscript was drafted against the XXX (the full name of reporting guidelines and citation) for XXX (type of research) research''.
-A complete list of reporting guidelines can be accessed via the equator network: \url{https://www.equator-network.org/}.}
-
-% Only for journal Nursing Reports
-%\useofartificialintelligence{Please describe in detail any and all uses of artificial intelligence (AI) or AI-assisted tools used in the preparation of the manuscript.
-This may include, but is not limited to, language translation, language editing and grammar, or generating text.
-Alternatively, please state that “AI or AI-assisted tools were not used in drafting any aspect of this manuscript”.}
-
 \acknowledgments{In this section you can acknowledge any support given which is not covered by the author contribution or funding sections.
@@
 
-\conflictsofinterest{Declare conflicts of interest or state ``The authors declare no conflicts of interest.'' Authors must identify and declare any personal circumstances or interest that may be perceived as inappropriately influencing the representation or interpretation of reported research results.
-Any role of the funders in the design of the study; in the collection, analyses or interpretation of data; in the writing of the manuscript; or in the decision to publish the results must be declared in this section.
-If there is no role, please state ``The funders had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript; or in the decision to publish the results''.}
+\conflictsofinterest{The authors declare no conflicts of interest.}
 
@@
 \appendixtitles{no} % Leave argument "no" if all appendix headings stay EMPTY (then no dot is printed after "Appendix A"). If the appendix sections contain a heading then change the argument to "yes".
-\appendixstart
-\appendix
-\section[\appendixname~\thesection]{}
-\subsection[\appendixname~\thesubsection]{}
-The appendix is an optional section that can contain details and data supplemental to the main text---for example, explanations of experimental details that would disrupt the flow of the main text but nonetheless remain crucial to understanding and reproducing the research shown; figures of replicates for experiments of which representative data are shown in the main text can be added here if brief, or as Supplementary Data.
-Mathematical proofs of results not central to the paper can be added as an appendix.
-
-\begin{table}[H] 
-\caption{This is a table caption.\label{tab5}}
-\begin{tabularx}{\textwidth}{CCC}
-\toprule
-\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
-\midrule
-Entry 1		& Data			& Data\\
-Entry 2		& Data			& Data\\
-\bottomrule
-\end{tabularx}
-\end{table}
-
-\section[\appendixname~\thesection]{}
-All appendix sections must be cited in the main text. In the appendices, Figures, Tables, etc. should be labeled, starting with ``A''---e.g., Figure A1, Figure A2, etc.
 
@@
 
-% Please provide either the correct journal abbreviation (e.g.
-according to the “List of Title Word Abbreviations” http://www.issn.org/services/online-services/access-to-the-ltwa/) or the full name of the journal.
-% Citations and References in Supplementary files are permitted provided that they also appear in the reference list here.
+\bibliography{paper/references}
+\PublishersNote{}
 
-%=====================================
-% References, variant A: external bibliography
-%=====================================
-%\bibliography{your_external_BibTeX_file}
-
-%=====================================
-% References, variant B: internal bibliography
-%=====================================
-\begin{thebibliography}{999}
-% Reference 1
-\bibitem[Author1(year)]{ref-journal}
-Author~1, T. The title of the cited article. {\em Journal Abbreviation} {\bf 2008}, {\em 10}, 142--149.
-% Reference 2
-\bibitem[Author2(year)]{ref-book1}
-Author~2, L. The title of the cited contribution. In {\em The Book Title}; Editor 1, F., Editor 2, A., Eds.; Publishing House: City, Country, 2007; pp. 32--58.
-% Reference 3
-\bibitem[Author3(year)]{ref-book2}
-Author 1, A.; Author 2, B. \textit{Book Title}, 3rd ed.; Publisher: Publisher Location, Country, 2008; pp. 154--196.
-% Reference 4
-\bibitem[Author4(year)]{ref-unpublish}
-Author 1, A.B.; Author 2, C. Title of Unpublished Work. \textit{Abbreviated Journal Name} year, \textit{phrase indicating stage of publication (submitted; accepted; in press)}.
-% Reference 5
-\bibitem[Author5(year)]{ref-communication}
-Author 1, A.B. (University, City, State, Country); Author 2, C. (Institute, City, State, Country). Personal communication, 2012.
-% Reference 6
-\bibitem[Author6(year)]{ref-proceeding}
-Author 1, A.B.; Author 2, C.D.; Author 3, E.F.
-Title of presentation.
-In Proceedings of the Name of the Conference, Location of Conference, Country, Date of Conference (Day Month Year); Abstract Number (optional), Pagination (optional).
-% Reference 7
-\bibitem[Author7(year)]{ref-thesis}
-Author 1, A.B. Title of Thesis. Level of Thesis, Degree-Granting University, Location of University, Date of Completion.
-% Reference 8
-\bibitem[Author8(year)]{ref-url}
-Title of Site. Available online: URL (accessed on Day Month Year).
-\end{thebibliography}
-
-% If authors have biography, please use the format below
-%\section*{Short Biography of Authors}
-%\bio
-%{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/author1.pdf}}}
-%{\textbf{Firstname Lastname} Biography of first author}
-%
-%\bio
-%{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/author2.jpg}}}
-%{\textbf{Firstname Lastname} Biography of second author}
-
-% For the MDPI journals use author-date citation, please follow the formatting guidelines on http://www.mdpi.com/authors/references
-% To cite two works by the same author: \citeauthor{ref-journal-1a} (\citeyear{ref-journal-1a}, \citeyear{ref-journal-1b}). This produces: Whittaker (1967, 1975)
-% To cite two works by the same author with specific pages: \citeauthor{ref-journal-3a} (\citeyear{ref-journal-3a}, p.
-328; \citeyear{ref-journal-3b}, p.475).
-This produces: Wong (1999, p.
-328; 2000, p.
-475)
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% for journal Sci
-%\reviewreports{\\
-%Reviewer 1 comments and authors’ response\\
-%Reviewer 2 comments and authors’ response\\
-%Reviewer 3 comments and authors’ response
-%}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\PublishersNote{}
 \end{adjustwidth}

### 2024-07-11 paper/paper.tex
initial update to paper with general outline and some thoughts for discussion, etc...

@@
     - discuss intrinsic dimensionality identification via abundance sparsity and BIC
-    - Discuss identification of non-linearity \textit{strenghth} via
+    - Discuss identification of non-linearity \textit{strength} via regularization parameter and BIC.
+    - reconstruction rmse can also be used to evaluate the relative performance for fixed model size (i.e. k and m) with varying regularization strength
 
@@
 
-\funding{Please add: ``This research received no external funding'' or ``This research was funded by NAME OF FUNDER grant number XXX.'' and  and ``The APC was funded by XXX''.
-Check carefully that the details given are accurate and use the standard spelling of funding agency names at \url{https://search.crossref.org/funding}, any errors may affect your future funding.}
+\funding{\textcolor{red}{Please add: ``This research received no external funding'' or ``This research was funded by NAME OF FUNDER grant number XXX.'' and  and ``The APC was funded by XXX''.
+Check carefully that the details given are accurate and use the standard spelling of funding agency names at \url{https://search.crossref.org/funding}, any errors may affect your future funding.}}
 
@@
 
-\dataavailability{We encourage all authors of articles published in MDPI journals to share their research data.
+\dataavailability{\textcolor{red}{We encourage all authors of articles published in MDPI journals to share their research data.
 In this section, please provide details regarding where data supporting reported results can be found, including links to publicly archived datasets analyzed or generated during the study.
 Where no new data were created, or where data is unavailable due to privacy or ethical restrictions, a statement is still required.
-Suggested Data Availability Statements are available in section ``MDPI Research Data Policies'' at \url{https://www.mdpi.com/ethics}.}
+Suggested Data Availability Statements are available in section ``MDPI Research Data Policies'' at \url{https://www.mdpi.com/ethics}.}}
 
-\acknowledgments{In this section you can acknowledge any support given which is not covered by the author contribution or funding sections.
-This may include administrative and technical support, or donations in kind (e.g., materials used for experiments).}
+\acknowledgments{\textcolor{red}{In this section you can acknowledge any support given which is not covered by the author contribution or funding sections.
+This may include administrative and technical support, or donations in kind (e.g., materials used for experiments).}}
 

### 2024-07-12 paper/paper.tex
some more minor changes
updated to fit parameter sweep for robot team dye data

@@
 \subsection{Nonlinear Mixing: Water Contaminant Identification}
-- describe robot team data collection & processing
+- describe robot team data collection and processing
 - describe
@@
 the n-simplex).
-    - Applications to source apportionment, e.g. linear GSM can be used for standard source apportionment problems
+    - Applications to source apportionment, e.g.
+linear GSM can be used for standard source apportionment problems (no reactions) and nonlinear GSM may be useful for modeling scenarios with aditional complications due to reactions and transfport from source to receptor.
 

### 2024-07-12 1b_usgs-fit-gsm.jl

@@ -39,8 +39,10 @@ snrs = [split(split(splitext(csv)[1], "_")[2], "-")[2] for csv ∈ csvs]
 stds = [parse(Float64, split(split(splitext(csv)[1], "_")[3], "-")[2]) for csv ∈ csvs]
 
 
-k = 75
-λs = [0.00001, 0.0001, 0.001, 0.01, 0.1]
+# k = 75
+k = 25
+# λs = [0.00001, 0.0001, 0.001, 0.01, 0.1]
+λs = [0.001,]
 Nᵥ = 3
 
 

### 2024-07-12 2b-robotteam-fit-gsm.jl

@@ -1,8 +1,12 @@
+println("Starting GSM fit script")
+println("\tLoading packages...")
 using MLJ, GenerativeTopographicMapping
 using DataFrames, CSV
 using StableRNGs
 using Random
 using JSON
+
+include("./utils/robot-team-config.jl")
 
 using CairoMakie
 include("./utils/makie-defaults.jl")
@@ -23,35 +27,36 @@ update_theme!(
 )
 
 
-outpath = "./output/robot-team"
-figpath = "./figures/robot-team"
+
+println("\tLoading data...")
+outpath = "./output/robot-team/dye"
+figpath = "./figures/robot-team/dye"
 mkpath(outpath)
 mkpath(figpath)
 
-
 include("./utils/fit-metrics.jl")
-
 
 # LOAD IN  COMBINED DATASET
 readdir("./data/robot-team/dye-test")
-datapath = "./data/robot-team/df_features.csv"
-df = CSV.read(datapath, DataFrame);
+readdir("./data/robot-team/")
+
+df_features = CSV.read("./data/robot-team/dye-test/df_features-nw.csv", DataFrame);
+# df_features = CSV.read("./data/robot-team/dye-test/df_features-nw.csv", DataFrame);
+# df_targets = CSV.read("./data/robot-team/dye-test/df_targets-nw.csv", DataFrame);
+
 
 
 # Explore data and infer intrinsic dimensionality (i.e. # of endmembers)
+println("\tAssessing dimensionality...")
+
 using LinearAlgebra
-
-size(Array(df))
-
-Svd = svd(Array(df) .- mean(Array(df), dims=1));
-
-σs = abs2.(Svd.S) ./ (nrow(df) - 1)
-
+Svd = svd(Array(df_features) .- mean(Array(df_features), dims=1));
+σs = abs2.(Svd.S) ./ (nrow(df_features) - 1)
 variances =  σs ./ sum(σs)
 cum_var = cumsum(σs ./ sum(σs))
 
 fig = Figure();
-ax = Axis(fig[1,1], xlabel="component", ylabel="Explained Variance", xticks=(1:length(σs)));
+ax = Axis(fig[1,1], xlabel="component", ylabel="Explained Variance", xticks=(1:length(σs)), xminorgridvisible=false);
 lines!(ax, 1:length(σs), variances, linewidth=2, linestyle=:dash)
 scatter!(ax, 1:length(σs), variances, markersize=25)
 xlims!(ax, 0.85, 10)
[…]

### 2024-07-12 utils/robot-team-config.jl

@@ -93,7 +93,7 @@ targets_latex= Dict(
 
 
 # wavelength bins used by sensor
-wavelengths = [390.960, 392.270, 393.590, 394.910, 396.230, 397.550, 398.870, 400.180, 401.500, 402.820, 404.140, 405.460, 406.780, 408.100, 409.420, 410.740, 412.060, 413.380, 414.700, 416.020, 417.340, 418.660, 419.980, 421.300, 422.620, 423.940, 425.260, 426.580, 427.910, 429.230, 430.550, 431.870, 433.190, 434.510, 435.840, 437.160, 438.480, 439.800, 441.120, 442.450, 443.770, 445.090, 446.410, 447.740, 449.060, 450.380, 451.710, 453.030, 454.350, 455.680, 457.000, 458.320, 459.650, 460.970, 462.300, 463.620, 464.940, 466.270, 467.590, 468.920, 470.240, 471.570, 472.890, 474.220, 475.540, 476.870, 478.190, 479.520, 480.840, 482.170, 483.500, 484.820, 486.150, 487.480, 488.800, 490.130, 491.450, 492.780, 494.110, 495.430, 496.760, 498.090, 499.420, 500.740, 502.070, 503.400, 504.730, 506.050, 507.380, 508.710, 510.040, 511.370, 512.700, 514.020, 515.350, 516.680, 518.010, 519.340, 520.670, 522.000, 523.330, 524.660, 525.990, 527.310, 528.640, 529.970, 531.300, 532.630, 533.960, 535.300, 536.630, 537.960, 539.290, 540.620, 541.950, 543.280, 544.610, 545.940, 547.270, 548.600, 549.940, 551.270, 552.600, 553.930, 555.260, 556.600, 557.930, 559.260, 560.590, 561.920, 563.260, 564.590, 565.920, 567.260, 568.590, 569.920, 571.260, 572.590, 573.920, 575.260, 576.590, 577.920, 579.260, 580.590, 581.930, 583.260, 584.600, 585.930, 587.260, 588.600, 589.930, 591.270, 592.600, 593.940, 595.280, 596.610, 597.950, 599.280, 600.620, 601.950, 603.290, 604.630, 605.960, 607.300, 608.640, 609.970, 611.310, 612.650, 613.980, 615.320, 616.660, 617.990, 619.330, 620.670, 622.010, 623.340, 624.680, 626.020, 627.360, 628.700, 630.040, 631.370, 632.710, 634
[…]

### 2024-07-15 paper/paper.tex
added idea to discussion about parallel GTM for large datasets

@@
 
-- Discuss  extensions to the GSM
-    - batch training as outlined in Bishop paper, extensions to noise model to allow adaptive uncertainty for each wavelength bin
+- Limitations of the GSM
+    - slow training due to computation of dinstance matrix between full dataset and embedding space
+    - An incremental version of the EM procedure as described in \cite{gtm-developments} can speed this up by considering batches $X_b$ of $X$ and performing only a partial E-step by updating responsibilities corresponding to the data batch
+    - However, M step still involves the "full" responsibility matrix. 
+    - This can be addressed by taking an ensembling approach as described in \cite{parallel-gtm} where multiple GTM are trained on subsets of $X$ and the results are averaged. 
+    - More generally, we may consider a mixture-of-GTMs as briefly desribed in \cite{gtm-orig} whereby the overall density is given as 
+    \begin{equation}
+        p(\mathbf{x}) = \sum_r P(r)p(\mathbf{x}\vert r)
+    \end{equation}
+    where there are $r$ individual GTM models with mixing coefficients $P(r)$ such that $\sum_r P(r) =1$
 

### 2024-07-15 paper/paper.tex
adding more to discussion section

@@
 
+- Other potential extensions to GSM model
+     - Adapt latent space  to allow for points whose barycentric coordinates sum to less than 1 in order to account for lighting effects (shouldn't matter for reflectance but would matter for radiance depending on incident light).
+This is the $\gamma$ factor that shows up in some of the mixing papers.
+
 - Discuss other applications of the GSM

### 2024-07-16 paper/paper.tex
added some equations

@@
 
+\newpage
+
 \section{Generative Simplex Mapping}
 \subsection{Method Derivation}
+
+\begin{equation}\label{eqn:data-space-distribution}
+    p(x \mid z, \mathbf{W}, \beta) = \left(\frac{\beta}{2\pi} \right)^{D/2}\exp\left( -\frac{\beta}{2}\lVert \psi(z; \mathbf{W}) - x \rVert^2 \right)
+\end{equation}
+
+\begin{equation}\label{eqn:latent-distribution}
+    p(z) = \sum\limits_k^K \pi_k \, \delta(z - z_k) \quad\text{where}\quad  \sum_k^K\pi_k = 1
+\end{equation}
+
+\begin{equation}\label{eqn:llh}
+    \mathcal{L}(\mathbf{W}, \beta) = \sum\limits_n^N \ln \left( \sum\limits_k^K \pi_k \, p(x_n \mid z_k, \mathbf{W}, \beta) \right)
+\end{equation}
+
+The activation function $\mathbf{\Phi}_{km} = \phi_m(z_k)$ is chosen to enable modeling linear and nonlinear mixing. For $m \leq N_v$, we take $\Phi_{km} = z_k$. For $m > N_v$, we set
+\begin{equation}\label{eq:act-function}
+    \mathbf{\Phi}_{km} = \begin{cases}
+        \frac{s - \lVert z_k - \mu_m \rVert}{s}; & \lVert z_k - \mu_m \rVert \leq s \\
+        0; & \lVert z_k - \mu_m \rVert > s
+    \end{cases}
+\end{equation}
+The activation function centers $\mu_m$ are selected to be greater than
+
+We include a prior distribution on model weights to enable
+\begin{align}\label{eq:weight-prior}
+\begin{split}
+    p(\mathbf{W} \mid \lambda_e, \lambda_w) &= \left(\frac{\lambda_e}{2\pi}\right)^{\frac{D\,N_v}{2}}\exp\left(-\frac{\lambda_e}{2}\sum\limits_d^D\sum\limits_{m=1}^{N_v}\mathbf{W}_{dm}^2 \right)\\ 
+    & \qquad \cdot \left(\frac{\lambda_w}{2\pi}\right)^{\frac{D(M-N_v)}{2}}\exp\left(-\frac{\lambda_w}{2}\sum\limits_d^D\sum_{m=N_v+1}^M \mathbf{W}_{dm}^2\right)
+\end{split}
+\end{align}
+
+\begin{equation}\label{eq:responsibility}
+    R_{kn}  = p(z_k \mid x_n, \mathbf{W}, \beta) = \dfrac{p(x_n \mid z_k, \mathbf{W}, \beta)}{\sum\limits_{k'}^K p(x_n \mid z_{k'}, \mathbf{W}, \beta)}
+\end{equation}
+
+Penalized complete-data log likelihood
+\begin{adjustwidth}{-\extralength}{0cm}
+\begin{equation}\label{eq:complete-data-llh}
+    Q = \sum\limits_n^N\sum\limits_k^K R_{kn} \left(\ln\pi_k + \frac{D}{2}\ln\left(\frac{\beta}{2\pi}\right) - \frac{\beta}{2}\sum\limits_d^D\left(\sum\limits_m^M \mathbf{W}_{dm}\mathbf{\Phi}_{km} - \mathbf{X}_{nd}\right)^2\right) + \frac{MD}{2}\ln\left(\frac{\lambda}{2\pi}\right) - \frac{\lambda}{2}\sum\limits_d^D \sum\limits_m^M \mathbf{W}_{dm}^2
+\end{equation}
+\end{adjustwidth}
+
+\begin{equation}\label{eq:pi-update}
+    \pi_k^{\text{new}}  = \frac{1}{N}\sum_n R_{kn}
+\end{equation}
+
+\begin{equation}\label{eq:beta-update}
+    \frac{1}{\beta^{\text{new}}}  = \frac{1}{ND}\sum\limits_n^N\sum\limits_k^K R_{kn}\lVert \psi(z_k; \mathbf{W}) - x_n \rVert^2
+\end{equation}
 
@@
 \subsection{Implementation Details}
+
+The mixing coefficients are initially chosen so that $\pi_k = \frac{1}{K}$.
 

### 2024-07-17 1b_usgs-fit-gsm.jl
updated figures

@@ -69,7 +69,7 @@ for i ∈ 1:length(csvs)
     println("\n------- Euclidean -----\n")
     nmf = NMF(k=Nᵥ, cost=:Euclidean, normalize_abundance=true, tol=1e-9, maxiters=1000, rng=StableRNG(42))
     mach = machine(nmf, df)
-    fit!(mach, verbosity=1)
+    fit!(mach, verbosity=0)
 
     fp = fitted_params(mach)
     H = fp.H
@@ -136,7 +136,7 @@ for i ∈ 1:length(csvs)
     println("\n------- KL -----------\n")
     nmf = NMF(k=Nᵥ, cost=:KL, normalize_abundance=true, tol=1e-9, maxiters=1000, rng=StableRNG(42))
     mach = machine(nmf, df)
-    fit!(mach, verbosity=1)
+    fit!(mach, verbosity=0)
 
     fp = fitted_params(mach)
     H = fp.H
@@ -204,7 +204,7 @@ for i ∈ 1:length(csvs)
 
     nmf = NMF(k=Nᵥ, cost=:L21, normalize_abundance=true, tol=1e-9, maxiters=1000, rng=StableRNG(42))
     mach = machine(nmf, df)
-    fit!(mach, verbosity=1)
+    fit!(mach, verbosity=0)
 
     fp = fitted_params(mach)
     H = fp.H
@@ -269,7 +269,7 @@ end
 
 
 
-# GSM
+# GSM Linear
 for i ∈ 1:length(csvs)
     csv = csvs[i]
     snr = snrs[i]
@@ -290,7 +290,7 @@ for i ∈ 1:length(csvs)
         println("\tλ = $(λ)\n")
         gsm = GSMLinear(k=k, Nv=Nᵥ, λ=λ, tol=1e-9, nepochs=500, niters=100, rng=StableRNG(42))
         mach = machine(gsm, df)
-        fit!(mach, verbosity=1)
+        fit!(mach, verbosity=0)
 
         Yorig = Matrix(df)
         Ŷ = data_reconstruction(mach, df)
@@ -313,6 +313,12 @@ for i ∈ 1:length(csvs)
         res_dict["σ_orignal"] = σnoise
         res_dict["SNR"] = snr
         res_dict["λ"] = λ
+
+        W = rpt[:W]
+        res_dict["W_min"] = minimum(W)
+        res_dict["W_mean"] = mean(W)
[…]

### 2024-07-17 1c_usgs-plots.jl

@@ -14,8 +14,8 @@ update_theme!(
     Axis=(
         xticklabelsize=20,
         yticklabelsize=20,
-        xlabelsize=22,
-        ylabelsize=22,
+        xlabelsize=20,
+        ylabelsize=20,
         titlesize=25,
     ),
     Colorbar=(
@@ -51,6 +51,8 @@ respath = joinpath(outpath, "linear")
 
 results_gsm = []
 results_gsm_big = []
+results_gsm_combo = []
+results_gsm_big_combo = []
 results_nmf_euc = []
 results_nmf_kl = []
 results_nmf_L21 = []
@@ -58,8 +60,14 @@ results_nmf_L21 = []
 for (root, dirs, files) ∈ walkdir(respath)
     for f ∈ files
         if endswith(f, ".json")
-            if occursin("gsm-big", f)
+            if occursin("gsm-big-combo_λ", f)
+                push!(results_gsm_big_combo, JSON.parsefile(joinpath(root, f)))
+            elseif occursin("gsm-combo_λ", f)
+                push!(results_gsm_combo, JSON.parsefile(joinpath(root, f)))
+            elseif occursin("gsm-big_λ", f)
                 push!(results_gsm_big, JSON.parsefile(joinpath(root, f)))
+            elseif occursin("gsm_λ", f)
+                push!(results_gsm, JSON.parsefile(joinpath(root, f)))
             elseif occursin("nmf-euclidean", f)
                 push!(results_nmf_euc, JSON.parsefile(joinpath(root, f)))
             elseif occursin("nmf-kl", f)
@@ -67,7 +75,7 @@ for (root, dirs, files) ∈ walkdir(respath)
             elseif occursin("nmf-L21", f)
                 push!(results_nmf_L21, JSON.parsefile(joinpath(root, f)))
             else
-                push!(results_gsm, JSON.parsefile(joinpath(root, f)))
+                continue
             end
         end
     end
[…]

### 2024-07-17 2a_robotteam-data-setup.jl

@@ -123,52 +123,73 @@ function get_h5_data(h5path, Δx = 0.1, skip_size = 5)
 end
 
 
-dfs = []
+dfs_dye = []
+dfs_nodye = []
+
 
 # loop over files and produce dataframes
 for (day, collections) in files_dict
     for (collection , files) in collections
         for f in files
-            println("Working on $(f)")
-            df = get_h5_data(f)
-            push!(dfs, df)
+            if collection == "dye"
+                println("Working on $(f)")
+                df = get_h5_data(f)
+                push!(dfs_dye, df)
+            else
+                println("Working on $(f)")
+                df = get_h5_data(f)
+                push!(dfs_nodye, df)
+            end
         end
     end
 end
 
 
-df_out = vcat(dfs...);
+df_dye = vcat(dfs_dye...);
+df_nodye = vcat(dfs_nodye...);
+
+# keep only water pixels
+idx_nw_dye = findall(df_dye.NDWI1 .≥ 0.25)
+idx_nw_nodye = findall(df_nodye.NDWI1 .≥ 0.25)
+
+df_dye_features = df_dye[idx_nw_dye, 1:idx_900]
+df_dye_targets = df_dye[idx_nw_dye, 463:end]
+
+df_nodye_features = df_nodye[idx_nw_nodye, 1:idx_900]
+df_nodye_targets = df_nodye[idx_nw_nodye, 463:end]
 
 
-df_features = df_out[:, 1:idx_900];
-df_targets = df_out[:, 463:end];
-
-CSV.write(joinpath(dye_path, "df_features.csv"), df_features)
-CSV.write(joinpath(dye_path, "df_targets.csv"), df_targets)
-
-
-idx_nw = findall(df_targets.NDWI1 .≥ 0.25)
-
-CSV.write(joinpath(dye_path, "df_features-nw.csv"),df_features[idx_nw, :])
-CSV.write(joinpath(dye_path, "df_targets-nw.csv"),df_targets[idx_nw, :])
+# save the data
+CSV.write(joinpath(dye_path, "df_dye_features.csv"), df_dye_features)
+CSV.write(joinpath(dye_path, "df_dye_targets.csv"), df_dye_targets)
+CSV.write(joinpath(dye_path, "df_nodye_features.csv"), df_nodye_features)
[…]

### 2024-07-17 2b-robotteam-fit-gsm.jl

@@ -1,10 +1,10 @@
-println("Starting GSM fit script")
-println("\tLoading packages...")
 using MLJ, GenerativeTopographicMapping
 using DataFrames, CSV
 using StableRNGs
 using Random
 using JSON
+using JLSO
+using LinearAlgebra
 
 include("./utils/robot-team-config.jl")
 
@@ -29,37 +29,33 @@ update_theme!(
 
 
 println("\tLoading data...")
-outpath = "./output/robot-team/dye"
-figpath = "./figures/robot-team/dye"
+outpath = "./output/robot-team/fits"
+figpath = "./figures/robot-team"
 mkpath(outpath)
 mkpath(figpath)
 
 include("./utils/fit-metrics.jl")
 
-# LOAD IN  COMBINED DATASET
-readdir("./data/robot-team/dye-test")
+# LOAD IN  DATASET
 readdir("./data/robot-team/")
-
-df_features = CSV.read("./data/robot-team/dye-test/df_features-nw.csv", DataFrame);
-# df_features = CSV.read("./data/robot-team/dye-test/df_features-nw.csv", DataFrame);
-# df_targets = CSV.read("./data/robot-team/dye-test/df_targets-nw.csv", DataFrame);
-
+df_features = CSV.read("./data/robot-team/df_features.csv", DataFrame);
+nrow(df_features)
 
 
 # Explore data and infer intrinsic dimensionality (i.e. # of endmembers)
 println("\tAssessing dimensionality...")
 
-using LinearAlgebra
 Svd = svd(Array(df_features) .- mean(Array(df_features), dims=1));
 σs = abs2.(Svd.S) ./ (nrow(df_features) - 1)
 variances =  σs ./ sum(σs)
 cum_var = cumsum(σs ./ sum(σs))
 
 fig = Figure();
-ax = Axis(fig[1,1], xlabel="component", ylabel="Explained Variance", xticks=(1:length(σs)), xminorgridvisible=false);
-lines!(ax, 1:length(σs), variances, linewidth=2, linestyle=:dash)
-scatter!(ax, 1:length(σs), variances, markersize=25)
-xlims!(ax, 0.85, 10)
+ax = Axis(fig[2,1], xlabel="Component", ylabel="Explained Variance (%)", xticks=(1:length(σs)), xminorgridvisible=false, yminorgridvisible=false);
+hl = hlines!(ax, [1,], color=mints_colors[2], linestyle=:solid, linewidth=2, label="1%")
[…]

### 2024-07-17 1a_usgs-setup.jl
adding in new figures

@@ -4,9 +4,10 @@ using Random, Statistics, LinearAlgebra
 using StableRNGs
 
 
-using CairoMakie, # TernaryDiagrams
+using CairoMakie #, TernaryDiagrams
 
 include("./utils/makie-defaults.jl")
+
 set_theme!(mints_theme)
 update_theme!(
     figure_padding=30,
@@ -39,9 +40,9 @@ if !ispath(joinpath(datapath, "linear"))
     mkpath(joinpath(datapath, "linear"))
 end
 
-if !ispath(joinpath(datapath, "bilinear"))
-    mkpath(joinpath(datapath, "bilinear"))
-end
+# if !ispath(joinpath(datapath, "bilinear"))
+#     mkpath(joinpath(datapath, "bilinear"))
+# end
 
 
 figpath = "./figures/1_usgs"
@@ -53,9 +54,9 @@ if !ispath(joinpath(figpath, "linear"))
     mkpath(joinpath(figpath, "linear"))
 end
 
-if !ispath(joinpath(figpath, "bilinear"))
-    mkpath(joinpath(figpath, "bilinear"))
-end
+# if !ispath(joinpath(figpath, "bilinear"))
+#     mkpath(joinpath(figpath, "bilinear"))
+# end
 
 
 outpath = "./output/1_usgs"
@@ -67,9 +68,9 @@ if !ispath(joinpath(outpath, "linear"))
     mkpath(joinpath(outpath, "linear"))
 end
 
-if !ispath(joinpath(outpath, "bilinear"))
-    mkpath(joinpath(outpath, "bilinear"))
-end
+# if !ispath(joinpath(outpath, "bilinear"))
+#     mkpath(joinpath(outpath, "bilinear"))
+# end
 
 
 
@@ -216,6 +217,7 @@ fig[1,1] = Legend(fig, [l1, l2, l3], [min_to_use...,], framevisible=false, orien
 xlims!(ax, λs[1], λs[end])
 
 save(joinpath(figpath, "endmembers-orig.png"), fig)
+save(joinpath(figpath, "endmembers-orig.svg"), fig)
 
 fig
 

### 2024-07-17 2b-robotteam-fit-gsm.jl

@@ -72,7 +72,7 @@ n_rbfs = 250
 # Nv_s = [3, 4, 5, 6, 7, 8, 9]
 
 
-Nv_s = [3, 4, 5, 6, 7]
+Nv_s = [3, 4, 5, 6, 7, 8, 9, 10]
 s = 1/10
 λe_s = [0.01, 0.001, 0.0001]
 λw_s = [10_000, 1_000, 100, 10, 1]

### 2024-07-18 2b-robotteam-fit-gsm.jl
updated figures

@@ -72,10 +72,10 @@ n_rbfs = 250
 # Nv_s = [3, 4, 5, 6, 7, 8, 9]
 
 
-Nv_s = [3, 4, 5, 6, 7, 8, 9, 10]
+Nv_s = [2, 3, 4, 5, 6, 7, 8, 9, 10]
 s = 1/10
-λe_s = [0.01, 0.001, 0.0001]
-λw_s = [10_000, 1_000, 100, 10, 1]
+λe_s = [0.01, 0.001, 0.0001, 0.00001]
+λw_s = [10_000, 1_000, 100, 10, 1, 0.1]
 
 println("\tNumber of models to train: ", length(Nv_s) * length(λe_s) * length(λw_s))
 println("\tTraiing models...")
@@ -159,18 +159,31 @@ CSV.write(joinpath("./output/robot-team/", "df_results.csv"), df_res)
 
 
 
+unique(df_res.λe)
+unique(df_res.λw)
+names(df_res)
 
 # Identify the best Nv via BIC, AIC, Q, LLH
-gdf = groupby(df_res, :Nv)
 
-df_Q = combine(gdf, :Q => maximum)
-df_llh = combine(gdf, :llhs => maximum)
-df_BIC = combine(gdf, :BIC => minimum)
-df_AIC = combine(gdf, :AIC => minimum)
+df_res[argmax(df_res.Q), [:Q, :Nv, :λe, :λw]]
+df_res[argmax(df_res.llhs),  [:llhs, :Nv, :λe, :λw]]
+df_res[argmin(df_res.BIC), [:BIC, :Nv, :λe, :λw]]
+df_res[argmin(df_res.AIC), [:AIC, :Nv, :λe, :λw]]
 
-Nv_Q = df_Q[argmax(df_Q.Q_maximum), :Nv]
-Nv_llh = df_llh[argmax(df_llh.llhs_maximum), :Nv]
-Nv_BIC = df_BIC[argmin(df_BIC.BIC_minimum), :Nv]
-Nv_AIC = df_AIC[argmin(df_AIC.AIC_minimum), :Nv]
+
+# gdf = groupby(df_res, :Nv)
+# df_Q = combine(gdf, :Q => maximum)
+# df_llh = combine(gdf, :llhs => maximum)
+# df_BIC = combine(gdf, :BIC => minimum)
+# df_AIC = combine(gdf, :AIC => minimum)
+
[…]

### 2024-07-18 paper/paper.tex

@@
 
+\begin{figure}[H]
+\includegraphics[width=\columnwidth]{paper/figures/methods/gsm/gsm-diagram.png}
+\caption{Illustration of the GSM.
+Points in the latent space (green dots) are uniformly distributed at the nodes of a regular grid on the n-simplex whose barycentric coordinates are to be interpreted as the relative abundance of n-many unique sources $z_1, ...
+z_{N_v}$.
+Here, $N_v=3$ has been chosen for illustrative purposes.
+These nodes are mapped into (spectral) data space via the mapping $\psi(z)$ utilizing $M$-many radially symmetric basis functions shown in red.
+The trained GSM also enables modelling of spectral variability via the precision parameter $\beta$ shown here as a light blue band around the estimated spectra in the data space.\label{fig:gsm-diagram}}
+\end{figure}
+
 \subsection{Implementation Details}

### 2024-07-18 paper/paper.tex
Updates from Overleaf

@@
 \begin{equation}\label{eq:responsibility}
-    R_{kn}  = p(z_k \mid x_n, \mathbf{W}, \beta) = \dfrac{p(x_n \mid z_k, \mathbf{W}, \beta)}{\sum\limits_{k'}^K p(x_n \mid z_{k'}, \mathbf{W}, \beta)}
+    \mathbf{R}_{kn}  = p(z_k \mid x_n, \mathbf{W}, \beta) = \dfrac{p(x_n \mid z_k, \mathbf{W}, \beta)}{\sum\limits_{k'}^K p(x_n \mid z_{k'}, \mathbf{W}, \beta)}
 \end{equation}
@@
 \begin{equation}\label{eq:complete-data-llh}
-    Q = \sum\limits_n^N\sum\limits_k^K R_{kn} \left(\ln\pi_k + \frac{D}{2}\ln\left(\frac{\beta}{2\pi}\right) - \frac{\beta}{2}\sum\limits_d^D\left(\sum\limits_m^M \mathbf{W}_{dm}\mathbf{\Phi}_{km} - \mathbf{X}_{nd}\right)^2\right) + \frac{MD}{2}\ln\left(\frac{\lambda}{2\pi}\right) - \frac{\lambda}{2}\sum\limits_d^D \sum\limits_m^M \mathbf{W}_{dm}^2
+    Q = \sum\limits_n^N\sum\limits_k^K \mathbf{R}_{kn} \left(\ln\pi_k + \frac{D}{2}\ln\left(\frac{\beta}{2\pi}\right) - \frac{\beta}{2}\sum\limits_d^D\left(\sum\limits_m^M \mathbf{W}_{dm}\mathbf{\Phi}_{km} - \mathbf{X}_{nd}\right)^2\right) + \frac{MD}{2}\ln\left(\frac{\lambda}{2\pi}\right) - \frac{\lambda}{2}\sum\limits_d^D \sum\limits_m^M \mathbf{W}_{dm}^2
 \end{equation}
@@
 \begin{equation}\label{eq:pi-update}
-    \pi_k^{\text{new}}  = \frac{1}{N}\sum_n R_{kn}
+    \pi_k^{\text{new}}  = \frac{1}{N}\sum_n \mathbf{R}_{kn}
 \end{equation}
@@
 \begin{equation}\label{eq:beta-update}
-    \frac{1}{\beta^{\text{new}}}  = \frac{1}{ND}\sum\limits_n^N\sum\limits_k^K R_{kn}\lVert \psi(z_k; \mathbf{W}) - x_n \rVert^2
+    \frac{1}{\beta^{\text{new}}}  = \frac{1}{ND}\sum\limits_n^N\sum\limits_k^K \mathbf{R}_{kn}\lVert \psi(z_k; \mathbf{W}) - x_n \rVert^2
 \end{equation}

### 2024-07-19 paper/paper.tex
Updates from Overleaf

@@
 \begin{figure}[H]
-\includegraphics[width=\columnwidth]{paper/figures/methods/gsm/gsm-diagram.png}
+\includegraphics[width=\columnwidth]{paper/figures/methods/gsm/gsm-diagram.pdf}
 \caption{Illustration of the GSM.
@@
 
+\begin{figure}[H]
+\includegraphics[width=0.85\columnwidth]{paper/figures/results-usgs/fit-comparison.pdf}
+\caption{Comparison of GSM against NMF on simulated USGS dataset. \label{fig:usgs-fits}}
+\end{figure}
+
 \subsection{Nonlinear Mixing: Water Contaminant Identification}
@@
 \subsection{Linear Mixing}
+
+\begin{figure}[H]
+\includegraphics[width=0.85\columnwidth]{paper/figures/results-usgs/fit-comparison.pdf}
+\caption{Comparison of GSM against NMF on simulated USGS dataset. \label{fig:usgs-fits}}
+\end{figure}
+
 \subsection{Nonlinear Mixing: Rhodamine Dye Plume}

### 2024-07-22 1c_usgs-plots.jl
updating with new fit results

@@ -459,4 +459,3 @@ xlims!(ax, λs[1], λs[end])
 fig
 
 save(joinpath(figpath, "linear", "extracted-endmembers.png"), fig)
-

### 2024-07-22 2b-robotteam-fit-gsm.jl

@@ -3,7 +3,6 @@ using DataFrames, CSV
 using StableRNGs
 using Random
 using JSON
-using JLSO
 using LinearAlgebra
 
 include("./utils/robot-team-config.jl")
@@ -69,16 +68,14 @@ n_rbfs = 250
 
 
 #k = 25 # this roughly corresponds to 3000 nodes
-# Nv_s = [3, 4, 5, 6, 7, 8, 9]
-
-
-Nv_s = [2, 3, 4, 5, 6, 7, 8, 9, 10]
+Nv_s = [2, 3, 4, 5, 6, 7, 8]
 s = 1/10
-λe_s = [0.01, 0.001, 0.0001, 0.00001]
-λw_s = [10_000, 1_000, 100, 10, 1, 0.1]
+λe_s = [0.1, 0.01, 0.001, 0.0001, 0.00001]
+λw_s = [10_000, 1_000, 100, 10, 1, 0.1, 0.01, 0.001]
+
 
 println("\tNumber of models to train: ", length(Nv_s) * length(λe_s) * length(λw_s))
-println("\tTraiing models...")
+println("\tTraining models...")
 
 for Nv ∈ Nv_s
     for λe ∈ λe_s
@@ -92,7 +89,7 @@ for Nv ∈ Nv_s
                 println("Nv: ", Nv, "\tλe: ", λe, "\tλw: ", λw)
                 println("\n")
 
-                gsm = GSMBigCombo(n_nodes=n_nodes, n_rbfs=n_rbfs, s=s, Nv=Nv, niters=10, tol=1e-9, nepochs=100)
+                gsm = GSMBigCombo(n_nodes=n_nodes, n_rbfs=n_rbfs, λe=λe, λw=λw, s=s, Nv=Nv, niters=10, tol=1e-9, nepochs=100)
 
                 mach = machine(gsm, df_features)
                 fit!(mach, verbosity=1)
@@ -116,23 +113,14 @@ for Nv ∈ Nv_s
                 open(json_path, "w") do f
                     JSON.print(f, res)
                 end
-
-                # Save the machine for later use
-                # https://juliaai.github.io/MLJ.jl/stable/machines/
-
-                smach = serializable(mach)
-                JLSO.save(jlso_path, :machine => smach)
             end
         end
     end
 end
 
 
-
-
 # Load in results and create single DataFrame
 df_res = []
-
[…]

### 2024-07-22 1b_usgs-fit-gsm.jl
updated figures

@@ -445,7 +445,7 @@ m = 10
 # ss = [0.5, 1.0, 1.5, 2.0, 2.5, 3.0]
 λe_s = [0.001,]
 #λw_s = [1000.0, 100.0, 10.0, 1.0]
-λw_s = [1000.0,]
+λw_s = [1.0, 100.0]
 Nᵥ = 3
 
 
@@ -495,7 +495,13 @@ for i ∈ 1:length(csvs)
             res_dict["SNR"] = snr
             res_dict["λe"] = λe
             res_dict["λw"] = λw
-            res_dict["W"] = rpt[:W]
+
+            W = rpt[:W]
+            res_dict["W"] = W
+            res_dict["W_min"] = minimum(W[:, Nᵥ+1:end])
+            res_dict["W_mean"] = mean(W[:, Nᵥ+1:end])
+            res_dict["W_median"] = median(W[:, Nᵥ+1:end])
+            res_dict["W_max"] = maximum(W[:, Nᵥ+1:end])
 
             res_dict["θ"] = mean([
                 minimum([spectral_angle(node_means[:, idx], R1) for idx ∈ idx_vertices]),
@@ -571,7 +577,15 @@ for i ∈ 1:length(csvs)
             res_dict["SNR"] = snr
             res_dict["λe"] = λe
             res_dict["λw"] = λw
-            res_dict["W"] = rpt[:W]
+
+            W = rpt[:W]
+            res_dict["W"] = W
+            res_dict["W_min"] = minimum(W[:, Nᵥ+1:end])
+            res_dict["W_mean"] = mean(W[:, Nᵥ+1:end])
+            res_dict["W_median"] = median(W[:, Nᵥ+1:end])
+            res_dict["W_max"] = maximum(W[:, Nᵥ+1:end])
+
+
 
             res_dict["θ"] = mean([
                 minimum([spectral_angle(node_means[:, idx], R1) for idx ∈ idx_vertices]),
@@ -613,5 +627,6 @@ for i ∈ 1:length(csvs)
         end
      end
 end
+
 
 

### 2024-07-22 1c_usgs-plots.jl

@@ -253,8 +253,8 @@ CSV.write(joinpath(respath, "df_nmf_L21.csv"), df_nmf_L21)
 # let's just take λ = 0.001
 df_gsm_plot = df_gsm[df_gsm.λ .== 0.001, :]
 df_gsm_big_plot = df_gsm_big[df_gsm_big.λ .== 0.001, :]
-df_gsm_combo_plot = df_gsm_combo[df_gsm_combo.λe .== 0.001 .&& df_gsm_combo.λw .== 1000.0, :]
-df_gsm_big_combo_plot = df_gsm_big_combo[df_gsm_big_combo.λe .== 0.001 .&& df_gsm_big_combo.λw .== 1000.0, :]
+df_gsm_combo_plot = df_gsm_combo[df_gsm_combo.λe .== 0.001 .&& df_gsm_combo.λw .== 100.0, :]
+df_gsm_big_combo_plot = df_gsm_big_combo[df_gsm_big_combo.λe .== 0.001 .&& df_gsm_big_combo.λw .== 100.0, :]
 
 
 # sort everything by SNR
@@ -276,11 +276,11 @@ df_gsm_big_combo_plot[:, [:SNR, :θ, :abund_rmse, :reconst_rmse]]
 fig = Figure(; px_per_unit=30);
 ax = Axis(fig[2,1], xlabel="SNR (dB)", ylabel="Mean Spectral Angle (degrees)", xticks=(1:9, ["∞", "35", "30", "25", "20", "15", "10", "5", "0"]), xminorgridvisible=false, yminorgridvisible=false);
 
-l_gsm = scatter!(ax, 1:9, df_gsm_plot[:, :θ], markersize=15, color=:green)
-lines!(ax, 1:9, df_gsm_big_combo_plot[:, :θ], linewidth=3, linestyle=:dash, color=:green)
+l_gsm_combo = scatter!(ax, 1:9, df_gsm_combo_plot[:, :θ], markersize=15, color=:green)
+lines!(ax, 1:9, df_gsm_combo_plot[:, :θ], linewidth=3, linestyle=:dash, color=:green)
 
-l_gsm_combo = scatter!(ax, 1:9, df_gsm_combo_plot[:, :θ], markersize=15, color=mints_colors[1], marker=:utriangle)
-lines!(ax, 1:9, df_gsm_combo_plot[:, :θ], linewidth=3, linestyle=:dash, color=mints_colors[1])
[…]

### 2024-07-22 2b-robotteam-fit-gsm.jl

@@ -61,17 +61,20 @@ fig
 save(joinpath(figpath, "pca-variance.png"), fig)
 
 
-
+fig
 # Based on this, between 4-8 endmembers should be fine
 n_nodes = 3000
 n_rbfs = 250
 
 
 #k = 25 # this roughly corresponds to 3000 nodes
-Nv_s = [2, 3, 4, 5, 6, 7, 8]
+# Nv_s = [2, 3, 4, 5, 6, 7, 8]
+Nv_s = [3, 4, 5, 6, 7, 8]
 s = 1/10
-λe_s = [0.1, 0.01, 0.001, 0.0001, 0.00001]
-λw_s = [10_000, 1_000, 100, 10, 1, 0.1, 0.01, 0.001]
+# λe_s = [0.1, 0.01, 0.001, 0.0001, 0.00001]
+# λw_s = [10_000, 1_000, 100, 10, 1, 0.1, 0.01, 0.001]
+λe_s = [0.1, 0.01, 0.001,]
+λw_s = [100, 10, 1, 0.1,]
 
 
 println("\tNumber of models to train: ", length(Nv_s) * length(λe_s) * length(λw_s))
@@ -171,6 +174,7 @@ df_res_1000[argmin(df_res_1000.BIC), [:BIC, :Nv, :λe]]
 df_res_100 = df_res[(df_res.λw .== 100), :]
 df_res_100[argmin(df_res_100.Q), [:Q, :Nv, :λe]]
 df_res_100[argmin(df_res_100.BIC), [:BIC, :Nv, :λe]]
+
 
 
 

### 2024-07-22 paper/paper.tex
Updates from Overleaf

@@
 \begin{figure}[H]
-\includegraphics[width=0.85\columnwidth]{paper/figures/results-usgs/fit-comparison.pdf}
-\caption{Comparison of GSM against NMF on simulated USGS dataset. \label{fig:usgs-fits}}
+\includegraphics[width=\columnwidth]{paper/figures/methods/usgs/usgs-dataset.pdf}
+\caption{Synthetic dataset from USGS spectra. \label{fig:usgs-data}}
 \end{figure}
@@
 
+\begin{figure}[H]
+\includegraphics[width=\columnwidth]{paper/figures/methods/robot-team/robot-team-overview.pdf}
+\caption{Robot team dataset. \label{fig:robotteam-data}}
+\end{figure}
+
 \section{Results}
@@
 \begin{figure}[H]
-\includegraphics[width=0.85\columnwidth]{paper/figures/results-usgs/fit-comparison.pdf}
+\includegraphics[width=0.85\columnwidth]{paper/figures/results/usgs/fit-comparison.pdf}
 \caption{Comparison of GSM against NMF on simulated USGS dataset. \label{fig:usgs-fits}}
@@
 
+\begin{figure}[H]
+\includegraphics[width=\columnwidth]{paper/figures/results/usgs/extracted-endmembers.png}
+\caption{Endmembers extracted using GSM for simulated USGS dataset with SNR$=20$. \label{fig:usgs-endmembers}}
+\end{figure}
+
 \subsection{Nonlinear Mixing: Rhodamine Dye Plume}
-\subsection{Nonlinear Mixing: Semi-supervised Analysis}
+
+Exploratory data analysis: 
+\begin{figure}[H]
+\includegraphics[width=0.60\columnwidth]{paper/figures/results/robot-team/pca-variance.png}
+\caption{Explained variance of PCA components for robot team dataset. A red horizontal line is superimposed on the plot indicating an explained variance of $1\%$. \label{fig:robot-team-pca}}
+\end{figure}
 

### 2024-07-29 1c_usgs-plots.jl
updated with new figures

@@ -456,7 +456,7 @@ l3_fit = lines!(ax, λs, vertex_2, color=(mints_colors[3], 0.75), linewidth=2) #
 
 fig[1,2] = Legend(fig, [l1, l2, l3, l1_fit, l2_fit, l3_fit], [min_to_use..., "Vertex 1", "Vertex 2", "Vertex 3"], framevisible=false, orientation=:vertical, padding=(0,0,0,0), labelsize=13, height=-5)
 xlims!(ax, λs[1], λs[end])
-
+ylims!(ax, 0, 1)
 fig
 
 save(joinpath(figpath, "linear", "extracted-endmembers.png"), fig)

### 2024-07-29 2a_robotteam-data-setup.jl

@@ -1,195 +1,253 @@
 using HDF5
 using CSV, DataFrames
 using ProgressMeter
+using Random, StableRNGs
 
 include("./utils/robot-team-config.jl")
-
-
 datapath = "./data/robot-team"
-dye_path = joinpath(datapath, "dye-test")
-sup_path = joinpath(datapath, "supervised")
+if !ispath(datapath)
+    mkpath(datapath)
+end
 
-mkpath(datapath)
-mkpath(dye_path)
-mkpath(sup_path)
-
-
-h5_basepath = "/Users/johnwaczak/data/robot-team/processed/hsi"
-@assert ispath(h5_basepath)
+# load the model
+h5_path = joinpath(datapath, "sample-datacube.h5")
+@assert ispath(h5_path)
 
 
+# open HDF5 file and gather data
+h5 = h5open(h5_path, "r")
+Δx = 0.1
+varnames = read(h5["data-Δx_$(Δx)/varnames"]);
+Data = read(h5["data-Δx_$(Δx)/Data"])[:, :, :];
+IsInbounds = read(h5["data-Δx_$(Δx)/IsInbounds"]);
+Longitudes = read(h5["data-Δx_$(Δx)/Longitudes"]);
+Latitudes = read(h5["data-Δx_$(Δx)/Latitudes"]);
+xs = read(h5["data-Δx_$(Δx)/X"]);
+ys = read(h5["data-Δx_$(Δx)/Y"]);
+close(h5)
+
+# get X and Y positions
+X = zeros(size(Data,2), size(Data,3))
+Y = zeros(size(Data,2), size(Data,3))
+
+# fill with values
+for x_i ∈ axes(Data,2)
+    for y_j ∈ axes(Data,3)
+        X[x_i, y_j] = xs[x_i]
+        Y[x_i, y_j] = ys[y_j]
+    end
+end
+
+idx_NDWI = findfirst(varnames .== "NDWI1")
 idx_900 = findfirst(wavelengths .≥ 900)
 
+IsPlume = zeros(Bool, size(IsInbounds))
+IsPlume[1:250, 150:end] .= true
 
-# generate file lists
-files_dict = Dict(
-    "11-23" => Dict(
-        "no-dye" => [
-            joinpath(h5_basepath, "11-23", "Scotty_1", "Scotty_1-2.h5"),
-            joinpath(h5_basepath, "11-23", "Scotty_1", "Scotty_1-7.h5"),
[…]

### 2024-07-29 2b-robotteam-fit-gsm.jl

@@ -1,9 +1,11 @@
-using MLJ, GenerativeTopographicMapping
+#using MLJ, GenerativeTopographicMapping
+using MLJ, GenerativeTopographicMapping, MLJNonnegativeMatrixFactorization
 using DataFrames, CSV
 using StableRNGs
 using Random
 using JSON
 using LinearAlgebra
+
 
 include("./utils/robot-team-config.jl")
 
@@ -51,8 +53,8 @@ cum_var = cumsum(σs ./ sum(σs))
 
 fig = Figure();
 ax = Axis(fig[2,1], xlabel="Component", ylabel="Explained Variance (%)", xticks=(1:length(σs)), xminorgridvisible=false, yminorgridvisible=false);
-hl = hlines!(ax, [1,], color=mints_colors[2], linestyle=:solid, linewidth=2, label="1%")
 barplot!(ax, 1:length(σs), variances .* 100, color=mints_colors[1], strokecolor=:darkgreen, strokewidth=2)
+hl = hlines!(ax, [1,], color=mints_colors[2], linestyle=:solid, linewidth=2, label="1%")
 fig[1,1] = Legend(fig, [hl,], ["1 %",], framevisible=false, orientation=:horizontal, padding=(0,0,-20,0), labelsize=14, height=-5, halign=:right)
 xlims!(ax, 0.5, 10)
 ylims!(ax, -0.01, nothing)
@@ -61,21 +63,17 @@ fig
 save(joinpath(figpath, "pca-variance.png"), fig)
 
 
-fig
 # Based on this, between 4-8 endmembers should be fine
 n_nodes = 3000
 n_rbfs = 250
 
-
 #k = 25 # this roughly corresponds to 3000 nodes
-# Nv_s = [2, 3, 4, 5, 6, 7, 8]
-Nv_s = [3, 4, 5, 6, 7, 8]
+Nv_s = [3, 4, 5, 6]
 s = 1/10
-# λe_s = [0.1, 0.01, 0.001, 0.0001, 0.00001]
-# λw_s = [10_000, 1_000, 100, 10, 1, 0.1, 0.01, 0.001]
-λe_s = [0.1, 0.01, 0.001,]
-λw_s = [100, 10, 1, 0.1,]
+λe_s = [1.0, 0.1, 0.01, 0.001,]
+λw_s = [1_000., 100, 10, 1, 0.1,]
 
+Yorig = Matrix(df_features);
 
[…]

### 2024-07-29 2c_robotteam-plots.jl

@@ -1,0 +1,410 @@
+using MLJ, GenerativeTopographicMapping
+using DataFrames, CSV
+using StableRNGs
+using Random
+using JSON
+using LinearAlgebra
+using HDF5
+using Statistics
+
+include("./utils/robot-team-config.jl")
+include("./utils/datacube-viz.jl")
+
+using CairoMakie
+include("./utils/makie-defaults.jl")
+set_theme!(mints_theme)
+update_theme!(
+    figure_padding=30,
+    Axis=(
+        xticklabelsize=20,
+        yticklabelsize=20,
+        xlabelsize=22,
+        ylabelsize=22,
+        titlesize=25,
+    ),
+    Colorbar=(
+        ticklabelsize=20,
+        labelsize=22
+    )
+)
+
+
+include("./utils/fit-metrics.jl")
+
+
+figpath = "./figures/robot-team"
+outpath = "./output/robot-team"
+json_path = joinpath(outpath, "fit_final.json")
+mdl_path = joinpath(outpath, "model_final.jls")
+
+res_dict = JSON.parsefile(json_path)
+mach = machine(mdl_path)
+rpt = report(mach)
+
+Nv = res_dict["Nv"]
+W = rpt[:W]
+
+df_features = CSV.read("./data/robot-team/df_features.csv", DataFrame);
+
+idx_vertices =rpt[:idx_vertices]
+node_means =rpt[:node_data_means]
+Rorig = Array(df_features);
+
+Ypred = data_reconstruction(mach, df_features)
+Zpred = DataFrame(MLJ.transform(mach, df_features))
+idx1_max = argmax(Zpred.Z1)
+idx2_max = argmax(Zpred.Z2)
+idx3_max = argmax(Zpred.Z3)
+# visualize endmembers
+
+# c_water = mints_colors[3]
+# c_veg = mints_colors[1]
+# c_dye = mints_colors[2]
+
+c_water = CairoMakie.RGBf(0,0,1)
+c_veg = CairoMakie.RGBf(0,1,0)
+c_dye = CairoMakie.RGBf(1,0,0)
+
+fig = Figure();
+ax = CairoMakie.Axis(fig[2,1], xlabel="λ (nm)", ylabel="Reflectance")
+
+# l_veg = lines!(ax, wavelengths, node_means[:,idx_vertices[1]], color=c_veg)
[…]

### 2024-07-29 2d_plume-evolution.jl

@@ -1,0 +1,309 @@
+include("./utils/background-viz.jl")
+
+# see https://en.wikipedia.org/wiki/Geographic_coordinate_system#Latitude_and_longitude
+# for detail
+ϕ_scale = 33.70093
+m_per_deg = 111412.84*cosd(ϕ_scale) - 93.5*cosd(3*ϕ_scale) + 0.118 * cosd(5*ϕ_scale)
+λ_scale_l = -97.7166
+λ_scale_r = λ_scale_l + 30/m_per_deg
+
+w= -97.717472
+n= 33.703572
+s= 33.700797
+e= -97.712413
+
+satmap = get_background_satmap(w,e,s,n)
+
+lon_min, lon_max = (-97.7168, -97.7125)
+lat_min, lat_max = (33.70075, 33.7035)
+
+
+using MLJ, GenerativeTopographicMapping
+using DataFrames, CSV
+using StableRNGs
+using Random
+using JSON
+using LinearAlgebra
+using HDF5
+using Statistics
+
+include("./utils/robot-team-config.jl")
+include("./utils/datacube-viz.jl")
+include("./utils/fit-metrics.jl")
+
+using CairoMakie
+include("./utils/makie-defaults.jl")
+set_theme!(mints_theme)
+update_theme!(
+    figure_padding=30,
+    Axis=(
+        xticklabelsize=20,
+        yticklabelsize=20,
+        xlabelsize=22,
+        ylabelsize=22,
+        titlesize=25,
+    ),
+    Colorbar=(
+        ticklabelsize=20,
+        labelsize=22
+    )
+)
+
+
+
+
+figpath = "./figures/robot-team"
+outpath = "./output/robot-team"
+json_path = joinpath(outpath, "fit_final.json")
+mdl_path = joinpath(outpath, "model_final.jls")
+mach = machine(mdl_path)
+
+
+f_list_dye = [
+    joinpath("./data", "robot-team", "dye_1.h5"),
+    joinpath("./data", "robot-team", "dye_21.h5"),
+    joinpath("./data", "robot-team", "dye_22.h5"),
+]
+@assert all(ispath.(f_list_dye))
+
+
+fpath_1 = joinpath("./data", "robot-team", "sample-datacube.h5")
[…]

### 2024-07-29 utils/background-viz.jl

@@ -1,0 +1,44 @@
+using CondaPkg
+CondaPkg.add("numpy")
+CondaPkg.add("contextily")
+using PythonCall
+using Images
+
+ctx = pyimport("contextily")
+
+struct SatMap
+    w
+    e
+    s
+    n
+    img end
+
+"""
+    get_background_satmap(w::Float64, e::Float64, s::Float64, n::Float64; out_name::String="Scotty)
+
+
+Grab Esri World Imagery tiles for region with corners (w,n), (e,s) in longitude and latitude.
+Saves resulting image to `outpath`
+
+**Note:** result images are saved in Web-Mercator projection by default. See `WebMercatorfromLLA` and `LLAfromWebMercator` from `Geodesy.jl` for conversion details.
+"""
+function get_background_satmap(w::Float64, e::Float64, s::Float64, n::Float64; out_name::String="Scotty")
+    # ctx = pyimport("contextily")
+    tiff, ext = ctx.bounds2raster(w, s, e, n, out_name*".tiff", source=ctx.providers["Esri"]["WorldImagery"], ll=true)
+    warped_tiff, warped_ext = ctx.warp_tiles(tiff, ext, "EPSG:4326")
+
+    warped_ext = pyconvert(Vector{Float64}, warped_ext)
+    tiff_array = permutedims(pyconvert(Array{Int}, warped_tiff)./255, (3,1,2))
+    tiff_img = colorview(RGBA, tiff_array)
+
+    tiff_img = rotr90(tiff_img)
+
+    return SatMap(
+        warped_ext[1],
+        warped_ext[2],
+        warped_ext[3],
+        warped_ext[4],
+        tiff_img
+    )
+end
+

### 2024-07-29 utils/datacube-viz.jl

@@ -1,0 +1,59 @@
+using Images
+
+
+function getRGB(h5::HDF5.File; λred=630.0, λgreen=532.0, λblue=465.0, Δx=0.10, α=10.0, β=0.0)
+    λs = h5["data-Δx_$(Δx)/λs"][:]
+
+    λred=630.0
+    λgreen=532.0
+    λblue=465.0
+
+    idx_r = argmin(abs.(λs .- λred))
+    idx_g = argmin(abs.(λs .- λgreen))
+    idx_b = argmin(abs.(λs .- λblue))
+
+    Rr = h5["data-Δx_$(Δx)/Data"][idx_r, :, :]
+    Rg = h5["data-Δx_$(Δx)/Data"][idx_g, :, :]
+    Rb = h5["data-Δx_$(Δx)/Data"][idx_b, :, :]
+
+    ij_pixels = findall(h5["data-Δx_$(Δx)/IsInbounds"][:,:])
+    img = zeros(4, size(Rr)...)
+
+    Threads.@threads for ij ∈ ij_pixels
+        img[1, ij] = Rr[ij]
+        img[2, ij] = Rg[ij]
+        img[3, ij] = Rb[ij]
+        img[4, ij] = 1.0
+    end
+
+    return img
+end
+
+
+function gamma_correct(img, γ=1/2)
+    # see https://en.wikipedia.org/wiki/Gamma_correction
+    img_out = copy(img)
+
+    img_out[1:3,:,:] .= img_out[1:3,:,:] .^ γ
+
+    return img_out
+end
+
+function brighten(img, α=0.13, β=0.0)
+    # see : https://www.satmapper.hu/en/rgb-images/
+    img_out = copy(img)
+    for i ∈ 1:3
+        img_out[i,:,:] .= clamp.(α .* img_out[i,:,:] .+ β, 0, 1)
+    end
+
+    return img_out
+end
+
+function get_h5_rgb(h5path)
+    h5 = h5open(h5path)
+    img = getRGB(h5)
+    close(h5)
+    img_out = brighten(gamma_correct(img, .75), 3.5, 0)
+    return colorview(RGBA, img_out)
+end
+

### 2024-07-29 1c_usgs-plots.jl
new figures\!

@@ -273,8 +273,9 @@ df_gsm_big_combo_plot[:, [:SNR, :θ, :abund_rmse, :reconst_rmse]]
 
 
 # figure out the lambda with the best θ for each SNR
-fig = Figure(; px_per_unit=30);
-ax = Axis(fig[2,1], xlabel="SNR (dB)", ylabel="Mean Spectral Angle (degrees)", xticks=(1:9, ["∞", "35", "30", "25", "20", "15", "10", "5", "0"]), xminorgridvisible=false, yminorgridvisible=false);
+fig = Figure(;);
+gl = fig[1,1] = GridLayout();
+ax = Axis(gl[2,1], xlabel="SNR (dB)", ylabel="Mean Spectral Angle (degrees)", xticks=(1:9, ["∞", "35", "30", "25", "20", "15", "10", "5", "0"]), xminorgridvisible=false, yminorgridvisible=false);
 
 l_gsm_combo = scatter!(ax, 1:9, df_gsm_combo_plot[:, :θ], markersize=15, color=:green)
 lines!(ax, 1:9, df_gsm_combo_plot[:, :θ], linewidth=3, linestyle=:dash, color=:green)
@@ -291,7 +292,16 @@ lines!(ax, 1:9, df_nmf_kl[:, :θ], linewidth=3, linestyle=:dash, color=:red)
 l_nmf_L21 = scatter!(ax, 1:9, df_nmf_L21[:, :θ], markersize=15, color=:orange, marker=:star5)
 lines!(ax, 1:9, df_nmf_L21[:, :θ], linewidth=3, linestyle=:dash, color=:orange)
 
-fig[1,1] = Legend(fig, [l_gsm_combo, l_gsm_big_combo, l_nmf_euc, l_nmf_kl, l_nmf_L21], ["GSM", "GSM (Big)", "NMF (ℓ₂)", "NMF (KL)", "NMF (ℓ₂,₁)"], framevisible=false, orientation=:horizontal, padding=(0,0,0,0), labelsize=14, height=-5)
[…]

### 2024-07-30 2b-robotteam-fit-gsm.jl
added table with top 10 models from hyperparam search

@@ -153,13 +153,16 @@ unique(df_res.λw)
 sort(df_res[df_res.λw .== 10.0 , [:Nv, :λe, :λw, :BIC, :AIC, :Q]], :BIC)
 
 
-df_best = df_res[df_res.λw .≥ 1.0 , :];
+df_best = sort(df_res[df_res.λw .≥ 1.0 , :], :BIC);
+
+df_out = df_best[1:10, [:Nv, :λe, :λw, :BIC, :AIC, :rmse_reconst,]]
+CSV.write("./paper/fit-res.csv", df_out)
 # df_best = df_res
 # Identify the best Nv via BIC, AIC, Q, LLH
 
 df_best[argmax(df_best.Q), [:Q, :Nv, :λe, :λw]]
+
 df_best[argmax(df_best.llhs),  [:llhs, :Nv, :λe, :λw]]
-df_best[argmin(df_best.BIC), [:BIC, :Nv, :λe, :λw]]
 df_best[argmin(df_best.AIC), [:AIC, :Nv, :λe, :λw]]
 
 

### 2024-07-30 paper/paper.tex
Updates from Overleaf

@@
 \begin{figure}[H]
-\includegraphics[width=0.85\columnwidth]{paper/figures/results/usgs/fit-comparison.pdf}
+\includegraphics[width=\columnwidth]{paper/figures/results/usgs/fit-comparison.pdf}
 \caption{Comparison of GSM against NMF on simulated USGS dataset. \label{fig:usgs-fits}}
@@
 \caption{Explained variance of PCA components for robot team dataset. A red horizontal line is superimposed on the plot indicating an explained variance of $1\%$. \label{fig:robot-team-pca}}
+\end{figure}
+
+\begin{table}[H] 
+\caption{This is a table caption. Tables should be placed in the main text near to the first time they are~cited.\label{table:fit-comparison}}
+\begin{tabularx}{\textwidth}{CCC}
+\toprule
+\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
+\midrule
+Entry 1		& Data			& Data\\
+Entry 2		& Data			& Data \textsuperscript{1}\\
+\bottomrule
+\end{tabularx}
+\noindent{\footnotesize{\textsuperscript{1} Tables may have a footer.}}
+\end{table}
+
+\begin{figure}[H]
+\begin{adjustwidth}{-\extralength}{0cm}
+\centering
+\includegraphics[width=1.25\columnwidth]{paper/figures/results/robot-team/extracted-endmembers.pdf}
+\end{adjustwidth}
+\caption{Nonlinear GSM applied to water spectra: \textbf{(a)} Spectra corresponding to maximum abundances for each vertex in the trained GSM identified for as water, near-shore vegetation, and rhodamine dye sources.
+\textbf{(b)} The original hyperspectral datacube segmented according to the relative abundance of each endmember.
+Each water pixel defined by an NDWI $\geq 0.25$ is colored by smoothing interpolating between red, green, and blue colors corresponding to the relative abundance estimated for rhodamine, vegetation, and water spectra.}
+\label{fig:robot-team-endmembers}
+\end{figure}
+
+\newpage
+\begin{figure}[H]
+\begin{adjustwidth}{-\extralength}{0cm}
+\centering
+\includegraphics[width=\columnwidth]{paper/figures/results/robot-team/endmember-abundances.pdf}
+\caption{Endmember distributions: \textbf{(a)} The distribution of abundance for the water class.
+This source dominates in the center of the water and decreases towards the edge of the pond where surface vegetation begins to dominate the reflectance signal.
+We note that the water abundance decreases near  the edge of the rhodamine plume reflecting dye mixing and diffusion.
+\textbf{(b)} The distribution of vegetation.
+This signal includes filamentous blue-green algae observed to accumulate in shallow waters near the shore.
+\textbf{(c)} The rhodamine dye plume extent segmented from the HSI.
+The total area for near-shore vegetation and rhodamine are estimated to be $378.6$ $\text{m}^2$ and $255.7$ $\text{m}^2$ respectively.}
+\end{adjustwidth}
+\label{fig:endmember-abundance-dist}
+\end{figure}  
+\newpage
+
+\begin{figure}[H]
+\begin{adjustwidth}{-\extralength}{0cm}
+\centering
+\includegraphics[width=1.25\columnwidth]{paper/figures/results/robot-team/plume-evo.pdf}
+\end{adjustwidth}
+\caption{Rhodamine plume evolution: Using the trained GSM we can track the dispersion of the rhodamine dye plume between successive drone flights.
+\textbf{(a)} The initial plume distribution after release.
+Here the dye subsumes an area of $255.7$ $\text{m}^2$.
+\textbf{(b)} The same plume imaged 15 minutes later now extends across an area of $571.8$ $\text{m}^2$}
+\label{fig:plume-evo}
 \end{figure}

### 2024-08-01 paper/paper.tex
updates to introduction

@@
 
+Hyperspectral imaging has emerged as a keystone technology in remote sensing where the ability to discern slight variations in high resolution spectra facilitates a plethora of critical applications from land use assessment...
+etc.
+- environmental monitoring, resource management, risk assessment, biodiversity conservation, sustainable agriculture,
+- Many hyperspectral remote sensing missions have already been launched with many more in development
+    - EnMAP (launched 2022) is german satellite with remote sensing imager
+    - PRISMA (launched 2019) is italian remote sensing satellite
+    - HISUI (launched 2019) is japanese HSI on international space station
+    - CHIME (future launch) - Copernicus Hyperspectral Imaging mission for the environment from the european space agency
+    - SBG (future launch) - surface biology and geology - NASA mission
+    - SHALOM (future launch) - Joint israeli-italian mission
+    - HyspIRI (future launch) - italian missions
+- Additionally, continued improvement to HSI technologies have dramatically reduced their size enabling their inclusion to the payloads of small unmanned aerial vehicles (UAV).
+
+- Key challenge: how to make sense of big data
+- Common approaches involve computation of spectral indices like the NDVI
+    - These are nice but can be challenging to compare between instruments due to varibility in spectral bands and sensor response
+    - They also throw away
+- A second common approach is to directly map spectral data to parameters of interest using in situ reference data 
+    - This is popular for water quality assessment where parameters like CDOM, etc. track water quality
+    - The key limitation is the availability of coincident reference data. This can requires decades of aligned observations to provide sufficient data volumes for statistical/machine learning models
+- In light of these difficulties, unsupervised methods which extract relevant information from HSI themselves can be advantageous
+- Reality: pixels captured by remote sensing platforms have spatial extents which typically cover multiple unique sources. 
+- overview of spectral unmixing
+- relevance to source identification in remote sensing, contaminant identication 
+- relevance to source apportionment in air quality monitoring
+
 2. Traditional approaches for endmember extraction and spectral unmixing
+- Standard assumption: linear mixing model (LMM)
+- Standard requirements for LMM and other mixing models
+    - abundance sum-to-one contraint $\sum_i a_i = 1$
+    - endmember non-negativity constraint
+- Secondary assumption: existence of pure pixels, e.g. VCA and other algorithms
+- PMF and NMF as unsupervised LMM models 
+- Relevance of nonlinear mixing, e.g. multiple scattering, interference effects, etc...
+- Approaches for nonlinear unmixing
+    - radiative transfer models
+    - Deep learning approaches
 
 3. Generative topographic mapping
+- Self organizing map is a data-driven unsupervised classification algorithm
+- GTM is a principled extension of the SOM to a fully probabalistic latent-space model
+- Uses of the GTM in cheminformatics
+- Our previous work using the GTM
+- Limitations of the GTM
+    - 2 dimensional latent space
+    - Abundance estimation requires other heuristics like NS3
+    - No direct guarantee of endmember non-negativity. In practice, this is not usually a problem when data are non-negative but it would be nice to have some guarantee
+- Flexibility of the GTM model suggest augmenting the latent space distribution may lead to a more interpretable model
 
 4. Proposed method: generative simplex mapping
+- We propose a novel model dubbed Generative Simplex Mapping which naturally extends the GTM to directly facilitate mixture models. 
+- The latent space is modified to an N-simplex where each vertex naturally corresponds to a unique source
+- positions within the simplex correspond to different mixing ratios
+- The latent space is then mapped to the data space using an activation function which allows for flexible models including both linear and nonlinear unmixing. 
+- Like the NMF, the GSM simultaneously extracts endmembers \textit{and} estimates abundances
+- Importantly, the GSM learns a probability distribution for the data allowing spectral
 
 5. Summary of paper structure
+- In this paper, we present the GSM and demonstrate its application to both synthetic and real datasets. The structure of the paper is as follows: 
+- In section 2 we provide a detailed overview of the GSM model and it's implementation
+- In section 3 we describe experiments using synthetic and real data to evaluate the performance of the GSM
+- In section 4 we present the experimental results
+- In sections 5 we discuss additional applications of the GSM and further extensions to be explored in future work. 
+- Finally, in section 6, we present a summary of our conclusions.
 
@@
     \frac{1}{\beta^{\text{new}}}  = \frac{1}{ND}\sum\limits_n^N\sum\limits_k^K \mathbf{R}_{kn}\lVert \psi(z_k; \mathbf{W}) - x_n \rVert^2
+\end{equation}
+
+\begin{equation}\label{eq:W-update}
+    W_{dm}^\text{new} = \dfrac{\left(X^TR^T\Phi\right)_{dm}}{\left(W\Phi^TG\Phi + \Lambda\right)_{dm}} 
 \end{equation}
@@
 \begin{table}[H] 
-\caption{This is a table caption. Tables should be placed in the main text near to the first time they are~cited.\label{table:fit-comparison}}
-\begin{tabularx}{\textwidth}{CCC}
+\caption{Hyperparameter optimization: Multiple models were trained to identify optimal hyperparameter values for the GSM applied to the water spectra dataset.
+Here we report the top $10$ models ranked according the the BIC.}
+\label{table:fit-comparison}
+\begin{tabularx}{\textwidth}{CCCCCC}
 \toprule
-\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
+\textbf{$N_v$}	& \textbf{$\lambda_e$}	& \textbf{$\lambda_w$} & \textbf{BIC} & \textbf{AIC} & \textbf{Reconstruction RMSE}\\
 \midrule
-Entry 1		& Data			& Data\\
-Entry 2		& Data			& Data \textsuperscript{1}\\
+$3$ & $0.01$    & $1.0$     & $-6.195\times10^7$   & $-6.269\times10^7$   & $0.000989$ \\
+$3$	& $0.001$   & $1.0$     & $-6.194\times10^7$   & $-6.268\times10^7$   & $0.000989$ \\
+$3$	& $0.1$     & $1.0$	    & $-6.192\times10^7$   & $-6.265\times10^7$   & $0.000991$ \\
+$3$	& $1.0$     & $1.0$	    & $-6.186\times10^7$   & $-6.260\times10^7$   & $0.001190$ \\
+$4$	& $1.0$     & $1.0$	    & $-6.181\times10^7$   & $-6.255\times10^7$   & $0.001002$ \\
+$4$	& $0.1$     & $1.0$	    & $-6.175\times10^7$   & $-6.249\times10^7$   & $0.001008$ \\
+$4$	& $0.01$    & $1.0$     & $-6.173\times10^7$   & $-6.247\times10^7$   & $0.001009$ \\ 
+$4$	& $0.001$   & $1.0$     & $-6.173\times10^7$   & $-6.247\times10^7$   & $0.001009$ \\
+$4$	& $0.1$     & $10.0$    & $-6.171\times10^7$   & $-6.245\times10^7$   & $0.001011$ \\
+$3$	& $1.0$     & $10.0$    & $-6.166\times10^7$   & $-6.239\times10^7$   & $0.001014$ \\
 \bottomrule
 \end{tabularx}
-\noindent{\footnotesize{\textsuperscript{1} Tables may have a footer.}}
 \end{table}
@@
     - Applications to source apportionment, e.g.
-linear GSM can be used for standard source apportionment problems (no reactions) and nonlinear GSM may be useful for modeling scenarios with aditional complications due to reactions and transfport from source to receptor.
+linear GSM can be used for standard source apportionment problems (no reactions) and nonlinear GSM may be useful for modeling scenarios with additional complications due to reactions and during transport from source to receptor.
 

### 2024-08-01 paper/paper.tex
more updates to the paper

@@
 % cleveref: load \crefname definitions after \begin{document}
+\graphicspath{{figures/}}
 
@@
 
-1. Overview of problem
+Hyperspectral imaging has emerged as a keystone technology in remote sensing where the ability to discern variations between high resolution spectra facilitates a plethora of critical applications from environmental monitoring and risk assessment to biodiversity convservation and sustainable agriculture.
+In recent years, many remote sensing platforms have been deployed with hyperspectral imaging payloads such as the Italian PRISMA mission launched in 2019 and the more recent German EnMAP launched in 2022 (\textcolor{red}{add citations}).
+Many future missions also plan to incorporate hyperal imaging capabilities such as the European Space Agency's CHIME which will include over $200$ bands spanning visible, near infrared (NIR), and short-wave infrared (SWIR) wavelengths (\textcolor{red}{add citations}).
+Additionally, the continued development of hyperspectral imaging technologies have resulted in a considerable reduction in size enabling their inclusion in the payloads of small unmanned aerial vehicles (UAV) (\textcolor{red}{add citations}).
+Despite these promising advances, the substantial increase in data volume generated by hyperspectral imagers as compared to their multi-spectral counterparts poses significant challenges to real-time analysis at scale.
 
-Hyperspectral imaging has emerged as a keystone technology in remote sensing where the ability to discern slight variations in high resolution spectra facilitates a plethora of critical applications from land use assessment...
-etc.
-- environmental monitoring, resource management, risk assessment, biodiversity conservation, sustainable agriculture,
-- Many hyperspectral remote sensing missions have already been launched with many more in development
-    - EnMAP (launched 2022) is german satellite with remote sensing imager
-    - PRISMA (launched 2019) is italian remote sensing satellite
-    - HISUI (launched 2019) is japanese HSI on international space station
-    - CHIME (future launch) - Copernicus Hyperspectral Imaging mission for the environment from the european space agency
-    - SBG (future launch) - surface biology and geology - NASA mission
-    - SHALOM (future launch) - Joint israeli-italian mission
-    - HyspIRI (future launch) - italian missions
-- Additionally, continued improvement to HSI technologies have dramatically reduced their size enabling their inclusion to the payloads of small unmanned aerial vehicles (UAV).
+Many approaches have been developed to make sense of hyperspectral images (HSI).
+For example, spectral indices such as the popular normalized difference vegetation index (NDVI) can be computed by taking ratios of spectral bands tailored to track specific reflectance characteristics (\textcolor{red}{add citations}).
+These indices have the advantage of being easy to compute but suffer from significant inter-instrument variability while ingoring most of the captured wavelength bins (\textcolor{red}{add citations}).
+An alternative approach is to pair HSI data with in-situ measurements to enable supervised models which map spectra directly to parameters of interest.
+However, this approach relies on serendipitous satellite overpasses above sensing sites to generate sufficient quantities of aligned data for model training and evaluation.
+For example, Aurin et al.
+combined data from over 30 years of oceanographic field campaigns with paired satellite imagery to develop robust models for the inversion of key water quality indicators such as colored dissolved organic matter (CDOM) \cite{aurin2018remote}.
+This approach can be accelerated by combining UAV-based hyperspectral imaging with rapid in-situ data collection using autonomous boats \cite{robot-team-1, robot-team-2}.
+Nevertheless, these supervised methods rely on \textit{a priori} knowledge of expected sources in order to select appropriate reference sensors for data collection.
+In light of these limitations, unsupervised methods are needed which can expediently extract source signatures from high dimensional HSI.
 
-- Key challenge: how to make sense of big data
-- Common approaches involve computation of spectral indices like the NDVI
-    - These are nice but can be challenging to compare between instruments due to varibility in spectral bands and sensor response
-    - They also throw away
-- A second common approach is to directly map spectral data to parameters of interest using in situ reference data 
-    - This is popular for water quality assessment where parameters like CDOM, etc. track water quality
-    - The key limitation is the availability of coincident reference data. This can requires decades of aligned observations to provide sufficient data volumes for statistical/machine learning models
-- In light of these difficulties, unsupervised methods which extract relevant information from HSI themselves can be advantageous
+The spatial resolution of hyperspectral imagers generally results in mixed HSI pixels comprised of many individual sources called endmembers.
+The task of unsupervised source identification using HSI data therefore comprises two steps: endmember extraction, and abundance estimation.
+Techniques such as vertex component analysis (VCA), the pixel purity index (PPI), and N-FINDR, solve this first task by identifying endmember spectra from HSI assuming the presense of some pure (unmixed) pixels (\textcolor{red}{add citations}).
+By further assuming a linear mixing model (LMM) whereby observed spectra are described by a linear combination of endmembers with non-negative abundances, HSI can be unmixed to obtain abundance maps using a variety of techniques such as constrained least-squares.
+
+Once extracted, HSI can be unmixed to estimate individual endmember abundances using a variety of techniques such as constrained least-squares optimization (\textcolor{red}{add citations}).
+Alternatively,
+
 - Reality: pixels captured by remote sensing platforms have spatial extents which typically cover multiple unique sources. 
@@
 - relevance to source identification in remote sensing, contaminant identication
-- relevance to source apportionment in air quality monitoring
 
@@
 \begin{figure}[H]
-\includegraphics[width=\columnwidth]{paper/figures/methods/gsm/gsm-diagram.pdf}
+\includegraphics[width=\columnwidth]{methods/gsm/gsm-diagram.pdf}
 \caption{Illustration of the GSM.
@@
 \begin{figure}[H]
-\includegraphics[width=\columnwidth]{paper/figures/methods/usgs/usgs-dataset.pdf}
+\includegraphics[width=\columnwidth]{methods/usgs/usgs-dataset.pdf}
 \caption{Synthetic dataset from USGS spectra. \label{fig:usgs-data}}
@@
 \begin{figure}[H]
-\includegraphics[width=\columnwidth]{paper/figures/methods/robot-team/robot-team-overview.pdf}
+\includegraphics[width=\columnwidth]{methods/robot-team/robot-team-overview.pdf}
 \caption{Robot team dataset. \label{fig:robotteam-data}}
@@
 \begin{figure}[H]
-\includegraphics[width=\columnwidth]{paper/figures/results/usgs/fit-comparison.pdf}
+\includegraphics[width=\columnwidth]{results/usgs/fit-comparison.pdf}
 \caption{Comparison of GSM against NMF on simulated USGS dataset. \label{fig:usgs-fits}}
@@
 \begin{figure}[H]
-\includegraphics[width=\columnwidth]{paper/figures/results/usgs/extracted-endmembers.png}
+\includegraphics[width=\columnwidth]{results/usgs/extracted-endmembers.png}
 \caption{Endmembers extracted using GSM for simulated USGS dataset with SNR$=20$. \label{fig:usgs-endmembers}}
@@
 \begin{figure}[H]
-\includegraphics[width=0.60\columnwidth]{paper/figures/results/robot-team/pca-variance.png}
+\includegraphics[width=0.60\columnwidth]{results/robot-team/pca-variance.png}
 \caption{Explained variance of PCA components for robot team dataset. A red horizontal line is superimposed on the plot indicating an explained variance of $1\%$. \label{fig:robot-team-pca}}
@@
 \centering
-\includegraphics[width=1.25\columnwidth]{paper/figures/results/robot-team/extracted-endmembers.pdf}
+\includegraphics[width=1.25\columnwidth]{results/robot-team/extracted-endmembers.pdf}
 \end{adjustwidth}
@@
 \centering
-\includegraphics[width=\columnwidth]{paper/figures/results/robot-team/endmember-abundances.pdf}
+\includegraphics[width=\columnwidth]{results/robot-team/endmember-abundances.pdf}
 \caption{Endmember distributions: \textbf{(a)} The distribution of abundance for the water class.
@@
 \centering
-\includegraphics[width=1.25\columnwidth]{paper/figures/results/robot-team/plume-evo.pdf}
+\includegraphics[width=1.25\columnwidth]{results/robot-team/plume-evo.pdf}
 \end{adjustwidth}
@@
 \begin{tabular}{@{}ll}
-MDPI & Multidisciplinary Digital Publishing Institute\\
-DOAJ & Directory of open access journals\\
-TLA & Three letter acronym\\
-LD & Linear dichroism
+PRISMA & Hyperspectral Precuror of the Application Mission \\
+EnMAP & Environmental Mapping and Analysis Program \\
+CHIME & Copernicus Hyperspectral Imaging Mission for the Environment \\
+NIR & Near Infrared \\
+SWIR & Short-wave Infrared \\
+UAV & Unmanned Aerial Vehicle \\
+HSI & Hyperspectral Image \\
+NDVI & Normalized Difference Vegetation Index \\
+VCA & Vertex Component Analysis\\
+PPI & Pixel Purity Index
 \end{tabular}
@@
 
-\bibliography{paper/references}
+\bibliography{./references.bib}
 \PublishersNote{}

### 2024-08-05 paper/paper.tex
updating text

--- 2024-07-30 paper/paper.tex
@@
 % Full title of the paper (Capitalized)
-\Title{Generative Simplex Mapping: Nonlinear Endmember Extraction and Unmixing for Hyperspectral Imagery}
+\Title{Generative Simplex Mapping: Nonlinear Endmember Extraction for Source Apportionment and Spectral Unmixing for Hyperspectral Imagery}
+% \Title{Generative Simplex Mapping: Nonlinear Endmember Extraction and Spectral Unmixing for Hyperspectral Imagery}
 
 % MDPI internal command: Title for citation in the left column
-\TitleCitation{Generative Simplex Mapping: Nonlinear Endmember Extraction and Unmixing for Hyperspectral Imagery}
+\TitleCitation{Generative Simplex Mapping: Nonlinear Endmember Extraction and Spectral Unmixing for Hyperspectral Imagery}
 
@@
 % Keywords
-\keyword{Endmember Extraction; Spectral Unmixing; Hyperspectral Imaging; Unsupervised Machine Learning}
+\keyword{Endmember Extraction; Spectral Unmixing; Hyperspectral Imaging; Unsupervised Machine Learning; Source Apportionment}
 
@@
 
-1. Overview of problem
+Hyperspectral imaging has emerged as a keystone technology in remote sensing where the ability to discern variations between high resolution spectra supports a plethora of critical applications such as environmental monitoring, biodiversity conservation, sustainable agriculture, and more.
+In recent years, many remote sensing platforms have been deployed with hyperspectral imaging payloads such as the Italian PRISMA mission launched in 2019, the German EnMAP launched in 2022, and recently NASA's PACE satellite launched in 2024 \cite{PRISMA-orig, EnMAP-orig, PACE-orig}.
+Many future missions also plan to incorporate hyperspectral imaging capabilities such as the European Space Agency's CHIME which will include over $200$ bands spanning visible, near infrared (NIR), and short-wave infrared (SWIR) wavelengths \cite{CHIME-orig}}.
+The continued development of hyperspectral imaging technology has also led to a considerable reduction in size enabling their inclusion in the payloads of small unmanned aerial vehicles (UAV) \cite{adao2017hyperspectral, arroyo2019implementation}.
+Despite the proliferation of hyperspectral imaging, the considerable increase in data volume associated with hyperspectral images (HSI) poses significant challenges to real-time analysis at scale.
 
-2. Traditional approaches for endmember extraction and spectral unmixing
+Many approaches have been developed to make sense of HSI data.
+For example, spectral indices like the popular normalized difference vegetation index (NDVI) can be computed by taking ratios of spectral bands tailored to track specific reflectance characteristics \cite{thenkabail-indices,thenkabail2018hyperspectral}.
+These indices have the advantage of being easy to compute but suffer from significant inter-instrument variability while ignoring most of the information captured in HSI spectra \cite{ndvi-variability}.
+An alternative approach is to pair HSI data with in-situ measurements to enable supervised models which map spectra directly to parameters of interest.
+However, this approach relies on serendipitous satellite overpasses above sensing sites to generate sufficient quantities of aligned data for model training and evaluation.
+For instance, Aurin et al.
+combined data from over 30 years of oceanographic field campaigns with paired satellite imagery to develop robust models for the inversion of key water quality indicators such as colored dissolved organic matter (CDOM) \cite{aurin2018remote}.
+This approach can be accelerated by combining UAV-based hyperspectral imaging with rapid in-situ data collection using autonomous boats \cite{robot-team-1, robot-team-2}.
+Nevertheless, these supervised methods rely on \textit{a priori} knowledge of expected sources in order to identify appropriate reference sensors for data collection.
+In light of these limitations, unsupervised methods are needed which can expediently extract source signatures from high dimensional HSI.
 
-3. Generative topographic mapping
+The spatial resolution of hyperspectral imagers generally results in  pixels with mixed signals from multiple sources called endmembers.
+The task of unsupervised source identification using HSI data therefore involves two steps: endmember extraction, and abundance estimation.
+Techniques such as vertex component analysis (VCA), the pixel purity index (PPI), and N-FINDR solve this first task by identifying endmember spectra from HSI assuming the presence of some pure (unmixed) pixels \cite{vca-orig, ppi-orig, N-FINDR-orig}.
+By further assuming a linear mixing model (LMM) whereby observed spectra are described by a linear combination of endmembers with non-negative abundances, HSI can then be unmixed using a variety of techniques such as constrained least-squares \cite{spectral-unmixing-orig, fcls-unmixing}.
+Among these methods Non-negative Matrix Factorization (NMF) is a widely used approach which extracts endmember spectra and unmixes abundances simultaneously via matrix factorization \cite{nmf-orig, unmixing-nmf-review, unmixing-nmf-review-2}.
+The update equations for NMF can be formulated into multiplicative updates which guarantee the non-negativity of endmember spectra and their associated abundances \cite{nmf-algorithms}.
+For this reason the continued development of new NMF varieties remains an active area of research.
 
-4. Proposed method: generative simplex mapping
+In realistic scenes multiple scattering and surface variability can easily challenge the assumption of linear mixing \cite{heylen2014review}.
+Water-based HSI specifically are prone to nonlinear mixing effects due to absorption features of dissolved and suspended substances, fluorescence of organic matter, and particulate scattering in turbid waters \cite{hsi-absorption, hsi-fluorescence, hsi-turibidity}.
+With the growing popularity of deep learning approaches in remote sensing, a variety of models based on autoencoder architectures have been introduced which enable nonlinear unmixing of HSI data \cite{non-negative-autoencoders,su2019daen,palsson2020convolutional,}.
+However, the complexity introduced by these models significantly impacts training time and decreases model interpretability.
+An ideal approach should enable both endmember extraction and nonlinear unmixing while accounting for spectral variability.
 
-5. Summary of paper structure
+The self-organizing map (SOM) is an unsupervised machine learning method which maps high-dimensional data onto a low-dimensional grid while preserving the topological relationships between data points \cite{kohonen-som-1}.
+This low dimensional representation provides a convenient way to visualize HSI data while the weight vectors for each SOM node can be interpreted as representative spectra \cite{cantero2004analysis, duran2007time,som-hsi}.
+If labelled reference spectra are available, the SOM can be used to enable semi-supervised labelling of HSI pixels \cite{riese2019supervised}.
+The SOM has also been shown to be effective for compression of HSIs acquired by a CubeSat \cite{som-satellite}.
+Despite these capabilities, the SOM does not offer a probabilistic interpretation and relies on a heuristic training procedure with hyperparameters that can be challenging to tune.
+To address these shortcomings, Bishop et al introduced the Generative Topographic Mapping (GTM), a probabilistic latent-variable model inspired by the SOM \cite{gtm-orig}.
+When the latent space is chosen to be two dimensional, the GTM can be used to visualize the distribution of HSI spectra while the mapping of latent-space nodes into the HSI data-space provides endmembers\cite{robot-team-gtm}.
+Unfortunately, the rectangular latent-space grid employed by the GTM does not directly translate into endmember abundances.
+Additionally, the expectation-maximization (EM) algorithm used to train the GTM does not guarantee non-negativity of GTM node spectra.
 
-\newpage
+In this paper, we introduce a new variant of the GTM dubbed Generative Simplex Mapping (GSM) which can extract endmember spectra and unmix nonlinear mixtures.
+By replacing the rectangular latent-space of the GTM with a gridded $n$-simplex, the vertices of the GSM can be immediately interpreted as endmembers corresponding to $n+1$ unique spectral signatures.
+The mapping from the latent-space to the HSI data-space models signal mixing while barycentric coordinates for the latent-space simplex estimate relative endmember abundances.
+Furthermore, by taking inspiration from the multiplicative updates of NMF, the GSM algorithm maintains non-negativity of resulting endmember spectra.
+If only linear mixing is present, the GSM algorithm drives nonlinear contributions to $0$.
+Prior distributions included for GSM model weights yield hyperparameters which can be tuned to control the smoothness of resulting spectra and the degree of nonlinear mixing applied.
 
-\section{Generative Simplex Mapping}
-\subsection{Method Derivation}
+The rest of the paper is structured as follows.
+Section~\ref{sec:gsm} describes the proposed method.
+Section~\ref{sec:experiments} and Section~\ref{sec:results} describe experiments using simulated and real HSI data to evaluate the GSM model.
+Section~\ref{sec:disucssion} discusses additional applications and extensions of the GSM to be explored in future work.
+Finally, Section~\ref{sec:conclusions} finishes the paper with some closing remarks.
 
+\section{Generative Simplex Mapping}\label{sec:gsm}
+
+In the original GTM formulation, data vectors $x$ (e.g.
+reflectance spectra) are described by latent variables $z$ mapped into the data-spae by a nonlinear function $\psi$.
+The data-space distribution is taken to be normal with precision parameter $\beta$ to account for measurement noise and spectral variability.
+The GSM follows uses same structure, that is
 \begin{equation}\label{eqn:data-space-distribution}
@@
 \end{equation}
+where $\mathbf{W}$ are weights which parameterize the mapping $\psi$.
 
+Assuming data are uniformly sampled from an embedded manifold in the data-space, the GTM models the latent-space using a rectangular grid with $K$-many nodes of equal prior probability.
+To adapt the GTM to describe endmember mixing, the GSM makes two key changes.
+The first is to replace the rectangular GTM grid with a gridded simplex having $N_v$ vertices.
+Barycentric coordinates for each node then describe the relative abundance of each endmember with simplex vertices corresponding to pure endmember spectra.
+The second change is to replace the equal prior probabilities with adaptive mixing coefficients $\pi_k$ to allow the GSM to model HSI with non-uniform mixing distributions.
+Together, this leads to a latent-space prior given by
 \begin{equation}\label{eqn:latent-distribution}
-    p(z) = \sum\limits_k^K \pi_k \, \delta(z - z_k) \quad\text{where}\quad  \sum_k^K\pi_k = 1
+    p(z) = \sum\limits_k^K \pi_k \, \delta(z - z_k) \quad\text{where}\quad  \sum_k^K\pi_k = 1.
 \end{equation}
 
+For a dataset containing $N$-many records, these definitions yield a log-likelihood function given by
 \begin{equation}\label{eqn:llh}
@@
 \end{equation}
+which can be maximized to obtain optimal values for model weights $\mathbf{W}$ and precision $\beta$.
+Rather than optimizing Eq~\ref{eqn:llh} directly, we instead choose a particular form for $\Psi$ to allow fitting the GSM via an EM algorithm.
 
-The activation function $\mathbf{\Phi}_{km} = \phi_m(z_k)$ is chosen to enable modeling linear and nonlinear mixing. For $m \leq N_v$, we take $\Phi_{km} = z_k$. For $m > N_v$, we set
+In the standard LMM model, the mapping $\Psi$ is given by $\psi(z;W) = \mathbf{W}z$ where the columns of $\mathbf{W}$ correspond to endmember spectra.
+To model nonlinear mixing, $z$ is replaced by the output of $M$-many activation functions such that $\Psi(z;W) = \mathbf{W}\phi(z)$.
+The activations applied to each GSM node can then be collected to form a matrix
+ $\mathbf{\Phi}_{km} = \phi_m(z_k)$.
+For $m \leq N_v$, we take $\Phi_{km} = z_k$ to account for linear mixing.
+The remaining  $M-N_v$ activations are computed using radial basis functions (RBF) with centers $\mu_m$ distributed throughout the simplex (but not at the vertices) so that
 \begin{equation}\label{eq:act-function}
     \mathbf{\Phi}_{km} = \begin{cases}
-        \frac{s - \lVert z_k - \mu_m \rVert}{s}; & \lVert z_k - \mu_m \rVert \leq s \\
+        \dfrac{s - \lVert z_k - \mu_m \rVert}{s}; & \lVert z_k - \mu_m \rVert \leq s \\
         0; & \lVert z_k - \mu_m \rVert > s
@@
 \end{equation}
-The activation function centers $\mu_m$ are selected to be greater than
+where $s$ is the spacing between RBF centers.
+In this form, the first $N_v$ columns of $\mathbf{W}$ correspond to endmember spectra while the remaining columns account for additional nonlinear effects.
+In the limit of linear mixing, the GSM training algorithm should therefore force $W_{dm} = 0$ for $m\geq N_v$.
+A visualization of the GSM model is shown in Fig~\ref{fig:gsm-diagram}.
 
-We include a prior distribution on model weights to enable
-\begin{align}\label{eq:weight-prior}
-\begin{split}
-    p(\mathbf{W} \mid \lambda_e, \lambda_w) &= \left(\frac{\lambda_e}{2\pi}\right)^{\frac{D\,N_v}{2}}\exp\left(-\frac{\lambda_e}{2}\sum\limits_d^D\sum\limits_{m=1}^{N_v}\mathbf{W}_{dm}^2 \right)\\ 
-    & \qquad \cdot \left(\frac{\lambda_w}{2\pi}\right)^{\frac{D(M-N_v)}{2}}\exp\left(-\frac{\lambda_w}{2}\sum\limits_d^D\sum_{m=N_v+1}^M \mathbf{W}_{dm}^2\right)
-\end{split}
-\end{align}
+\begin{figure}[H]
+\includegraphics[width=\columnwidth]{methods/gsm/gsm-diagram.pdf}
+\caption{Illustration of the GSM.
+The latent space consists of a grid of $K$-many points (green dots) distributed throughout a simplex with $N_v$ vertices.
+Barycentric coordinates of each node in the simplex correspond to the relative abundance of $N_v$-many unique sources.
+Here, $N_v=3$ has been chosen for illustrative purposes.
+Nodes are mapped into the data space via the map $\psi(z)$ utilizing $M$-many radially symmetric basis functions (red).
+Spectral variability is estimated via the precision parameter $\beta$ shown here in the data space as a light blue band around the spectrum given by $\psi(z)$.}
+\label{fig:gsm-diagram}
+\end{figure}
 
+To further constrain the model, we introduce prior distributions on the weights $\mathbf{W}$.
+For $m\leq N_v$ we take $W_{dm}\sim\mathcal{N}(0, \lambda_e^{-1})$ corresponding to a zero-mean Gaussian with variance $\lambda_e^{-1}$.
+For $m>N_v$ we use a zero-mean Laplace distribution, $W_{dm}\sim\dfrac{\lambda_w}{2}\exp(-\lambda_w\lvert W_{dm}\rvert)$, with scale parameter $\lambda_w^{-1}$.
+Under these choices, $\lambda_e$ corresponds to $L_2$ regularization on endmember spectra while $\lambda_w$ corresponds to $L_1$ regularization on the nonlinear activations.
+In other words, $\lambda_e$ governs the smoothness of the resulting endmembers while $\lambda_w$ encourages sparsity for the nonlinear contributions.
+
+An EM algorithm for the GSM model can now be formulated as follows.
+Suppose that we have current estimates for the model weights $\mathbf{W}$, mixing coefficients $\pi_k$, and precision parameter $\beta$.
+During the expectation step we compute the posterior probabilities, that is, the responsibility of each GSM node for each spectrum in the dataset:
 \begin{equation}\label{eq:responsibility}
-    \mathbf{R}_{kn}  = p(z_k \mid x_n, \mathbf{W}, \beta) = \dfrac{p(x_n \mid z_k, \mathbf{W}, \beta)}{\sum\limits_{k'}^K p(x_n \mid z_{k'}, \mathbf{W}, \beta)}
+    \mathbf{R}_{kn}  = p(z_k \mid x_n, \mathbf{W}, \beta) = \dfrac{\pi_k \, p(x_n \mid z_k, \mathbf{W}, \beta)}{\sum\limits_{k'}^K \pi_{k'} \, p(x_n \mid z_{k'}, \mathbf{W}, \beta)}.
 \end{equation}
 
-Penalized complete-data log likelihood
-\begin{adjustwidth}{-\extralength}{0cm}
+Then, holding the $\mathbf{R}_{kn}$ fixed, we optimize the expectation of the penalized complete-data log-likelihood given by
+%\begin{adjustwidth}{-\extralength}{0cm}
 \begin{equation}\label{eq:complete-data-llh}
-    Q = \sum\limits_n^N\sum\limits_k^K \mathbf{R}_{kn} \left(\ln\pi_k + \frac{D}{2}\ln\left(\frac{\beta}{2\pi}\right) - \frac{\beta}{2}\sum\limits_d^D\left(\sum\limits_m^M \mathbf{W}_{dm}\mathbf{\Phi}_{km} - \mathbf{X}_{nd}\right)^2\right) + \frac{MD}{2}\ln\left(\frac{\lambda}{2\pi}\right) - \frac{\lambda}{2}\sum\limits_d^D \sum\limits_m^M \mathbf{W}_{dm}^2
+\begin{aligned}
+    Q &= \sum\limits_n^N\sum\limits_k^K \mathbf{R}_{kn} \left(\ln\pi_k + \frac{D}{2}\ln\left(\frac{\beta}{2\pi}\right) - \frac{\beta}{2}\sum\limits_d^D\left(\sum\limits_m^M \mathbf{W}_{dm}\mathbf{\Phi}_{km} - \mathbf{X}_{nd}\right)^2\right) \\
+    &\qquad + \frac{MD}{2}\ln\left(\frac{\lambda}{2\pi}\right) - \frac{\lambda}{2}\sum\limits_d^D \sum\limits_m^M \mathbf{W}_{dm}^2
+\end{aligned}
 \end{equation}
-\end{adjustwidth}
+%\end{adjustwidth}
 
@@
 
+\begin{equation}\label{eq:W-update}
+    W_{dm}^\text{new} = \dfrac{\left(X^TR^T\Phi\right)_{dm}}{\left(W\Phi^TG\Phi + \Lambda\right)_{dm}} 
+\end{equation}
+
 - multiplicative updated to guarantee non-negativity (Generalized EM procedure)
 - sum-to-one constraint is guaranteed via representation of latent space with barycentric coordinates.
-
-\begin{figure}[H]
-\includegraphics[width=\columnwidth]{paper/figures/methods/gsm/gsm-diagram.pdf}
-\caption{Illustration of the GSM.
-Points in the latent space (green dots) are uniformly distributed at the nodes of a regular grid on the n-simplex whose barycentric coordinates are to be interpreted as the relative abundance of n-many unique sources $z_1, ...
-z_{N_v}$.
-Here, $N_v=3$ has been chosen for illustrative purposes.
-These nodes are mapped into (spectral) data space via the mapping $\psi(z)$ utilizing $M$-many radially symmetric basis functions shown in red.
-The trained GSM also enables modelling of spectral variability via the precision parameter $\beta$ shown here as a light blue band around the estimated spectra in the data space.\label{fig:gsm-diagram}}
-\end{figure}
 
@@
 
-\section{Experiments}
+\section{Experiments}\label{sec:experiments}
 \subsection{Linear Mixing: Comparison to NMF}
@@
 
-\section{Results}
+\section{Results}\label{sec:results}
 \subsection{Linear Mixing}
@@
 
-\section{Discussion}
+\section{Discussion}\label{sec:disucssion}
 
@@
     - Further note that both NMF and GSM do not rely on assumption of pure pixels (unlike VCA and other methods) 
-    - GSM also allows estimation of endmember variability via the precision parameter $\beta$ i.e. nodes in data space follow Gaussian distribution with $\sigma = \sqrt{\beta^{-1}}$
+    - GSM also allows estimation of endmember variability via the precision parameter $\beta$ i.e. nodes in data-space follow Gaussian distribution with $\sigma = \sqrt{\beta^{-1}}$
 
@@
     - Applications to source apportionment, e.g.
-linear GSM can be used for standard source apportionment problems (no reactions) and nonlinear GSM may be useful for modeling scenarios with aditional complications due to reactions and transfport from source to receptor.
+linear GSM can be used for standard source apportionment problems (no reactions) and nonlinear GSM may be useful for modeling scenarios with additional complications due to reactions and during transport from source to receptor.
 
-\section{Conclusions}
+\section{Conclusions}\label{sec:conclusions}
 
@@
 \begin{tabular}{@{}ll}
-MDPI & Multidisciplinary Digital Publishing Institute\\
-DOAJ & Directory of open access journals\\
-TLA & Three letter acronym\\
-LD & Linear dichroism
+PRISMA & Hyperspectral Precuror of the Application Mission \\
+EnMAP & Environmental Mapping and Analysis Program \\
+PACE & Plankton, Aerosol, Cloud, ocean Ecosystem \\
+CHIME & Copernicus Hyperspectral Imaging Mission for the Environment \\
+NIR & Near Infrared \\
+SWIR & Short-wave Infrared \\
+UAV & Unmanned Aerial Vehicle \\
+HSI & Hyperspectral Image \\
+NDVI & Normalized Difference Vegetation Index \\
+VCA & Vertex Component Analysis\\
+PPI & Pixel Purity Index \\ 
+LMM & Linear Mixing Model \\
+NMF & Non-negative Matrix Factorization \\ 
+SOM & Self-Organizing Map \\ 
+GTM & Generative Topographic Mapping  \\ 
+EM & Expectation-Maximization \\
+GSM & Generative Simplex Mapping \\ 
+RBF & Radial Basis Function
 \end{tabular}
@@
 
-\bibliography{paper/references}
+\bibliography{references.bib}
 \PublishersNote{}

### 2024-08-05 paper/paper.tex
added more updates to paper

@@
 
-For a dataset containing $N$-many records, these definitions yield a log-likelihood function given by
+For a dataset containing $N$-many records, these definitions yield a log likelihood function given by
 \begin{equation}\label{eqn:llh}
@@
 \begin{equation}\label{eq:act-function}
-    \mathbf{\Phi}_{km} = \begin{cases}
+    \Phi_{km} = \begin{cases}
         \dfrac{s - \lVert z_k - \mu_m \rVert}{s}; & \lVert z_k - \mu_m \rVert \leq s \\
@@
 In this form, the first $N_v$ columns of $\mathbf{W}$ correspond to endmember spectra while the remaining columns account for additional nonlinear effects.
-In the limit of linear mixing, the GSM training algorithm should therefore force $W_{dm} = 0$ for $m\geq N_v$.
+For linear mixing, the GSM training algorithm should therefore drive $W_{dm}$ to $0$ for $m\geq N_v$.
 A visualization of the GSM model is shown in Fig~\ref{fig:gsm-diagram}.
@@
 For $m>N_v$ we use a zero-mean Laplace distribution, $W_{dm}\sim\dfrac{\lambda_w}{2}\exp(-\lambda_w\lvert W_{dm}\rvert)$, with scale parameter $\lambda_w^{-1}$.
-Under these choices, $\lambda_e$ corresponds to $L_2$ regularization on endmember spectra while $\lambda_w$ corresponds to $L_1$ regularization on the nonlinear activations.
+Under these choices $\lambda_e$ corresponds to $L_2$ regularization on endmember spectra while $\lambda_w$ corresponds to $L_1$ regularization on the nonlinear activations.
 In other words, $\lambda_e$ governs the smoothness of the resulting endmembers while $\lambda_w$ encourages sparsity for the nonlinear contributions.
@@
 \begin{equation}\label{eq:responsibility}
-    \mathbf{R}_{kn}  = p(z_k \mid x_n, \mathbf{W}, \beta) = \dfrac{\pi_k \, p(x_n \mid z_k, \mathbf{W}, \beta)}{\sum\limits_{k'}^K \pi_{k'} \, p(x_n \mid z_{k'}, \mathbf{W}, \beta)}.
+    R_{kn}  = p(z_k \mid x_n, \mathbf{W}, \beta) = \dfrac{\pi_k \, p(x_n \mid z_k, \mathbf{W}, \beta)}{\sum\limits_{k'}^K \pi_{k'} \, p(x_n \mid z_{k'}, \mathbf{W}, \beta)}.
+\end{equation}
+For the maximization step, we consider the expectation of the penalized complete-data log likelihood given by
+\begin{equation}\label{eq:complete-data-llh}
+\begin{aligned}
+    Q &= \sum_n^N\sum_k^K R_{kn} \left(\ln\pi_k + \frac{D}{2}\ln\left(\frac{\beta}{2\pi}\right) - \frac{\beta}{2}\sum_d^D\left(\sum_m^M W_{dm}\Phi_{km} - X_{nd}\right)^2\right) \\ 
+    &\qquad + \frac{N_vD}{2}\ln\left(\frac{\lambda_e}{2\pi}\right) - \frac{\lambda}{2}\sum_d^D \sum_{m=1}^{N_v} W_{dm}^2  \\ 
+    &\qquad + (M-N_v)D\ln\left(\dfrac{\lambda_w}{2}\right) - \lambda_w\sum_d^D\sum_{m=N_v+1}^{M} W_{dm}
+\end{aligned}
+\end{equation}
+where $X_{nd}$ is the $d$-th component of the $n$-th spectrum in the dataset.
+Eq~\ref{eq:complete-data-llh} is then maximized with respect to $\pi_k$, $\beta$, and $\mathbf{W}$ to obtain new parameter values.
+A detailed overview of the EM procedure can be found in Bishop \cite{bishop-prml}.
+
+For $\pi_k$, optimization can be performed using Lagrange multipliers to maintain the condition that $\sum_k\pi_k=1$. Doing so yields
+\begin{equation}\label{eq:pi-update}
+    \pi_k^{\text{new}}  = \frac{1}{N}\sum_n R_{kn}
 \end{equation}
 
-Then, holding the $\mathbf{R}_{kn}$ fixed, we optimize the expectation of the penalized complete-data log-likelihood given by
-%\begin{adjustwidth}{-\extralength}{0cm}
-\begin{equation}\label{eq:complete-data-llh}
-\begin{aligned}
-    Q &= \sum\limits_n^N\sum\limits_k^K \mathbf{R}_{kn} \left(\ln\pi_k + \frac{D}{2}\ln\left(\frac{\beta}{2\pi}\right) - \frac{\beta}{2}\sum\limits_d^D\left(\sum\limits_m^M \mathbf{W}_{dm}\mathbf{\Phi}_{km} - \mathbf{X}_{nd}\right)^2\right) \\
-    &\qquad + \frac{MD}{2}\ln\left(\frac{\lambda}{2\pi}\right) - \frac{\lambda}{2}\sum\limits_d^D \sum\limits_m^M \mathbf{W}_{dm}^2
-\end{aligned}
+Optimization of Eq~\ref{eq:complete-data-llh} with respect to $\mathbf{W}$ leads to a linear system which can be solved using standard numerical methods.
+However, in this form we can not guarantee the non-negativity of $\mathbf{W}$.
+We therefore take inspiration from the multiplicative updates for NMF introduced by Lee and Seung \cite{nmf-orig}.
+A gradient-based update for $\mathbf{W}$ would take the form
+\begin{equation}
+    \mathbf{W}_{\text{new}} = \mathbf{W} + \eta\frac{\partial Q}{\partial \mathbf{W}}
 \end{equation}
-%\end{adjustwidth}
+for some learning rate $\eta$. Differentiating, we obtain
+\begin{equation}
+    \frac{\partial Q}{\partial \mathbf{W}} = -\beta \mathbf{W}\mathbf{\Phi}^T\mathbf{G}\mathbf{\Phi} - \mathbf{\Lambda} + \beta \mathbf{X}^T\mathbf{R}^T\mathbf{\Phi}
+\end{equation}
+where $\mathbf{G}$ is a diagonal matrix with $G_{kk} = \sum_n R_{kn}$ and $\mathbf{\Lambda}$ is given by 
+\begin{equation}
+    \Lambda_{dm} = \begin{cases}
+        \lambda_e W_{dm}; & m \leq N_v \\ 
+        \lambda_w; & m > N_v
+    \end{cases}
+\end{equation}
+If we allow for individual learning rates $\eta_{dm}$ for each element of $\mathbf{W}$, then choosing 
+\begin{equation}
+    \eta_{dm} = \frac{W_{dm}}{\left(\beta \mathbf{W}\mathbf{\Phi}^T\mathbf{G}\mathbf{\Phi}\right)_{dm} + \Lambda_{dm}}
+\end{equation}
+results in update equations for elements of $\mathbf{W}$ given by
+\begin{equation}\label{eq:W-update}
+    W_{dm}^{new}  = W_{dm} \cdot \dfrac{\left(\beta \mathbf{X}^T\mathbf{R}^T\mathbf{\Phi}\right)_{dm}}{\left(\beta \mathbf{W}\mathbf{\Phi}^T\mathbf{G}\mathbf{\Phi}\right)_{dm} + \Lambda_{dm}}
+\end{equation}
+From Eq~\ref{eq:W-update}, it is clear that we are multiplying $W_{dm}$ by non-negative values and therefore, $W_{dm}$ remains non-negative.
 
-\begin{equation}\label{eq:pi-update}
-    \pi_k^{\text{new}}  = \frac{1}{N}\sum_n \mathbf{R}_{kn}
+Finally, optimizing Eq~\ref{eq:complete-data-llh} with respect to $\beta$ yields
+\begin{equation}\label{eq:beta-update}
+    \frac{1}{\beta^{\text{new}}}  = \frac{1}{ND}\sum\limits_n^N\sum\limits_k^K R_{kn}\lVert \psi(z_k; \mathbf{W}) - x_n \rVert^2.
 \end{equation}
-
-\begin{equation}\label{eq:beta-update}
-    \frac{1}{\beta^{\text{new}}}  = \frac{1}{ND}\sum\limits_n^N\sum\limits_k^K \mathbf{R}_{kn}\lVert \psi(z_k; \mathbf{W}) - x_n \rVert^2
-\end{equation}
-
-\begin{equation}\label{eq:W-update}
-    W_{dm}^\text{new} = \dfrac{\left(X^TR^T\Phi\right)_{dm}}{\left(W\Phi^TG\Phi + \Lambda\right)_{dm}} 
-\end{equation}
-
-- multiplicative updated to guarantee non-negativity (Generalized EM procedure)
-- sum-to-one constraint is guaranteed via representation of latent space with barycentric coordinates.
+The expectation and maximization steps are repeated in turn until $Q$ converges to a predetermined tolerance level.
 

### 2024-08-06 paper/paper.tex
updates to experiment section

@@
 To address these shortcomings, Bishop et al introduced the Generative Topographic Mapping (GTM), a probabilistic latent-variable model inspired by the SOM \cite{gtm-orig}.
-When the latent space is chosen to be two dimensional, the GTM can be used to visualize the distribution of HSI spectra while the mapping of latent-space nodes into the HSI data-space provides endmembers\cite{robot-team-gtm}.
-Unfortunately, the rectangular latent-space grid employed by the GTM does not directly translate into endmember abundances.
+When the latent space is chosen to be two dimensional, the GTM can be used to visualize the distribution of HSI spectra while the mapping of latent space nodes into the HSI data space provides endmembers\cite{robot-team-gtm}.
+Unfortunately, the rectangular latent space grid employed by the GTM does not directly translate into endmember abundances.
 Additionally, the expectation-maximization (EM) algorithm used to train the GTM does not guarantee non-negativity of GTM node spectra.
@@
 In this paper, we introduce a new variant of the GTM dubbed Generative Simplex Mapping (GSM) which can extract endmember spectra and unmix nonlinear mixtures.
-By replacing the rectangular latent-space of the GTM with a gridded $n$-simplex, the vertices of the GSM can be immediately interpreted as endmembers corresponding to $n+1$ unique spectral signatures.
-The mapping from the latent-space to the HSI data-space models signal mixing while barycentric coordinates for the latent-space simplex estimate relative endmember abundances.
+By replacing the rectangular latent space of the GTM with a gridded $n$-simplex, the vertices of the GSM can be immediately interpreted as endmembers corresponding to $n+1$ unique spectral signatures.
+The mapping from the latent space to the HSI data space models signal mixing while barycentric coordinates for the latent space simplex estimate relative endmember abundances.
 Furthermore, by taking inspiration from the multiplicative updates of NMF, the GSM algorithm maintains non-negativity of resulting endmember spectra.
@@
 
-In the original GTM formulation, data vectors $x$ (e.g.
-reflectance spectra) are described by latent variables $z$ mapped into the data-spae by a nonlinear function $\psi$.
-The data-space distribution is taken to be normal with precision parameter $\beta$ to account for measurement noise and spectral variability.
-The GSM follows uses same structure, that is
+In the original GTM formulation, data vectors $x$ (reflectance spectra) are described by latent variables $z$ mapped into the data space by a nonlinear function $\psi$.
+The data space distribution is taken to be normal with precision parameter $\beta$ to account for measurement noise and spectral variability.
+The GSM uses same structure, that is
 \begin{equation}\label{eqn:data-space-distribution}
@@
 \end{equation}
-where $\mathbf{W}$ are weights which parameterize the mapping $\psi$.
+where $\mathbf{W}$ are model weights which parameterize the mapping $\psi$.
 
-Assuming data are uniformly sampled from an embedded manifold in the data-space, the GTM models the latent-space using a rectangular grid with $K$-many nodes of equal prior probability.
+Assuming data are uniformly sampled from an embedded manifold in the data space, the GTM models the latent space using a rectangular grid with $K$-many nodes of equal prior probability.
 To adapt the GTM to describe endmember mixing, the GSM makes two key changes.
 The first is to replace the rectangular GTM grid with a gridded simplex having $N_v$ vertices.
-Barycentric coordinates for each node then describe the relative abundance of each endmember with simplex vertices corresponding to pure endmember spectra.
+Barycentric coordinates for each node then describe the relative abundance of each endmember with each vertex corresponding to a pure endmember.
 The second change is to replace the equal prior probabilities with adaptive mixing coefficients $\pi_k$ to allow the GSM to model HSI with non-uniform mixing distributions.
-Together, this leads to a latent-space prior given by
+Together, this leads to a latent space prior distribution given by
 \begin{equation}\label{eqn:latent-distribution}
@@
 \begin{equation}\label{eqn:llh}
-    \mathcal{L}(\mathbf{W}, \beta) = \sum\limits_n^N \ln \left( \sum\limits_k^K \pi_k \, p(x_n \mid z_k, \mathbf{W}, \beta) \right)
+    \mathcal{L} = \sum\limits_n^N \ln \left( \sum\limits_k^K \pi_k \, p(x_n \mid z_k, \mathbf{W}, \beta) \right)
 \end{equation}
-which can be maximized to obtain optimal values for model weights $\mathbf{W}$ and precision $\beta$.
-Rather than optimizing Eq~\ref{eqn:llh} directly, we instead choose a particular form for $\Psi$ to allow fitting the GSM via an EM algorithm.
+which can be maximized to obtain optimal values for model weights $\mathbf{W}$, mixing coefficients $\pi_k$, and precision $\beta$.
+Rather than optimizing Eq~\ref{eqn:llh} directly, we instead choose a particular form for $\psi$ to allow fitting the GSM via an EM algorithm.
 
-In the standard LMM model, the mapping $\Psi$ is given by $\psi(z;W) = \mathbf{W}z$ where the columns of $\mathbf{W}$ correspond to endmember spectra.
-To model nonlinear mixing, $z$ is replaced by the output of $M$-many activation functions such that $\Psi(z;W) = \mathbf{W}\phi(z)$.
-The activations applied to each GSM node can then be collected to form a matrix
- $\mathbf{\Phi}_{km} = \phi_m(z_k)$.
-For $m \leq N_v$, we take $\Phi_{km} = z_k$ to account for linear mixing.
-The remaining  $M-N_v$ activations are computed using radial basis functions (RBF) with centers $\mu_m$ distributed throughout the simplex (but not at the vertices) so that
+In the standard LMM model, the mapping $\psi$ is given by $\psi(z;W) = \mathbf{W}z$ where the columns of $\mathbf{W}$ correspond to endmember spectra.
+To model nonlinear mixing, $z$ is replaced by the output of $M$-many activation functions such that $\psi(z;W) = \mathbf{W}\phi(z)$.
+The activations applied to each GSM node can then be collected to form a matrix with elements $\Phi_{km} = \phi_m(z_k)$.
+For $m \leq N_v$, we take $\Phi_{km} = z_k$ to model linear mixing.
+The remaining  $M-N_v$ activations are computed using radial basis functions (RBF) with centers $\mu_m$ distributed throughout the simplex (but not at the vertices) with
 \begin{equation}\label{eq:act-function}
@@
 Optimization of Eq~\ref{eq:complete-data-llh} with respect to $\mathbf{W}$ leads to a linear system which can be solved using standard numerical methods.
-However, in this form we can not guarantee the non-negativity of $\mathbf{W}$.
+However, in this form we can not guarantee the non-negativity of $\mathbf{W}$ required to describe reflectance spectra.
 We therefore take inspiration from the multiplicative updates for NMF introduced by Lee and Seung \cite{nmf-orig}.
-A gradient-based update for $\mathbf{W}$ would take the form
+A standard gradient-based update for $\mathbf{W}$ would take the form
 \begin{equation}
@@
 \end{equation}
-for some learning rate $\eta$. Differentiating, we obtain
+for some learning rate $\eta$. Therefore, we differentiate to obtain
 \begin{equation}
@@
 \end{equation}
-results in update equations for elements of $\mathbf{W}$ given by
+results in a multiplicative update rule given by
 \begin{equation}\label{eq:W-update}
@@
 \end{equation}
-From Eq~\ref{eq:W-update}, it is clear that we are multiplying $W_{dm}$ by non-negative values and therefore, $W_{dm}$ remains non-negative.
+From Eq~\ref{eq:W-update}, it is clear that we are multiplying $W_{dm}$ by strictly non-negative values, and therefore, non-negative $W_{dm}$ will remain so during each update.
+This update can also be repeated multiple times during each M-step to accelerate convergence.
 
-Finally, optimizing Eq~\ref{eq:complete-data-llh} with respect to $\beta$ yields
+Lastly, optimizing Eq~\ref{eq:complete-data-llh} with respect to $\beta$ yields
 \begin{equation}\label{eq:beta-update}
@@
 \end{equation}
-The expectation and maximization steps are repeated in turn until $Q$ converges to a predetermined tolerance level.
 
-\subsection{Implementation Details}
+To train a GSM model, weights $\mathbf{W}$ are randomly initialized to positive values.
+The mixing coefficients are initially set to $\pi_k = 1/K$.
+Finally, the precision parameter $\beta$ is initialized to the variance of the $(N_v+1)$-th principal component.
+After initialization, the expectation and maximization steps are repeated in turn until $Q$ converges to a predetermined tolerance level.
+For large $N_v$ we note that generating a regular grid within the simplex becomes cumbersome as the number grid nodes scales as ${k + N_v - 2 \choose N_v - 1}$ for $k$ nodes per edge.
+An alternative approach is to randomly sample points within the simplex to obtain a total of $K$ nodes.
+A Dirichlet distribution
+\begin{equation}
+    p(z_1,...z_{N_v}) = \frac{\Gamma(\sum_i^{N_v}\alpha_i)}{\prod_{i}^{N_v}\Gamma(\alpha_i)} \prod_{i}^{N_v} z_i^{\alpha_i-1}
+\end{equation}
+with all $\alpha_i=1$ can be used to uniformly sample within the simplex.
+Since the mixing coefficients $\pi_k$ are adaptive, variability in node separation should not significantly impact the resulting GSM.
 
-The mixing coefficients are initially chosen so that $\pi_k = \frac{1}{K}$.
+The probabilistic form of the GSM means that a variety of information criteria can be used to evaluate the model fits. In this paper we consider two metrics, the Bayesian Information Criterion (BIC),
+\begin{equation}
+   \text{BIC} = P\ln(N) - 2\mathcal{L},
+\end{equation}
+and the Akaike Information Criterion (AIC),
+\begin{equation}
+    \text{AIC} = 2P - 2\mathcal{L}
+\end{equation}
+where $P$ is the total number of model parameters and $\mathcal{L}$ is the log likelihood from Eq~\ref{eqn:llh}.
 
-- Linear Method
+The map $\psi$ provides the representation of each node $z_k$ in the data space.
+Importantly, applying $\psi$ to each vertex extracts endmember spectra from the GSM.
+Slices $R_{\left[:,n\right]}$ of the matrix $\mathbf{R}$ define the responsibility of each latent node $z_k$ for the $n$-th spectrum $x_n$ in the dataset.
+Therefore once the GSM has been trained, $\mathbf{R}$ can be used to unmix endmember abundances by representing each record in the latent space via the mean:
+\begin{equation}
+    \hat{z}_n = \sum_k^K R_{kn}z_k.
+\end{equation}
 
-- Nonlinear Method
-
-- Mixed Method
+A freely available implementation of the GSM is accessible at \cite{gtm-code}.
+The code is written in the Julia programming language and follows the Machine Learning in Julia (MLJ) common interface \cite{bezanson2012julia, blaom2020mlj}.
 
 \section{Experiments}\label{sec:experiments}
+
 \subsection{Linear Mixing: Comparison to NMF}
-- describe USGS data
-- describe NMF method (for comparisson)
-- describe evaluation criteria
+
+To illustrate the effectiveness of the GSM, we first demonstrate its ability to model simple linear mixing by driving nonlinear weights to zero during the fitting process.
+To this end, a synthetic dataset comprising linear mixtures of three sample spectra from the U.S.
+Geological Survey digital spectral library was generated by sampling $1000$ abundance vectors from a Dirichlet distribution with $\alpha_1=\alpha_2=\alpha_3=1/3$ \cite{usgs-spectra}.
+These known spectra and abundances are visualized in Figure~\ref{fig:usgs-data}
 
@@
 \includegraphics[width=\columnwidth]{methods/usgs/usgs-dataset.pdf}
-\caption{Synthetic dataset from USGS spectra. \label{fig:usgs-data}}
+\caption{Synthetic dataset fromed from USGS spectra.
+\textbf{(a)} spectra from the USGS spectral database used as the ground truth endmembers.
+These spectra were selected to match \cite{vca-orig}.
+\textbf{(b)} The distribution of abundances used in the dataset.
+Samples were generated using a Dirichlet distribution with $\alpha_1=\alpha_2=\alpha_3=1/3$.}
+\label{fig:usgs-data}
 \end{figure}
+
+Zero-mean Gaussian noise was added to the data to yield $9$ datasets with signal-to-noise ratios (SNR) ranging from $0$ to $\infty$ to examine the impact of random noise on GSM performance.
+For each of these datasets, a GSM was trained using $25$ nodes per edge with $\lambda_e$ set to the default value of $0.01$ and fixed to $\lambda_w = 100$.
+The large value of $\lambda_w$ was chosen to encourage the GSM to prefer models with limited nonlinear mixing.
+
+For comparison, NMF models were also trained on each dataset as this method is both highly popular and does not include the pure-pixel assumption common to other methods like VCA and PPI.
+Countless variations on the original NMF method have been introduced into the literature with one review identifying more than $100$ distinct NMF variations \cite{unmixing-nmf-review}.
+For the purpose of evaluating the GSM, we considered the standard $\ell_2$ and KL-divergence formulations introduced by Lee and Seung \cite{nmf-algorithms}.
+We also consider the robust $\ell_{2,1}$ NMF as introduced by Kong et al \cite{nmf-l21}.
+
+For comparison, three versions of NMF were also trained on each dataset. NMF was used as it is a highly popular method that does not include the pure-pixel assumption of other models
 
@@
     - Further note that both NMF and GSM do not rely on assumption of pure pixels (unlike VCA and other methods) 
-    - GSM also allows estimation of endmember variability via the precision parameter $\beta$ i.e. nodes in data-space follow Gaussian distribution with $\sigma = \sqrt{\beta^{-1}}$
+    - GSM also allows estimation of endmember variability via the precision parameter $\beta$ i.e. nodes in data space follow Gaussian distribution with $\sigma = \sqrt{\beta^{-1}}$
 

### 2024-08-06 paper/paper.tex
finished up methods section

@@
 In recent years, many remote sensing platforms have been deployed with hyperspectral imaging payloads such as the Italian PRISMA mission launched in 2019, the German EnMAP launched in 2022, and recently NASA's PACE satellite launched in 2024 \cite{PRISMA-orig, EnMAP-orig, PACE-orig}.
-Many future missions also plan to incorporate hyperspectral imaging capabilities such as the European Space Agency's CHIME which will include over $200$ bands spanning visible, near infrared (NIR), and short-wave infrared (SWIR) wavelengths \cite{CHIME-orig}}.
+Many future missions also plan to incorporate hyperspectral imaging capabilities such as the European Space Agency's CHIME which will include over $200$ bands spanning visible, near infrared (NIR), and short-wave infrared (SWIR) wavelengths \cite{CHIME-orig}.
 The continued development of hyperspectral imaging technology has also led to a considerable reduction in size enabling their inclusion in the payloads of small unmanned aerial vehicles (UAV) \cite{adao2017hyperspectral, arroyo2019implementation}.
@@
 Geological Survey digital spectral library was generated by sampling $1000$ abundance vectors from a Dirichlet distribution with $\alpha_1=\alpha_2=\alpha_3=1/3$ \cite{usgs-spectra}.
-These known spectra and abundances are visualized in Figure~\ref{fig:usgs-data}
+These known spectra and abundances are visualized in Figure~\ref{fig:usgs-data}.
 
@@
 
-For comparison, NMF models were also trained on each dataset as this method is both highly popular and does not include the pure-pixel assumption common to other methods like VCA and PPI.
+For comparison, NMF models were also trained on each dataset as this method is both highly popular and does not include the pure-pixel assumption common to other technqieus like VCA and PPI.
 Countless variations on the original NMF method have been introduced into the literature with one review identifying more than $100$ distinct NMF variations \cite{unmixing-nmf-review}.
 For the purpose of evaluating the GSM, we considered the standard $\ell_2$ and KL-divergence formulations introduced by Lee and Seung \cite{nmf-algorithms}.
-We also consider the robust $\ell_{2,1}$ NMF as introduced by Kong et al \cite{nmf-l21}.
+We also consider the robust $\ell_{2,1}$ NMF as described by Kong et al \cite{nmf-l21}.
 
-For comparison, three versions of NMF were also trained on each dataset. NMF was used as it is a highly popular method that does not include the pure-pixel assumption of other models
+Four metrics were used to compare model performance.
+For endmember extraction the mean spectral angle and mean RMSE between true endmembers $\rho_i$ and extracted endmembers $\hat{\rho}_i$ were computed where the spectral angle for the $i$-th endmember is defined as
+\begin{equation}
+    \theta(\rho_i, \hat{\rho}_i) =  \arcos\left( \dfrac{\langle \rho_i, \hat{\rho}_i \rangle}{\lVert \rho_i \rVert \cdot \lVert \hat{\rho}_i \rVert}\right)
+\end{equation}
+and the RMSE for the $i$-th endmember is
+\begin{equation}
+    \text{RMSE}(\rho_i, \hat{\rho}_i) = \sqrt{\frac{1}{D-1}\sum_d^D\left(\rho_i(\lambda_d) - \hat{\rho}_i(\lambda_d) \right)^2}.
+\end{equation}
+Abundance estimation was similarly evaluated using the Mean RMSE between true abundance and estimated abundance values for each endmember.
+Finally, the reconstruction RMSE, that is, the RMSE computed between the original dataset and the dataset reconstructed via the extracted endmembers and their associated abundances was computed.
+This provides a model-agnostic criterion to guarantee that each model sufficiently converged during training.
 
 \subsection{Nonlinear Mixing: Water Contaminant Identification}
-- describe robot team data collection and processing
-- describe
 
+To assess the ability of the GSM to unmix realistic scenes likely to involve nonlinear mixing effects, we consider a dataset of real HSI collected in Montague, North Texas on 9 December 2020.
+A Freefly Alta-X autonomous quadcopter was used as the UAV platform and was equipped with a Resonon Pika XC2 visible+near-infrared (VNIR) hyperspectral imager to acquire multiple HSI.
+Each HSI pixel included 462 wavelength bins ranging from $391$ to $1011$ nm.
+To evaluate the ability of the GSM to identify potential contaminant sources, rhodamine dye, a commonly used tracer in hydrological studies, was released into a pond.
+Two UAV flights spaced 15 minutes apart were used to capture the evolution of the plume as it dispersed into the surrounding water.
+
+The hyperspectral imager is in a pushbroom configuration so that HSI were captured one scan-line at a time.
+Data from an embedded GPS/INS unit enable direct georectification of captured imagery using the method outlined in \cite{muller2002program}.
+Additionally, an upward-facing Ocean Optics UV-Vis NIR spectrometer with a cosine corrector was included on the top of the UAV to measure incident solar irradiance.
+The configuration of the UAV together with a sample HSI are shown in Figure~\ref{fig:robotteam-data}.
+For a more detailed description of the system, the reader is directed to ref.
+\cite{robot-team-1, robot-team-2}.
+
+\begin{figure}[t]
+\includegraphics[width=\columnwidth]{methods/robot-team/robot-team-overview.pdf}
+\caption{Real HSI Dataset.
+\textbf{(a)} the UAV used to collect hyperspectral images.
+\textbf{(b)} The Resonon Pika XC2 hyperspectral imager used to acquire HSI.
+\textbf{(c)} A sample hyperspectral data cube.
+Spectra a plotted using at their geographic position with the log10-reflectance colored along the z axis and a pseudocolor image on top.
+The signature of the rhodamine dye plume is clearly identifiable in the water.
+\label{fig:robotteam-data}}
+\end{figure}
+
+Raw HSI were converted to reflectance using the downwelling irradiance spectrum captured simultaneously with each HSI.
+Given the UAV flies with the imager oriented to nadir, the reflectance is then given by
+\begin{equation}
+    \rho(\lambda) = \pi\,L(\lambda)/E_d(\lambda)
+\end{equation}
+where $L$ is the spectral radiance, $E_d$ is the downwelling irradiance, and a factor of $\pi$ is included resulting from the assumption of diffuse upwelling radiance \cite{ruddick2019review}.
+The UAV flights were performed near solar noon to maximize the amount of sunlight illuminating the water.
+For this Pond in North Texas, this corresponded to an average solar zenith angle of $56.7^{\circ}$ resulting in HSI with negligible sunglint effects.
+
+From the collected HSI, a water-only pixel mask was generated by identifying all pixel spectra with a normalized difference water index (NDWI) above $0.25$ as defined in ref.
+\cite{ndwi}.
+Of these water pixels, a combined dataset of $15,000$ spectra were sampled for GSM training.
+As a final processing step, reflectance spectra were limited to $\lambda \leq 900$ nm as wavelengths above this threshold showed significant noise.
+
+In order to justify values for the appropriate number of endmembers, the final dataset was decomposed using principal component analysis (PCA).
+A plot of the sorted explained variance for each PCA component is shown in Figure~\ref{fig:robot-team-pca}
 \begin{figure}[H]
-\includegraphics[width=\columnwidth]{methods/robot-team/robot-team-overview.pdf}
-\caption{Robot team dataset. \label{fig:robotteam-data}}
+\begin{center}
+\includegraphics[width=0.60\columnwidth]{results/robot-team/pca-variance.png}
+\end{center}
+\vspace{-20pt}
+\caption{Explained variance of PCA components for the real HSI dataset.
+A red horizontal line is superimposed on the plot marking an explained variance of $1\%$.
+All components past the fourth explain less than $1\%$ of the observed variance.}
+\label{fig:robot-team-pca}
 \end{figure}  
+The PCA decomposition of the data suggests that at least $3$ endmembers should be used for a mixing model and beyond $6$ there is little added benefit.
+Based on these observations, multiple GSM models were trained with $N_v$ ranging from $3$ to $6$, $\lambda_e$ ranging from $0.001$ to $1.0$, and $\lambda_w$ ranging from $1$ to $1000$ in order to explore the GSM parameter space.
+From these, a final model was identified using the BIC, AIC, and reconstruction RMSE.
+The resulting GSM was then explored to examine extracted endmembers and map the evolution of the rhodamaine plume by using the abundances given by the latent space representation of each pixel in the HSI.
 
@@
 Exploratory data analysis:
-\begin{figure}[H]
-\includegraphics[width=0.60\columnwidth]{results/robot-team/pca-variance.png}
-\caption{Explained variance of PCA components for robot team dataset. A red horizontal line is superimposed on the plot indicating an explained variance of $1\%$. \label{fig:robot-team-pca}}
-\end{figure}
 
@@
 GSM & Generative Simplex Mapping \\ 
-RBF & Radial Basis Function
+RBF & Radial Basis Function \\
+RMSE & Root Mean Square Error \\ 
+PCA & Principal Component Analysis
 \end{tabular}

### 2024-08-07 paper/paper.tex
pushing updates to results section

@@
 
+The results of GSM training on the synthetic linear mixing dataset described in Sec~\ref{sec:experiments} are illustrated in Figure~\ref{fig:usgs-fits}.
+Three versions of NMF were trained corresponding to Euclidean ($\ell_2$), KL-Divergence, and $\ell_{2,1}$ cost functions.
+GSM models using both a regular simplex grid and a grid of points sampled using a uniform Dirichlet distribution (referred to as a \textit{big} GSM model) were trained to compare performance on linear mixing tasks.
+
 \begin{figure}[H]
 \includegraphics[width=\columnwidth]{results/usgs/fit-comparison.pdf}
-\caption{Comparison of GSM against NMF on simulated USGS dataset. \label{fig:usgs-fits}}
+\caption{Comparison of GSM against NMF on simulated linear mixing dataset using USGS spectra.
+\textbf{(a)} The mean spectral angle computed between extracted endmembers and original endmembers.
+\textbf{(b)} the mean RMSE computed between extracted endmembers and original endmembers.
+\textbf{(c)} The mean abundance RMSE computed between original abundance data for each endmember and extracted abundances.
+\textbf{(d)} The reconstruction RMSE which evaluates the quality of fit.
+All models realized similar values reflecting convergence of the models to the level of random noise introduced into the data.}
+ \label{fig:usgs-fits}
 \end{figure}
 
+The quality of endmember extraction is measured by the mean spectral angle and mean endmember RMSE.
+As Figure~\ref{fig:usgs-fits} indicates, both versions of the GSM outperformed their NMF counterparts.
+Additionally, we note that for all GSM models, even including $\text{SNR}=0$, all model weights $W_{dm}$ for $m>N_v$ corresponding to nonlinear mixing were driven identically to $0.0$.
+This confirms that for datasets with purely linear mixing and random noise, the GSM correctly fits a mixing model without introducing unnecessary complexity.
+
+The quality of unmixing, i.e.
+abundance estimation, performed by each model was evaluated using the mean abundance RMSE.
+Here we again see that both versions of the GSM outperformed NMF.
+To justify that all models were fairly trained to convergence, the data reconstruction RMSE was also computed.
+This metric uses the trained mixing model to compute the error between the original dataset and the reconstructed spectra generated using the extracted endmembers and their abundances.
+For all GSM and NMF models, the data reconstruction RMSE converged to the level of random noise introduced into the data.
+These results reflect a fair comparison between NMF and GSM models.
+
+In Figure~\ref{fig:usgs-endmembers}, we plot the extracted endmembers for the GSM model trained on the synthetic dataset with $\text{SNR}=20$.
 \begin{figure}[H]
-\includegraphics[width=\columnwidth]{results/usgs/extracted-endmembers.png}
-\caption{Endmembers extracted using GSM for simulated USGS dataset with SNR$=20$. \label{fig:usgs-endmembers}}
+\begin{adjustwidth}{-\extralength}{-4.25cm}
+\begin{center}
+\includegraphics[width=1.1\columnwidth]{results/usgs/extracted-endmembers.png}
+\end{center}
+\end{adjustwidth}
+\caption{Endmembers extracted by the GSM for the simulated linear mixing dataset with SNR$=20$.
+Dashed lines correspond to orignal endmember spectra from the USGS spectral database.
+Solid lines superimposed on the plot indicate the extracted endmember spectra.
+Colored bands are included around each spectrum corresponding to the spectral variability estimated by the GSM precision parameter $\beta$ where the band width is $2\sqrt{\beta^{-1}}$ corresponding to $2$ standard deviations.}
+\label{fig:usgs-endmembers}
 \end{figure}  
+The extracted endmembers clearly fit the original source spectra while capturing local reflectance features.
+Additionally, the GSM precision parameter $\beta$ which is tuned during model training provides an assessment of spectral varaiibilty due to random noise.
+In Figure~\ref{fig:usgs-endmembers}
 

### 2024-08-07 paper/paper.tex
further updates to results section

@@
 The extracted endmembers clearly fit the original source spectra while capturing local reflectance features.
-Additionally, the GSM precision parameter $\beta$ which is tuned during model training provides an assessment of spectral varaiibilty due to random noise.
-In Figure~\ref{fig:usgs-endmembers}
+Additionally, the GSM precision parameter $\beta$ which is tuned during model training provides an assessment of spectral variability due to random noise.
+In Figure~\ref{fig:usgs-endmembers} this is indicated by the colored band centered around each extracted spectrum with a width of $2\sqrt{\beta^{-1}}$ corresponding to 2 standard deviations.
+The SNR of $20$ added to this example corresponds to zero-mean Gaussian noise with a standard deviation of $\sigma=0.0493$.
+After training the GSM found $\sqrt{\beta^{-1}}=0.0495$ which accurately captures the introduced noise.
+This ability to assess the spectral variability of extracted endmembers is a key advantage of the GSM.
 
@@
 
-Exploratory data analysis:
-
+For the dataset of real HSI spectra described in Section~\ref{sec:experiments}, 80 GSM models were trained to explore the GSM hyperparameter space.
+Model performance was compared using the BIC, AIC, and reconstruction RMSE with the results of the top $10$ performing models shown in Table~\ref{table:fit-comparison}.
 \begin{table}[H] 
-\caption{Hyperparameter optimization: Multiple models were trained to identify optimal hyperparameter values for the GSM applied to the water spectra dataset.
-Here we report the top $10$ models ranked according the the BIC.}
+\caption{GSM hyperparameter optimization for the real HSI dataset: Multiple GSM models were trained to explore the impact of model hyperparameters.
+$N_v$ values from $3$ to $6$ were explored as suggested by the PCA decomposition of the dataset.
+$\lambda_e$ was varied from $0.001$ to $1.0$ and $\lambda_w$ ranged from $1$ to $1000$.
+Here we report the top $10$ models ranked by increasing BIC.
+The AIC and reconstruction RMSE are also included for comparison.}
 \label{table:fit-comparison}
@@
 \end{table}
+As the table indicates, a nonlinear GSM with $N_v=3$, $\lambda_e=0.01$, and $\lambda_w=1.0$ achieved minimum BIC, AIC, and reconstruction RMSE values.
+The lower value of $\lambda_w$ identified from these models reflects the presence of nonlinear mixing effects in the HSI data.
+While the values of model weights $W_{dm}$ for $m>N_v$ corresponding to nonlinear mixing were $0$ for the synthetic dataset, here these weights obtained a small but non-negligible median value of $0.0012$.
+
+Reflectance spectra generated by the trained GSM corresponding to maximum abundance values for each endmember are plotted in Figure~\ref{fig:robot-team-endmembers}.
+Based on their signatures, these endmembers are identified with water, vegetation (including near shore, filamentous blue-green algae) and the rhodmaine dye plume.
 
@@
 \end{adjustwidth}
-\caption{Nonlinear GSM applied to water spectra: \textbf{(a)} Spectra corresponding to maximum abundances for each vertex in the trained GSM identified for as water, near-shore vegetation, and rhodamine dye sources.
-\textbf{(b)} The original hyperspectral datacube segmented according to the relative abundance of each endmember.
-Each water pixel defined by an NDWI $\geq 0.25$ is colored by smoothing interpolating between red, green, and blue colors corresponding to the relative abundance estimated for rhodamine, vegetation, and water spectra.}
+\caption{GSM applied to water spectra from real HSI dataset: \textbf{(a)} Spectra generated by the trained GSM for samples with maximum abundance for each endmember.
+Based on these spectral profiles, endmembers are identified with water, near-shore vegetation, and rhodamine dye.
+\textbf{(b)} A HSI segmented according to the relative abundance of each endmember.
+Each water pixel is colored by smoothing interpolating between red, green, and blue colors using the relative abundance estimated for rhodamine, vegetation, and water spectra.
+The rhodamine plume is clearly identifiable in the western portion of the HSI.}
 \label{fig:robot-team-endmembers}
 \end{figure}
+
+By assigning a unique color to each endmember, the spatial distribution of HSI spectra can be visualized by using estimated abundances to smoothly interpolate between endmember colors as shown in panel b of Figure~\ref{fig:robot-team-endmembers}.
+In other words, this method provides a fuzzy semantic segmentation of HSI pixels.
+Alternatively, each HSI pixel could be assigned a single class corresponding to the endmember with maximal abundance.
+In this way, the GSM can be used to visualize high-dimensional HSI spectra.
+
+To showcase the ability of the GSM to identify spatially localized contaminant sources, abundance maps for each individual endmember were generated by using the trained GTM to unmix all water pixels the HSI.
+In Figure~\ref{fig:endmember-abundance-dist} these abundance maps are compared for each endmember.
+In particular, the abundance map for the water endmember covers a majority of the center of the pond and decreases near the edge of the rhodamine plume where the dye and water mix.
+Vegetation in shallow water near the shore and the rhodamine dye plume in the western half of the pond are also clearly identified by their abundance maps.
+Given a HSI pixel resolution of $0.1 \times 0.1$ $\text{m}^2$, the spatial extent of vegetation is estimated to be $378.6$ $\text{m}^2$ while the plume initially extends across $255.7$ $\text{m}^2$.
 
@@
 \includegraphics[width=\columnwidth]{results/robot-team/endmember-abundances.pdf}
-\caption{Endmember distributions: \textbf{(a)} The distribution of abundance for the water class.
-This source dominates in the center of the water and decreases towards the edge of the pond where surface vegetation begins to dominate the reflectance signal.
-We note that the water abundance decreases near  the edge of the rhodamine plume reflecting dye mixing and diffusion.
-\textbf{(b)} The distribution of vegetation.
-This signal includes filamentous blue-green algae observed to accumulate in shallow waters near the shore.
+\caption{Endmember abundance distributions: \textbf{(a)} The spatial distribution of abundance for the water class.
+This source dominates in the center of the pond and decreases towards the shore where vegetation begins to dominate the reflectance signal.
+The water endmember abundance is also observed to decreases near the edge of the rhodamine plume reflecting dye mixing and diffusion.
+\textbf{(b)} The spatial distribution of vegetation.
+This endmember includes filamentous blue-green algae observed to accumulate in shallow waters near the shore.
 \textbf{(c)} The rhodamine dye plume extent segmented from the HSI.
-The total area for near-shore vegetation and rhodamine are estimated to be $378.6$ $\text{m}^2$ and $255.7$ $\text{m}^2$ respectively.}
+The total area for near-shore vegetation and rhodamine are estimated to be $378.6$ $\text{m}^2$ and $255.7$ $\text{m}^2$, respectively.}
+\label{fig:endmember-abundance-dist}
 \end{adjustwidth}
-\label{fig:endmember-abundance-dist}
 \end{figure}  
@@
 \end{figure}
+
+Applying the trained GSM to unmix HSI from UAV flights provides an efficient way to map contaminant dispersion.
+Figure~\ref{fig:plume-evo} demonstrates this by mapping the growing extent of the rhodamine dye between successive UAV flights.
+In just $15$ minutes, the plume expands from an initial area of $255.7$ $\text{m}^2$ to over $571.8$ $\text{m}^2$ as the dye disperses into the surrounding water.
 
@@
 This is the $\gamma$ factor that shows up in some of the mixing papers.
+    - Augment the data-space distribution to allow for more complicated noise models than a fixed $\beta$ parameter.
+A diagonal covariance matrix would allow the model to infer seperate uncertainty criterion for each wavelength bin which is relevant for real world HSI which may have different noise levels for different wavelength bins.
 

### 2024-08-07 paper/paper.tex
update results again...

@@
 \centering
-\includegraphics[width=1.25\columnwidth]{results/robot-team/plume-evo.pdf}
+\includegraphics[width=1.35\columnwidth]{results/robot-team/plume-evo.pdf}
 \end{adjustwidth}
@@
 
-Applying the trained GSM to unmix HSI from UAV flights provides an efficient way to map contaminant dispersion.
+Applying the trained GSM to unmix HSI from UAV flights provides an efficient way to map contaminant dispersion and transport.
 Figure~\ref{fig:plume-evo} demonstrates this by mapping the growing extent of the rhodamine dye between successive UAV flights.
-In just $15$ minutes, the plume expands from an initial area of $255.7$ $\text{m}^2$ to over $571.8$ $\text{m}^2$ as the dye disperses into the surrounding water.
+In just $15$ minutes, the plume expands from an initial area of $255.7$ $\text{m}^2$ to over $571.8$ $\text{m}^2$ as the dye mixes into the surrounding water.
 

### 2024-08-08 paper/paper.tex
added conclusions section

@@
 % Full title of the paper (Capitalized)
-\Title{Generative Simplex Mapping: Nonlinear Endmember Extraction for Source Apportionment and Spectral Unmixing for Hyperspectral Imagery}
-% \Title{Generative Simplex Mapping: Nonlinear Endmember Extraction and Spectral Unmixing for Hyperspectral Imagery}
+% \Title{Generative Simplex Mapping: Nonlinear Endmember Extraction for Source Apportionment and Spectral Unmixing for Hyperspectral Imagery}
+\Title{Generative Simplex Mapping: Nonlinear Endmember Extraction and Spectral Unmixing for Hyperspectral Imagery}
 
@@
 Water-based HSI specifically are prone to nonlinear mixing effects due to absorption features of dissolved and suspended substances, fluorescence of organic matter, and particulate scattering in turbid waters \cite{hsi-absorption, hsi-fluorescence, hsi-turibidity}.
-With the growing popularity of deep learning approaches in remote sensing, a variety of models based on autoencoder architectures have been introduced which enable nonlinear unmixing of HSI data \cite{non-negative-autoencoders,su2019daen,palsson2020convolutional,}.
+With the growing popularity of deep learning approaches in remote sensing, a variety of models based on autoencoder architectures have been introduced for unmixing HSI data \cite{non-negative-autoencoders,su2019daen,borsoi2019deep,palsson2020convolutional}.
 However, the complexity introduced by these models significantly impacts training time and decreases model interpretability.
@@
 
-- Comparison to NMF on linear mixing problem
-    - note that there are many versions of NMF with different constraints/regularization. The purpose here was to show that GSM is at least \textit{comparable} to NMF for linear unmixing problems
-    - Further note that both NMF and GSM do not rely on assumption of pure pixels (unlike VCA and other methods) 
-    - GSM also allows estimation of endmember variability via the precision parameter $\beta$ i.e. nodes in data space follow Gaussian distribution with $\sigma = \sqrt{\beta^{-1}}$
+The proliferation of hyperspectral imaging technologies in the remote sensing and environmental monitoring communities underscores the need for efficient algorithms which can make sense of high-dimensional HSI.
+Of particular interest are unsupervised algorithms which utilize all available wavelength bins to identify unique source profiles present realistic scenes.
+The GSM introduced in this paper is a novel method which simultaneously performs endmember extraction and spectral unmixing.
+Unlike popular endmember extraction algorithms such as VCA, PPI, and N-FINDR, the GSM does not assume the presence of pure pixels in HSI.
+Furthermore, the flexible structure of the mapping from the GSM latent space to the spectral data space allows the GSM to model nonlinear mixing effects, distinguishing it from other widely used linear models such as NMF.
+Being a probabilistic model also enables the GSM to quantify spectral variability via the precision parameter $\beta$ as indicated in Figure~\ref{fig:usgs-endmembers}.
+This additional capability is critical for analysis of realistic HSI where sensor noise, viewing geometry, and scene illumination all introduce uncertainty into extracted endmembers.
 
-- Nonlinear GSM for robot team 
-    - discuss identification of rhodamine plume and easy abundance mapping via the embedding coordinate
-    - discuss intrinsic dimensionality identification via abundance sparsity and BIC
-    - Discuss identification of non-linearity \textit{strength} via regularization parameter and BIC.
-    - reconstruction rmse can also be used to evaluate the relative performance for fixed model size (i.e. k and m) with varying regularization strength
+The growing popularity of deep learning methods has led many to explore applications of deep learning to the spectral unmixing problem.
+Variations on autoencoder architectures are a popular choice for both endmember extraction and unmixing.
+For example, Borsoi et al.
+utilized a variational autoencoder to more accurately identify spectral variability in a LMM \cite{borsoi2019deep}.
+Similarly, Palson et al.
+introduced an autoencoder with convolution layers to extend the LMM by incorporating spatial context from neighboring HSI pixels \cite{palsson2020convolutional}.
+However, the increased complexity of deep neural network approaches makes interpreting the latent space learned by encoder layers challenging and can dramatically increase training time and data volume requirements.
+In contrast, the latent space of the GSM is immediately interpretable due to the imposed simplex geometry while the form of $\psi$ leaves room to explore a variety of nonlinear mixing models.
+In this paper, $\psi$ was constructed to add nonlinear effects as additions or perturbations to an underlying linear mixing model with the $\lambda_w$ hyperparameter provided to control the degree of nonlinearity considered.
+In the original GTM formulation which inspired the GSM, the mapping $\psi$ strictly uses nonlinear features generated by Gaussian RBFs.
+Alternatively, one could easily construct $\psi$ for specific nonlinear models such as bilinear mixing by introducing features which depend on higher order combinations of the latent space (abundance) coordinates $z_k$.
+It is important to note, though, that the EM algorithm formulated for GSM training is made possible due to the linear combination of activations $\phi(z)$ with model weights $\mathbf{W}$.
+For mixing models which instead depend nonlinearly on weights $\mathbf{W}$, an EM algorithm may require iterative nonlinear solves during each M step.
 
-- Limitations of the GSM
-    - slow training due to computation of dinstance matrix between full dataset and embedding space
-    - An incremental version of the EM procedure as described in \cite{gtm-developments} can speed this up by considering batches $X_b$ of $X$ and performing only a partial E-step by updating responsibilities corresponding to the data batch
-    - However, M step still involves the "full" responsibility matrix. 
-    - This can be addressed by taking an ensembling approach as described in \cite{parallel-gtm} where multiple GTM are trained on subsets of $X$ and the results are averaged. 
-    - More generally, we may consider a mixture-of-GTMs as briefly desribed in \cite{gtm-orig} whereby the overall density is given as 
-    \begin{equation}
-        p(\mathbf{x}) = \sum_r P(r)p(\mathbf{x}\vert r)
-    \end{equation}
-    where there are $r$ individual GTM models with mixing coefficients $P(r)$ such that $\sum_r P(r) =1$
+The main limitation of the GSM is curse of dimensionality encountered when generating a grid on the $(N_v-1)$-simplex for large numbers of endmembers, $N_v$.
+This can be mitigated by instead randomly sampling points within the interior of the simplex using a uniform Dirichlet distribution to obtain a pre-determined number of nodes across the latent space.
+As the mixing coefficients $\pi_k$ are adapted during training, variability in inter-node spacing should not significantly impact model performance.
+Indeed, this was confirmed for the simulated dataset as summarized in Figure~\ref{fig:usgs-fits}.
+For considerably large datasets, the size of the responsibility matrix $\mathbf{R}$ can lead to extended training times.
+This can be addressed by augmenting the EM procedure to updated using mini-batches training samples during each EM step as described by Bishop et al.
+for the GTM in ref.
+\cite{gtm-developments}.
+Rather than updating the full responsibility matrix during each E step, a subset of $\mathbf{R}$ corresponding to a single batch of training data can be evaluated with all other entries held constant.
+The GSM can be extended in other ways, for example, by replacing the precision parameter $\beta$ with a covariance matrix to model wavelength dependent spectral variability common to many hyperspectral imaging platforms.
 
-- Other potential extensions to GSM model
-     - Adapt latent space  to allow for points whose barycentric coordinates sum to less than 1 in order to account for lighting effects (shouldn't matter for reflectance but would matter for radiance depending on incident light).
-This is the $\gamma$ factor that shows up in some of the mixing papers.
-    - Augment the data-space distribution to allow for more complicated noise models than a fixed $\beta$ parameter.
-A diagonal covariance matrix would allow the model to infer seperate uncertainty criterion for each wavelength bin which is relevant for real world HSI which may have different noise levels for different wavelength bins.
+Based on the results from applying the GSM to real HSI as illustrated in Figure~\ref{fig:endmember-abundance-dist} and Figure~\ref{fig:plume-evo}, one clear application of the GSM is to contaminant identification and water quality assessment.
+Modern UAV platforms make it possible to rapidly image bodies of water with limited access for in situ data collection.
+By equipping UAVs with hyperspectral imagers, the GSM can be used to identify abnormal spectral signatures corresponding to localized contaminant sources.
+Generating semantic segmentations of collected HSI as in panel b of Figure~\ref{fig:robot-team-endmembers} makes it easy to compare multiple HSI without resorting to pseudocolor images generated from a limited number of wavelength bands.
+Since the GSM models the distribution of all HSI spectra rather than individual parameters, it may also aide in situ data collection where sampling points can be chosen to maximize the area traversed in the GSM latent space rather than uniformly sampling across a wide spatial extent.
+Similar approaches have been developed to guide autonomous data collection vehicles by casting route planning as a prize collecting travelling salesman problem \cite{balas2007prize, suryan2020learning}.
 
-- Discuss other applications of the GSM
-    - Optimal route planning for in-situ data collection based on prize-collecting travelling salesman problem, i.e.
-construct a route which maximizies area traversed in GSM latent space (e.g.
-the n-simplex).
-    - Applications to source apportionment, e.g.
-linear GSM can be used for standard source apportionment problems (no reactions) and nonlinear GSM may be useful for modeling scenarios with additional complications due to reactions and during transport from source to receptor.
+As a final consideration, we note that the problem of endmember extraction and spectral unmixing for HSI is identical to source apportionment in the context of air quality.
+Here the goal is to identify measurement profiles associated with sources of ambient air pollution based on measurements at a receptor site.
+A popular model for source apportionment studies is Positive Matrix Factorization (PMF) introduced by Paatero and Tapper \cite{pmf-orig, ulbrich2009interpretation}.
+Just like NMF, PMF decomposes measurements into a linear combination of source profiles and relative abundances subject to non-negativity constraints.
+These linear receptor models assume that sources are not transformed during transport to the receptor, ignoring changes due to chemical reactions.
+Nonlinear mixing models like the GSM introduced here may therefore prove beneficial in this additional domain.
 
 \section{Conclusions}\label{sec:conclusions}
+
+In this paper, we introduce the GSM as a novel method for unsupervised endmember extraction and spectral unmixing.
+The model was compared to three varieties of NMF on a synthetic dataset of linearly mixed spectra from the USGS spectral database.
+Across all noise levels, the GSM outperformed NMF for both endmember accuracy and abundance estimation.
+When applied to real HSI collected over a north Texas pond, the GSM accurately models spectral mixing including nonlinear effects.
+Rhodamine dye was released into the pond the pond to simulate a localized contaminant source.
+Endmember abundances estimated by the GSM corresponding to the dye accurately tracked the evolution of the plume as it mixed into the surrounding water.
+These examples illustrate the power of combining nonlinear mixing models together with UAV-based hyperspectral imaging for environmental monitoring and water quality analysis.
+Future work will further develop the GSM to incorporate wavelength dependent spectral variability.
+Additionally the GSM may prove useful for air quality studies as a nonlinear source apportionment model.
 
@@
 RMSE & Root Mean Square Error \\ 
-PCA & Principal Component Analysis
+PCA & Principal Component Analysis \\ 
+PMF & Positive Matrix Factorization
 \end{tabular}

### 2024-08-09 paper/paper.tex
updating GSM paper

@@
 % Abstract (Do not insert blank lines, i.e. \\) 
-\abstract{\textcolor{red}{UPDATE ME!}}
+\abstract{\textcolor{violet}{In this paper, we introduce a new model for nonlinear endmember extraction and spectral unmixing of hyperspectral imagery called Generative Simplex Mapping (GSM).
+The model represents endmember mixing using a latent space with points sampled within a $(n-1)$-simplex corresponding to the abundance of $n$ unique sources.
+Points in this latent space are nonlinearly mapped to reflectance spectra via a flexible function combining linear and nonlinear mixing.
+Due to the probabilistic formulation of the GSM, spectral variability is also estimated by a precision parameter describing the distribution of observed spectra.
+Model parameters are determined using a generalized expectation-maximization algorithm.
+In the event of purely linear mixing, nonlinear contributions are naturally driven to zero.
+The GSM outperforms three varieties of non-negative matrix factorization for both endmember extraction accuracy and abundance estimation on a synthetic dataset of linearly mixed spectra from the USGS spectral library.
+In a second experiment, the GSM is applied to real hyperspectral imagery captured over a pond in North Texas.
+The model is able to accurately identify spectral signatures corresponding to near-shore algae, water, and rhodmaine tracer dye introduced into the pond to simulate water contamination by a localized source.
+Abundance maps generated using the GSM accurately track evolution of the dye plume as it mixes into the surrounding water.}}
 
@@
 The continued development of hyperspectral imaging technology has also led to a considerable reduction in size enabling their inclusion in the payloads of small unmanned aerial vehicles (UAV) \cite{adao2017hyperspectral, arroyo2019implementation}.
-Despite the proliferation of hyperspectral imaging, the considerable increase in data volume associated with hyperspectral images (HSI) poses significant challenges to real-time analysis at scale.
+Despite the proliferation of hyperspectral imaging data sources, the considerable increase in data volume associated with hyperspectral images (HSI) poses significant challenges to real-time analysis at scale.
 
@@
 For instance, Aurin et al.
-combined data from over 30 years of oceanographic field campaigns with paired satellite imagery to develop robust models for the inversion of key water quality indicators such as colored dissolved organic matter (CDOM) \cite{aurin2018remote}.
-This approach can be accelerated by combining UAV-based hyperspectral imaging with rapid in-situ data collection using autonomous boats \cite{robot-team-1, robot-team-2}.
+combined data from over 30 years of oceanographic field campaigns with paired satellite imagery to develop robust models for the inversion of key water quality indicators such as colored dissolved organic matter \cite{aurin2018remote}.
+This approach can be accelerated by combining UAV-based hyperspectral imaging with rapid, in-situ data collection using autonomous boats \cite{robot-team-1, robot-team-2}.
 Nevertheless, these supervised methods rely on \textit{a priori} knowledge of expected sources in order to identify appropriate reference sensors for data collection.
@@
 To address these shortcomings, Bishop et al introduced the Generative Topographic Mapping (GTM), a probabilistic latent-variable model inspired by the SOM \cite{gtm-orig}.
-When the latent space is chosen to be two dimensional, the GTM can be used to visualize the distribution of HSI spectra while the mapping of latent space nodes into the HSI data space provides endmembers\cite{robot-team-gtm}.
+When the latent space is chosen to be two dimensional, the GTM can be used to visualize the distribution of HSI spectra while the mapping of latent space nodes into the HSI data space provides endmembers \cite{robot-team-gtm}.
 Unfortunately, the rectangular latent space grid employed by the GTM does not directly translate into endmember abundances.
@@
 Eq~\ref{eq:complete-data-llh} is then maximized with respect to $\pi_k$, $\beta$, and $\mathbf{W}$ to obtain new parameter values.
-A detailed overview of the EM procedure can be found in Bishop \cite{bishop-prml}.
+For a detailed overview of the EM procedure, the reader is directed to ref.
+\cite{bishop-prml}.
 
@@
 We therefore take inspiration from the multiplicative updates for NMF introduced by Lee and Seung \cite{nmf-orig}.
-A standard gradient-based update for $\mathbf{W}$ would take the form
+A standard gradient-based update for $\mathbf{W}$ would normally take the form
 \begin{equation}
@@
 \end{equation}
-If we allow for individual learning rates $\eta_{dm}$ for each element of $\mathbf{W}$, then choosing 
+If we allow individual learning rates $\eta_{dm}$ for each element of $\mathbf{W}$, then choosing 
 \begin{equation}
@@
 
-Lastly, optimizing Eq~\ref{eq:complete-data-llh} with respect to $\beta$ yields
+Optimizing Eq~\ref{eq:complete-data-llh} with respect to $\beta$ yields the final update equation:
 \begin{equation}\label{eq:beta-update}
@@
 
-A freely available implementation of the GSM is accessible at \cite{gtm-code}.
+A freely available implementation of the GSM is provided at \cite{gtm-code}.
 The code is written in the Julia programming language and follows the Machine Learning in Julia (MLJ) common interface \cite{bezanson2012julia, blaom2020mlj}.
@@
 Geological Survey digital spectral library was generated by sampling $1000$ abundance vectors from a Dirichlet distribution with $\alpha_1=\alpha_2=\alpha_3=1/3$ \cite{usgs-spectra}.
-These known spectra and abundances are visualized in Figure~\ref{fig:usgs-data}.
+These source spectra and abundances are visualized in Figure~\ref{fig:usgs-data}.
 
@@
 \caption{Synthetic dataset fromed from USGS spectra.
-\textbf{(a)} spectra from the USGS spectral database used as the ground truth endmembers.
-These spectra were selected to match \cite{vca-orig}.
-\textbf{(b)} The distribution of abundances used in the dataset.
-Samples were generated using a Dirichlet distribution with $\alpha_1=\alpha_2=\alpha_3=1/3$.}
+\textbf{(a)} Spectra from the USGS spectral database used as the ground truth endmembers.
+These spectra were selected following the example in ref.
+\cite{vca-orig}.
+\textbf{(b)} The abundance distribution sampled for in the dataset.
+Samples were generated from a Dirichlet distribution with $\alpha_1=\alpha_2=\alpha_3=1/3$.}
 \label{fig:usgs-data}
@@
 Zero-mean Gaussian noise was added to the data to yield $9$ datasets with signal-to-noise ratios (SNR) ranging from $0$ to $\infty$ to examine the impact of random noise on GSM performance.
-For each of these datasets, a GSM was trained using $25$ nodes per edge with $\lambda_e$ set to the default value of $0.01$ and fixed to $\lambda_w = 100$.
+For each of these datasets, a GSM was trained using $25$ nodes per edge with $\lambda_e$ set to the default value of $0.01$ and $\lambda_w$ fixed to $100$.
 The large value of $\lambda_w$ was chosen to encourage the GSM to prefer models with limited nonlinear mixing.
@@
 For the purpose of evaluating the GSM, we considered the standard $\ell_2$ and KL-divergence formulations introduced by Lee and Seung \cite{nmf-algorithms}.
-We also consider the robust $\ell_{2,1}$ NMF as described by Kong et al \cite{nmf-l21}.
+We also included the robust $\ell_{2,1}$ NMF as described by Kong et al \cite{nmf-l21}.
 
@@
 
-To showcase the ability of the GSM to identify spatially localized contaminant sources, abundance maps for each individual endmember were generated by using the trained GTM to unmix all water pixels the HSI.
+To showcase the ability of the GSM to identify spatially localized contaminant sources, abundance maps for each individual endmember were generated by using the trained GTM to unmix all water pixels in the HSI.
 In Figure~\ref{fig:endmember-abundance-dist} these abundance maps are compared for each endmember.
-In particular, the abundance map for the water endmember covers a majority of the center of the pond and decreases near the edge of the rhodamine plume where the dye and water mix.
+The abundance map for the water endmember covers a majority of the center of the pond and decreases near the edge of the rhodamine plume where the dye and water mix.
 Vegetation in shallow water near the shore and the rhodamine dye plume in the western half of the pond are also clearly identified by their abundance maps.
@@
 In the original GTM formulation which inspired the GSM, the mapping $\psi$ strictly uses nonlinear features generated by Gaussian RBFs.
-Alternatively, one could easily construct $\psi$ for specific nonlinear models such as bilinear mixing by introducing features which depend on higher order combinations of the latent space (abundance) coordinates $z_k$.
+Alternatively, one could easily construct $\psi$ for specific nonlinear models such as bilinear mixing by introducing features which depend on higher order combinations of the latent space (abundance) coordinates.
 It is important to note, though, that the EM algorithm formulated for GSM training is made possible due to the linear combination of activations $\phi(z)$ with model weights $\mathbf{W}$.
@@
 
-The main limitation of the GSM is curse of dimensionality encountered when generating a grid on the $(N_v-1)$-simplex for large numbers of endmembers, $N_v$.
+The main limitation of the GSM is the curse-of-dimensionality encountered when generating a grid on the $(N_v-1)$-simplex for large numbers of endmembers, $N_v$.
 This can be mitigated by instead randomly sampling points within the interior of the simplex using a uniform Dirichlet distribution to obtain a pre-determined number of nodes across the latent space.
@@
 Indeed, this was confirmed for the simulated dataset as summarized in Figure~\ref{fig:usgs-fits}.
-For considerably large datasets, the size of the responsibility matrix $\mathbf{R}$ can lead to extended training times.
-This can be addressed by augmenting the EM procedure to updated using mini-batches training samples during each EM step as described by Bishop et al.
+For considerably large datasets, the size of the responsibility matrix $\mathbf{R}$ can also lead to extended training times.
+This can be addressed by augmenting the EM procedure to updated using mini-batches of training samples during each E-step as described by Bishop et al.
 for the GTM in ref.
 \cite{gtm-developments}.
-Rather than updating the full responsibility matrix during each E step, a subset of $\mathbf{R}$ corresponding to a single batch of training data can be evaluated with all other entries held constant.
-The GSM can be extended in other ways, for example, by replacing the precision parameter $\beta$ with a covariance matrix to model wavelength dependent spectral variability common to many hyperspectral imaging platforms.
+Rather than updating the full responsibility matrix, a subset of $\mathbf{R}$ corresponding to a single batch of training data can be evaluated with all other entries held constant.
+The GSM may also be extended in other ways, for example, by replacing the precision parameter $\beta$ with a covariance matrix to model wavelength dependent spectral variability common to many hyperspectral imaging platforms.
 
 Based on the results from applying the GSM to real HSI as illustrated in Figure~\ref{fig:endmember-abundance-dist} and Figure~\ref{fig:plume-evo}, one clear application of the GSM is to contaminant identification and water quality assessment.
-Modern UAV platforms make it possible to rapidly image bodies of water with limited access for in situ data collection.
+Modern UAV platforms make it possible to rapidly image bodies of water where direct access for in-situ data collection is restricted.
 By equipping UAVs with hyperspectral imagers, the GSM can be used to identify abnormal spectral signatures corresponding to localized contaminant sources.
 Generating semantic segmentations of collected HSI as in panel b of Figure~\ref{fig:robot-team-endmembers} makes it easy to compare multiple HSI without resorting to pseudocolor images generated from a limited number of wavelength bands.
-Since the GSM models the distribution of all HSI spectra rather than individual parameters, it may also aide in situ data collection where sampling points can be chosen to maximize the area traversed in the GSM latent space rather than uniformly sampling across a wide spatial extent.
-Similar approaches have been developed to guide autonomous data collection vehicles by casting route planning as a prize collecting travelling salesman problem \cite{balas2007prize, suryan2020learning}.
+Since the GSM models the distribution of all HSI spectra rather than individual parameters, it may also aide in-situ data collection by suggesting sampling points which maximize the area traversed in the GSM latent space rather than uniformly sampling across a wide spatial extent.
+Similar approaches have been developed to guide autonomous data collection vehicles by casting route planning as a prize collecting travelling salesman problem subject to resource constraints \cite{balas2007prize, suryan2020learning}.
 
@@
 
-In this paper, we introduce the GSM as a novel method for unsupervised endmember extraction and spectral unmixing.
+\textcolor{purple}{In this paper, we introduce the GSM as a novel method for unsupervised endmember extraction and spectral unmixing.
 The model was compared to three varieties of NMF on a synthetic dataset of linearly mixed spectra from the USGS spectral database.
@@
 When applied to real HSI collected over a north Texas pond, the GSM accurately models spectral mixing including nonlinear effects.
-Rhodamine dye was released into the pond the pond to simulate a localized contaminant source.
+Rhodamine dye was released into the pond to simulate a localized contaminant source.
 Endmember abundances estimated by the GSM corresponding to the dye accurately tracked the evolution of the plume as it mixed into the surrounding water.
@@
 Future work will further develop the GSM to incorporate wavelength dependent spectral variability.
-Additionally the GSM may prove useful for air quality studies as a nonlinear source apportionment model.
+Additionally the GSM may prove useful for air quality studies as a nonlinear source apportionment model.}
 
@@
 
-\funding{\textcolor{red}{Please add: ``This research received no external funding'' or ``This research was funded by NAME OF FUNDER grant number XXX.'' and  and ``The APC was funded by XXX''.
-Check carefully that the details given are accurate and use the standard spelling of funding agency names at \url{https://search.crossref.org/funding}, any errors may affect your future funding.}}
+\funding{\textcolor{red}{This research was funded by the following grants: the Texas National Security Network Excellence Fund award for Environmental Sensing Security Sentinels; the SOFWERX award for Machine Learning for Robotic Teams and NSF Award OAC-2115094; support from the University of Texas at Dallas Office of Sponsored Programs, Dean of Natural Sciences and Mathematics, and Chair of the Physics Department is gratefully acknowledged; TRECIS CC* Cyberteam (NSF \#2019135); NSF OAC-2115094 Award; and EPA P3 grant number 84057001-0.}}
 
@@
 
-\acknowledgments{\textcolor{red}{In this section you can acknowledge any support given which is not covered by the author contribution or funding sections.
-This may include administrative and technical support, or donations in kind (e.g., materials used for experiments).}}
+\acknowledgments{\textcolor{red}{David Shaefer is gratefully acknowledged for his help coordinating UAV flights during HSI collection.
+We thank Adam Aker, Lakitha Wijeratne, Shawhin Talebi, Ashen Fernando, Prabuddha Dewage, Mazhar Iqbal, Matthew Lary, and Gokul Balagopal for their help deploying the UAV.
+Don MacLaughlin, Scotty MacLaughlin, and the city of Plano, TX, are gratefully acknowledged for allowing us to deploy the autonomous robot team on their property.
+Christopher Simmons is gratefully acknowledged for his computational support.
+We thank  Antonio Mannino for his advice with regard to selecting the robotic boat's sensing suite.
+Annette Rogers is gratefully acknowledged for supporting the arrangement of insurance coverage.
+Steven Lyles is gratefully acknowledged for supporting the arrangement of a secure place for the robot team.}}
 

### 2024-08-09 paper/paper.tex
switching pdf to png for gsm diagram
fixing figure issue

@@
 % Full title of the paper (Capitalized)
-% \Title{Generative Simplex Mapping: Nonlinear Endmember Extraction for Source Apportionment and Spectral Unmixing for Hyperspectral Imagery}
-\Title{Generative Simplex Mapping: Nonlinear Endmember Extraction and Spectral Unmixing for Hyperspectral Imagery}
+% \Title{Generative Simplex Mapping: Non-linear Endmember Extraction for Source Apportionment and Spectral Unmixing for Hyperspectral Imagery}
+\Title{Generative Simplex Mapping: Non-linear Endmember Extraction and Spectral Unmixing for Hyperspectral Imagery}
 
 % MDPI internal command: Title for citation in the left column
-\TitleCitation{Generative Simplex Mapping: Nonlinear Endmember Extraction and Spectral Unmixing for Hyperspectral Imagery}
+\TitleCitation{Generative Simplex Mapping: Non-linear Endmember Extraction and Spectral Unmixing for Hyperspectral Imagery}
 
@@
 % Abstract (Do not insert blank lines, i.e. \\) 
-\abstract{\textcolor{violet}{In this paper, we introduce a new model for nonlinear endmember extraction and spectral unmixing of hyperspectral imagery called Generative Simplex Mapping (GSM).
+\abstract{\textcolor{violet}{In this paper, we introduce a new model for non-linear endmember extraction and spectral unmixing of hyperspectral imagery called Generative Simplex Mapping (GSM).
 The model represents endmember mixing using a latent space with points sampled within a $(n-1)$-simplex corresponding to the abundance of $n$ unique sources.
-Points in this latent space are nonlinearly mapped to reflectance spectra via a flexible function combining linear and nonlinear mixing.
+Points in this latent space are non-linearly mapped to reflectance spectra via a flexible function combining linear and non-linear mixing.
 Due to the probabilistic formulation of the GSM, spectral variability is also estimated by a precision parameter describing the distribution of observed spectra.
 Model parameters are determined using a generalized expectation-maximization algorithm.
-In the event of purely linear mixing, nonlinear contributions are naturally driven to zero.
-The GSM outperforms three varieties of non-negative matrix factorization for both endmember extraction accuracy and abundance estimation on a synthetic dataset of linearly mixed spectra from the USGS spectral library.
+In the event of purely linear mixing, non-linear contributions are naturally driven to zero.
+The GSM outperforms three varieties of non-negative matrix factorization for both endmember extraction accuracy and abundance estimation on a synthetic data set of linearly mixed spectra from the USGS spectral library.
 In a second experiment, the GSM is applied to real hyperspectral imagery captured over a pond in North Texas.
@@
 
-Hyperspectral imaging has emerged as a keystone technology in remote sensing where the ability to discern variations between high resolution spectra supports a plethora of critical applications such as environmental monitoring, biodiversity conservation, sustainable agriculture, and more.
+Hyperspectral imaging has emerged as a keystone technology in remote sensing, where the ability to discern variations between high-resolution spectra supports a plethora of critical applications such as environmental monitoring, biodiversity conservation, sustainable agriculture, and more.
 In recent years, many remote sensing platforms have been deployed with hyperspectral imaging payloads such as the Italian PRISMA mission launched in 2019, the German EnMAP launched in 2022, and recently NASA's PACE satellite launched in 2024 \cite{PRISMA-orig, EnMAP-orig, PACE-orig}.
-Many future missions also plan to incorporate hyperspectral imaging capabilities such as the European Space Agency's CHIME which will include over $200$ bands spanning visible, near infrared (NIR), and short-wave infrared (SWIR) wavelengths \cite{CHIME-orig}.
-The continued development of hyperspectral imaging technology has also led to a considerable reduction in size enabling their inclusion in the payloads of small unmanned aerial vehicles (UAV) \cite{adao2017hyperspectral, arroyo2019implementation}.
+Many future missions also plan to incorporate hyperspectral imaging capabilities such as the European Space Agency's CHIME which will include over $200$ bands spanning visible, near-infrared (NIR), and short-wave infrared (SWIR) wavelengths \cite{CHIME-orig}.
+The continued development of hyperspectral imaging technology has also led to a considerable reduction in size that enables its inclusion in the payloads of small unmanned aerial vehicles (UAVs) \cite{adao2017hyperspectral, arroyo2019implementation}.
 Despite the proliferation of hyperspectral imaging data sources, the considerable increase in data volume associated with hyperspectral images (HSI) poses significant challenges to real-time analysis at scale.
 
-Many approaches have been developed to make sense of HSI data.
-For example, spectral indices like the popular normalized difference vegetation index (NDVI) can be computed by taking ratios of spectral bands tailored to track specific reflectance characteristics \cite{thenkabail-indices,thenkabail2018hyperspectral}.
-These indices have the advantage of being easy to compute but suffer from significant inter-instrument variability while ignoring most of the information captured in HSI spectra \cite{ndvi-variability}.
-An alternative approach is to pair HSI data with in-situ measurements to enable supervised models which map spectra directly to parameters of interest.
+Many approaches have been developed to make sense of the HSI data.
+For example, spectral indices such as the popular normalized difference vegetation index (NDVI) can be computed by taking ratios of spectral bands tailored to track specific reflectance characteristics \cite{thenkabail-indices,thenkabail2018hyperspectral}.
+These indices have the advantage of being easy to compute, but suffer from significant variability between instruments while ignoring most of the information captured in HSI spectra \cite{ndvi-variability}.
+An alternative approach is to pair HSI data with in situ measurements to enable supervised models that map spectra directly to parameters of interest.
 However, this approach relies on serendipitous satellite overpasses above sensing sites to generate sufficient quantities of aligned data for model training and evaluation.
-For instance, Aurin et al.
+For example, Aurin et al.
 combined data from over 30 years of oceanographic field campaigns with paired satellite imagery to develop robust models for the inversion of key water quality indicators such as colored dissolved organic matter \cite{aurin2018remote}.
-This approach can be accelerated by combining UAV-based hyperspectral imaging with rapid, in-situ data collection using autonomous boats \cite{robot-team-1, robot-team-2}.
-Nevertheless, these supervised methods rely on \textit{a priori} knowledge of expected sources in order to identify appropriate reference sensors for data collection.
-In light of these limitations, unsupervised methods are needed which can expediently extract source signatures from high dimensional HSI.
+This approach can be accelerated by combining UAV-based hyperspectral imaging with rapid in situ data collection using autonomous boats \cite{robot-team-1, robot-team-2}.
+However, these supervised methods rely on \textit{a priori} knowledge of expected sources to identify appropriate reference sensors for data collection.
+In light of these limitations, unsupervised methods are needed that can expediently extract source signatures from high-dimensional HSI.
 
@@
 The task of unsupervised source identification using HSI data therefore involves two steps: endmember extraction, and abundance estimation.
-Techniques such as vertex component analysis (VCA), the pixel purity index (PPI), and N-FINDR solve this first task by identifying endmember spectra from HSI assuming the presence of some pure (unmixed) pixels \cite{vca-orig, ppi-orig, N-FINDR-orig}.
-By further assuming a linear mixing model (LMM) whereby observed spectra are described by a linear combination of endmembers with non-negative abundances, HSI can then be unmixed using a variety of techniques such as constrained least-squares \cite{spectral-unmixing-orig, fcls-unmixing}.
-Among these methods Non-negative Matrix Factorization (NMF) is a widely used approach which extracts endmember spectra and unmixes abundances simultaneously via matrix factorization \cite{nmf-orig, unmixing-nmf-review, unmixing-nmf-review-2}.
-The update equations for NMF can be formulated into multiplicative updates which guarantee the non-negativity of endmember spectra and their associated abundances \cite{nmf-algorithms}.
-For this reason the continued development of new NMF varieties remains an active area of research.
+Techniques such as vertex component analysis (VCA), the pixel purity index (PPI), and N-FINDR solve this first task by identifying HSI endmember spectra assuming the presence of some pure (unmixed) pixels \cite{vca-orig, ppi-orig, N-FINDR-orig}.
+By further assuming a linear mixing model (LMM) in which observed spectra are described by a linear combination of endmembers with non-negative abundances, HSI can be unmixed using a variety of techniques such as constrained least squares \cite{spectral-unmixing-orig, fcls-unmixing}.
+Among these methods, Nonnegative Matrix Factorization (NMF) is a widely used approach which extracts endmember spectra and unmixes abundances simultaneously via matrix factorization \cite{nmf-orig, unmixing-nmf-review, unmixing-nmf-review-2}.
+The update equations for NMF can be formulated as multiplicative updates which guarantee the non-negativity of endmember spectra and their associated abundances \cite{nmf-algorithms}.
+For this reason, the continued development of new NMF varieties remains an active area of research.
 
-In realistic scenes multiple scattering and surface variability can easily challenge the assumption of linear mixing \cite{heylen2014review}.
-Water-based HSI specifically are prone to nonlinear mixing effects due to absorption features of dissolved and suspended substances, fluorescence of organic matter, and particulate scattering in turbid waters \cite{hsi-absorption, hsi-fluorescence, hsi-turibidity}.
+In realistic scenes, multiple scattering and surface variability can easily challenge the assumption of linear mixing \cite{heylen2014review}.
+Water-based HSI specifically are prone to non-linear mixing effects due to absorption features of dissolved and suspended substances, fluorescence of organic matter, and particulate scattering in turbid waters \cite{hsi-absorption, hsi-fluorescence, hsi-turibidity}.
 With the growing popularity of deep learning approaches in remote sensing, a variety of models based on autoencoder architectures have been introduced for unmixing HSI data \cite{non-negative-autoencoders,su2019daen,borsoi2019deep,palsson2020convolutional}.
-However, the complexity introduced by these models significantly impacts training time and decreases model interpretability.
-An ideal approach should enable both endmember extraction and nonlinear unmixing while accounting for spectral variability.
+However, the complexity introduced by these models significantly impacts training time and decreases the interpretability of the model.
+An ideal approach should enable both endmember extraction and non-linear unmixing while accounting for spectral variability.
 
-The self-organizing map (SOM) is an unsupervised machine learning method which maps high-dimensional data onto a low-dimensional grid while preserving the topological relationships between data points \cite{kohonen-som-1}.
-This low dimensional representation provides a convenient way to visualize HSI data while the weight vectors for each SOM node can be interpreted as representative spectra \cite{cantero2004analysis, duran2007time,som-hsi}.
-If labelled reference spectra are available, the SOM can be used to enable semi-supervised labelling of HSI pixels \cite{riese2019supervised}.
-The SOM has also been shown to be effective for compression of HSIs acquired by a CubeSat \cite{som-satellite}.
+The self-organizing map (SOM) is an unsupervised machine learning method that maps high-dimensional data to a low-dimensional grid while preserving the topological relationships between data points \cite{kohonen-som-1}.
+This low-dimensional representation provides a convenient way to visualize HSI data while the weight vectors for each SOM node can be interpreted as representative spectra \cite{cantero2004analysis, duran2007time,som-hsi}.
+If labelled reference spectra are available, the SOM can be used to enable semi-supervised labeling of HSI pixels \cite{riese2019supervised}.
+The SOM has also been shown to be effective for the compression of HSIs acquired by a CubeSat \cite{som-satellite}.
 Despite these capabilities, the SOM does not offer a probabilistic interpretation and relies on a heuristic training procedure with hyperparameters that can be challenging to tune.
-To address these shortcomings, Bishop et al introduced the Generative Topographic Mapping (GTM), a probabilistic latent-variable model inspired by the SOM \cite{gtm-orig}.
-When the latent space is chosen to be two dimensional, the GTM can be used to visualize the distribution of HSI spectra while the mapping of latent space nodes into the HSI data space provides endmembers \cite{robot-team-gtm}.
+To address these shortcomings, Bishop et al.
+introduced the Generative Topographic Mapping (GTM), a probabilistic latent-variable model inspired by the SOM \cite{gtm-orig}.
+When choosing a two-dimensional latent space, the GTM can be used to visualize the distribution of HSI spectra while mapping the latent space nodes to the HSI data space provides endmembers \cite{robot-team-gtm}.
 Unfortunately, the rectangular latent space grid employed by the GTM does not directly translate into endmember abundances.
@@
 
-In this paper, we introduce a new variant of the GTM dubbed Generative Simplex Mapping (GSM) which can extract endmember spectra and unmix nonlinear mixtures.
+In this paper, we introduce a new variant of the GTM dubbed Generative Simplex Mapping (GSM), which can extract endmember spectra and unmix non-linear mixtures.
 By replacing the rectangular latent space of the GTM with a gridded $n$-simplex, the vertices of the GSM can be immediately interpreted as endmembers corresponding to $n+1$ unique spectral signatures.
 The mapping from the latent space to the HSI data space models signal mixing while barycentric coordinates for the latent space simplex estimate relative endmember abundances.
-Furthermore, by taking inspiration from the multiplicative updates of NMF, the GSM algorithm maintains non-negativity of resulting endmember spectra.
-If only linear mixing is present, the GSM algorithm drives nonlinear contributions to $0$.
-Prior distributions included for GSM model weights yield hyperparameters which can be tuned to control the smoothness of resulting spectra and the degree of nonlinear mixing applied.
+Furthermore, by taking inspiration from the multiplicative updates of NMF, the GSM algorithm maintains the non-negativity of resulting endmember spectra.
+If only linear mixing is present, the GSM algorithm drives non-linear contributions to $0$.
+Prior distributions included for GSM model weights yield hyperparameters which can be tuned to control the smoothness of the resulting spectra and the degree of non-linear mixing applied.
 
-The rest of the paper is structured as follows.
+The remainder of the paper is structured as follows.
 Section~\ref{sec:gsm} describes the proposed method.
-Section~\ref{sec:experiments} and Section~\ref{sec:results} describe experiments using simulated and real HSI data to evaluate the GSM model.
+Section~\ref{sec:experiments} and Section~\ref{sec:results} describe experiments using real and simulated HSI data to evaluate the GSM model.
 Section~\ref{sec:disucssion} discusses additional applications and extensions of the GSM to be explored in future work.
@@
 
-In the original GTM formulation, data vectors $x$ (reflectance spectra) are described by latent variables $z$ mapped into the data space by a nonlinear function $\psi$.
-The data space distribution is taken to be normal with precision parameter $\beta$ to account for measurement noise and spectral variability.
+In the original GTM formulation, the data vectors $x$ (reflectance spectra) are described by latent variables $z$ mapped into the data space by a non-linear function $\psi$.
+The data space distribution is taken to be normal with the precision parameter $\beta$ to account for measurement noise and spectral variability.
 The GSM uses same structure, that is
@@
 To adapt the GTM to describe endmember mixing, the GSM makes two key changes.
-The first is to replace the rectangular GTM grid with a gridded simplex having $N_v$ vertices.
-Barycentric coordinates for each node then describe the relative abundance of each endmember with each vertex corresponding to a pure endmember.
-The second change is to replace the equal prior probabilities with adaptive mixing coefficients $\pi_k$ to allow the GSM to model HSI with non-uniform mixing distributions.
+The first is to replace the rectangular GTM grid with a gridded simplex with $N_v$ vertices.
+The barycentric coordinates for each node then describe the relative abundance of each endmember with each vertex corresponding to a pure endmember.
+The second change is to replace equal prior probabilities with adaptive mixing coefficients $\pi_k$ to allow the GSM to model HSI with nonuniform mixing distributions.
 Together, this leads to a latent space prior distribution given by
@@
 
-For a dataset containing $N$-many records, these definitions yield a log likelihood function given by
+For a data set containing $N$-many records, these definitions yield a log likelihood function given by
 \begin{equation}\label{eqn:llh}
@@
 In the standard LMM model, the mapping $\psi$ is given by $\psi(z;W) = \mathbf{W}z$ where the columns of $\mathbf{W}$ correspond to endmember spectra.
-To model nonlinear mixing, $z$ is replaced by the output of $M$-many activation functions such that $\psi(z;W) = \mathbf{W}\phi(z)$.
+To model non-linear mixing, $z$ is replaced by the output of $M$-many activation functions such that $\psi(z;W) = \mathbf{W}\phi(z)$.
 The activations applied to each GSM node can then be collected to form a matrix with elements $\Phi_{km} = \phi_m(z_k)$.
@@
 where $s$ is the spacing between RBF centers.
-In this form, the first $N_v$ columns of $\mathbf{W}$ correspond to endmember spectra while the remaining columns account for additional nonlinear effects.
+In this form, the first $N_v$ columns of $\mathbf{W}$ correspond to endmember spectra, while the remaining columns account for additional non-linear effects.
 For linear mixing, the GSM training algorithm should therefore drive $W_{dm}$ to $0$ for $m\geq N_v$.
-A visualization of the GSM model is shown in Fig~\ref{fig:gsm-diagram}.
+A visualization of the GSM model is shown in Figure~\ref{fig:gsm-diagram}.
 
 \begin{figure}[H]
-\includegraphics[width=\columnwidth]{methods/gsm/gsm-diagram.pdf}
+\includegraphics[width=\columnwidth]{methods/gsm/gsm-diagram.png}
 \caption{Illustration of the GSM.
@@
 For $m>N_v$ we use a zero-mean Laplace distribution, $W_{dm}\sim\dfrac{\lambda_w}{2}\exp(-\lambda_w\lvert W_{dm}\rvert)$, with scale parameter $\lambda_w^{-1}$.
-Under these choices $\lambda_e$ corresponds to $L_2$ regularization on endmember spectra while $\lambda_w$ corresponds to $L_1$ regularization on the nonlinear activations.
-In other words, $\lambda_e$ governs the smoothness of the resulting endmembers while $\lambda_w$ encourages sparsity for the nonlinear contributions.
+Under these choices $\lambda_e$ corresponds to $L_2$ regularization on endmember spectra while $\lambda_w$ corresponds to $L_1$ regularization on the non-linear activations.
+In other words, $\lambda_e$ governs the smoothness of the resulting endmembers while $\lambda_w$ encourages sparsity for the non-linear contributions.
 
@@
 Suppose that we have current estimates for the model weights $\mathbf{W}$, mixing coefficients $\pi_k$, and precision parameter $\beta$.
-During the expectation step we compute the posterior probabilities, that is, the responsibility of each GSM node for each spectrum in the dataset:
+During the expectation step we compute the posterior probabilities, that is, the responsibility of each GSM node for each spectrum in the data set:
 \begin{equation}\label{eq:responsibility}
@@
 \end{equation}
-where $X_{nd}$ is the $d$-th component of the $n$-th spectrum in the dataset.
+where $X_{nd}$ is the $d$-th component of the $n$-th spectrum in the data set.
 Eq~\ref{eq:complete-data-llh} is then maximized with respect to $\pi_k$, $\beta$, and $\mathbf{W}$ to obtain new parameter values.
@@
 Optimization of Eq~\ref{eq:complete-data-llh} with respect to $\mathbf{W}$ leads to a linear system which can be solved using standard numerical methods.
-However, in this form we can not guarantee the non-negativity of $\mathbf{W}$ required to describe reflectance spectra.
-We therefore take inspiration from the multiplicative updates for NMF introduced by Lee and Seung \cite{nmf-orig}.
+However, in this form, we cannot guarantee the non-negativity of $\mathbf{W}$ required to describe reflectance spectra.
+Therefore, we take inspiration from the multiplicative updates for NMF introduced by Lee and Seung \cite{nmf-orig}.
 A standard gradient-based update for $\mathbf{W}$ would normally take the form
@@
 Importantly, applying $\psi$ to each vertex extracts endmember spectra from the GSM.
-Slices $R_{\left[:,n\right]}$ of the matrix $\mathbf{R}$ define the responsibility of each latent node $z_k$ for the $n$-th spectrum $x_n$ in the dataset.
+Slices $R_{\left[:,n\right]}$ of the matrix $\mathbf{R}$ define the responsibility of each latent node $z_k$ for the $n$-th spectrum $x_n$ in the data set.
 Therefore once the GSM has been trained, $\mathbf{R}$ can be used to unmix endmember abundances by representing each record in the latent space via the mean:
@@
 
-To illustrate the effectiveness of the GSM, we first demonstrate its ability to model simple linear mixing by driving nonlinear weights to zero during the fitting process.
-To this end, a synthetic dataset comprising linear mixtures of three sample spectra from the U.S.
+To illustrate the effectiveness of the GSM, we first demonstrate its ability to model simple linear mixing by driving non-linear weights to zero during the fitting process.
+To this end, a synthetic data set comprising linear mixtures of three sample spectra from the U.S.
 Geological Survey digital spectral library was generated by sampling $1000$ abundance vectors from a Dirichlet distribution with $\alpha_1=\alpha_2=\alpha_3=1/3$ \cite{usgs-spectra}.
@@
 \begin{figure}[H]
-\includegraphics[width=\columnwidth]{methods/usgs/usgs-dataset.pdf}
-\caption{Synthetic dataset fromed from USGS spectra.
+\includegraphics[width=\columnwidth]{paper/figures/methods/usgs/usgs-dataset.pdf}
+\caption{Synthetic data set formed from USGS spectra.
 \textbf{(a)} Spectra from the USGS spectral database used as the ground truth endmembers.
@@
 \cite{vca-orig}.
-\textbf{(b)} The abundance distribution sampled for in the dataset.
+\textbf{(b)} The abundance distribution sampled for in the data set.
 Samples were generated from a Dirichlet distribution with $\alpha_1=\alpha_2=\alpha_3=1/3$.}
@@
 
-Zero-mean Gaussian noise was added to the data to yield $9$ datasets with signal-to-noise ratios (SNR) ranging from $0$ to $\infty$ to examine the impact of random noise on GSM performance.
-For each of these datasets, a GSM was trained using $25$ nodes per edge with $\lambda_e$ set to the default value of $0.01$ and $\lambda_w$ fixed to $100$.
-The large value of $\lambda_w$ was chosen to encourage the GSM to prefer models with limited nonlinear mixing.
+Zero-mean Gaussian noise was added to the data to yield $9$ data sets with signal-to-noise ratios (SNR) ranging from $0$ to $\infty$ to examine the impact of random noise on GSM performance.
+For each of these data sets, a GSM was trained using $25$ nodes per edge with $\lambda_e$ set to the default value of $0.01$ and $\lambda_w$ fixed to $100$.
+The large value of $\lambda_w$ was chosen to encourage the GSM to prefer models with limited non-linear mixing.
 
-For comparison, NMF models were also trained on each dataset as this method is both highly popular and does not include the pure-pixel assumption common to other technqieus like VCA and PPI.
+For comparison, NMF models were also trained on each data set as this method is both highly popular and does not include the pure-pixel assumption common to other technqieus like VCA and PPI.
 Countless variations on the original NMF method have been introduced into the literature with one review identifying more than $100$ distinct NMF variations \cite{unmixing-nmf-review}.
@@
 Abundance estimation was similarly evaluated using the Mean RMSE between true abundance and estimated abundance values for each endmember.
-Finally, the reconstruction RMSE, that is, the RMSE computed between the original dataset and the dataset reconstructed via the extracted endmembers and their associated abundances was computed.
+Finally, the reconstruction RMSE, that is, the RMSE computed between the original data set and the data set reconstructed via the extracted endmembers and their associated abundances was computed.
 This provides a model-agnostic criterion to guarantee that each model sufficiently converged during training.
 
-\subsection{Nonlinear Mixing: Water Contaminant Identification}
+\subsection{Non-linear Mixing: Water Contaminant Identification}
 
-To assess the ability of the GSM to unmix realistic scenes likely to involve nonlinear mixing effects, we consider a dataset of real HSI collected in Montague, North Texas on 9 December 2020.
-A Freefly Alta-X autonomous quadcopter was used as the UAV platform and was equipped with a Resonon Pika XC2 visible+near-infrared (VNIR) hyperspectral imager to acquire multiple HSI.
+To assess the ability of the GSM to unmix realistic scenes likely to involve non-linear mixing effects, we consider a data set of real HSI collected in Montague, North Texas on 9 December 2020.
+A Freefly Alta-X autonomous quadcopter was used as a UAV platform and equipped with a Resonon Pika XC2 visible+near-infrared (VNIR) hyperspectral imager to acquire multiple HSI.
 Each HSI pixel included 462 wavelength bins ranging from $391$ to $1011$ nm.
-To evaluate the ability of the GSM to identify potential contaminant sources, rhodamine dye, a commonly used tracer in hydrological studies, was released into a pond.
+To evaluate the ability of GSM to identify potential contaminant sources, rhodamine dye, a commonly used tracer in hydrological studies, was released into a pond.
 Two UAV flights spaced 15 minutes apart were used to capture the evolution of the plume as it dispersed into the surrounding water.
 
-The hyperspectral imager is in a pushbroom configuration so that HSI were captured one scan-line at a time.
-Data from an embedded GPS/INS unit enable direct georectification of captured imagery using the method outlined in \cite{muller2002program}.
+The hyperspectral imager is in a pushbroom configuration so that HSI was captured one scan line at a time.
+Data from an embedded GPS/INS unit enable direct georectification of captured imagery using the method described in \cite{muller2002program}.
 Additionally, an upward-facing Ocean Optics UV-Vis NIR spectrometer with a cosine corrector was included on the top of the UAV to measure incident solar irradiance.
@@
 \includegraphics[width=\columnwidth]{methods/robot-team/robot-team-overview.pdf}
-\caption{Real HSI Dataset.
+\caption{Real HSI Data set.
 \textbf{(a)} the UAV used to collect hyperspectral images.
@@
 
-From the collected HSI, a water-only pixel mask was generated by identifying all pixel spectra with a normalized difference water index (NDWI) above $0.25$ as defined in ref.
+From the collected HSI, a water-only pixel mask was generated by identifying all pixel spectra with a normalized difference water index (NDWI) greater than $0.25$ as defined in ref.
 \cite{ndwi}.
-Of these water pixels, a combined dataset of $15,000$ spectra were sampled for GSM training.
+Of these water pixels, a combined data set of $15,000$ spectra was sampled for GSM training.
 As a final processing step, reflectance spectra were limited to $\lambda \leq 900$ nm as wavelengths above this threshold showed significant noise.
 
-In order to justify values for the appropriate number of endmembers, the final dataset was decomposed using principal component analysis (PCA).
-A plot of the sorted explained variance for each PCA component is shown in Figure~\ref{fig:robot-team-pca}
+In order to justify values for the appropriate number of endmembers, the final data set was decomposed using principal component analysis (PCA).
+A plot of the explained variance organized for each PCA component is shown in Figure~\ref{fig:robot-team-pca}
 \begin{figure}[H]
@@
 \vspace{-20pt}
-\caption{Explained variance of PCA components for the real HSI dataset.
-A red horizontal line is superimposed on the plot marking an explained variance of $1\%$.
+\caption{Explained variance of PCA components for the real HSI data set.
+A red horizontal line is superimposed on the graph, marking an explained variance of $1\%$.
 All components past the fourth explain less than $1\%$ of the observed variance.}
@@
 
-The results of GSM training on the synthetic linear mixing dataset described in Sec~\ref{sec:experiments} are illustrated in Figure~\ref{fig:usgs-fits}.
+The results of GSM training on the synthetic linear mixing data set described in Sec~\ref{sec:experiments} are illustrated in Figure~\ref{fig:usgs-fits}.
 Three versions of NMF were trained corresponding to Euclidean ($\ell_2$), KL-Divergence, and $\ell_{2,1}$ cost functions.
-GSM models using both a regular simplex grid and a grid of points sampled using a uniform Dirichlet distribution (referred to as a \textit{big} GSM model) were trained to compare performance on linear mixing tasks.
+GSM models using both a regular simplex grid and a grid of points sampled using a uniform Dirichlet distribution (referred to as a \textit{big} GSM model) were trained to compare performance in linear mixing tasks.
 
@@
 \includegraphics[width=\columnwidth]{results/usgs/fit-comparison.pdf}
-\caption{Comparison of GSM against NMF on simulated linear mixing dataset using USGS spectra.
+\caption{Comparison of GSM against NMF on simulated linear mixing data set using USGS spectra.
 \textbf{(a)} The mean spectral angle computed between extracted endmembers and original endmembers.
@@
 
-The quality of endmember extraction is measured by the mean spectral angle and mean endmember RMSE.
+The quality of endmember extraction is measured by the mean spectral angle and the mean endmember RMSE.
 As Figure~\ref{fig:usgs-fits} indicates, both versions of the GSM outperformed their NMF counterparts.
-Additionally, we note that for all GSM models, even including $\text{SNR}=0$, all model weights $W_{dm}$ for $m>N_v$ corresponding to nonlinear mixing were driven identically to $0.0$.
-This confirms that for datasets with purely linear mixing and random noise, the GSM correctly fits a mixing model without introducing unnecessary complexity.
+Additionally, we note that for all GSM models, even including $\text{SNR}=0$, all model weights $W_{dm}$ for $m>N_v$ corresponding to non-linear mixing were driven identically to $0.0$.
+This confirms that, for data sets with purely linear mixing and random noise, the GSM correctly fits a mixing model without introducing unnecessary complexity.
 
-The quality of unmixing, i.e.
-abundance estimation, performed by each model was evaluated using the mean abundance RMSE.
+The quality of the unmixing, that is, the estimation of the abundance, performed by each model was evaluated using the mean abundance RMSE.
 Here we again see that both versions of the GSM outperformed NMF.
-To justify that all models were fairly trained to convergence, the data reconstruction RMSE was also computed.
-This metric uses the trained mixing model to compute the error between the original dataset and the reconstructed spectra generated using the extracted endmembers and their abundances.
-For all GSM and NMF models, the data reconstruction RMSE converged to the level of random noise introduced into the data.
-These results reflect a fair comparison between NMF and GSM models.
+To justify that all models were fairly trained to convergence, the RMSE data reconstruction was also computed.
+This metric uses the trained mixing model to compute the error between the original data set and the reconstructed spectra generated using the extracted endmembers and their abundances.
+For all GSM and NMF models, the RMSE data reconstruction converged to the level of random noise introduced into the data.
+These results reflect a fair comparison between the NMF and GSM models.
 
-In Figure~\ref{fig:usgs-endmembers}, we plot the extracted endmembers for the GSM model trained on the synthetic dataset with $\text{SNR}=20$.
+In Figure~\ref{fig:usgs-endmembers}, we plot the endmembers extracted for the GSM model trained on the synthetic data set with $\text{SNR}=20$.
 \begin{figure}[H]
@@
 \end{adjustwidth}
-\caption{Endmembers extracted by the GSM for the simulated linear mixing dataset with SNR$=20$.
-Dashed lines correspond to orignal endmember spectra from the USGS spectral database.
+\caption{Endmembers extracted by the GSM for the simulated linear mixing data set with SNR$=20$.
+The dashed lines correspond to original endmember spectra from the USGS spectral database.
 Solid lines superimposed on the plot indicate the extracted endmember spectra.
@@
 The extracted endmembers clearly fit the original source spectra while capturing local reflectance features.
-Additionally, the GSM precision parameter $\beta$ which is tuned during model training provides an assessment of spectral variability due to random noise.
+Furthermore, the GSM precision parameter $\beta$ which is tuned during model training, provides an assessment of spectral variability due to random noise.
 In Figure~\ref{fig:usgs-endmembers} this is indicated by the colored band centered around each extracted spectrum with a width of $2\sqrt{\beta^{-1}}$ corresponding to 2 standard deviations.
 The SNR of $20$ added to this example corresponds to zero-mean Gaussian noise with a standard deviation of $\sigma=0.0493$.
-After training the GSM found $\sqrt{\beta^{-1}}=0.0495$ which accurately captures the introduced noise.
+After training, the GSM found $\sqrt{\beta^{-1}}=0.0495$ that accurately captures the introduced noise.
 This ability to assess the spectral variability of extracted endmembers is a key advantage of the GSM.
 
-\subsection{Nonlinear Mixing: Rhodamine Dye Plume}
+\subsection{Non-linear Mixing: Rhodamine Dye Plume}
 
-For the dataset of real HSI spectra described in Section~\ref{sec:experiments}, 80 GSM models were trained to explore the GSM hyperparameter space.
-Model performance was compared using the BIC, AIC, and reconstruction RMSE with the results of the top $10$ performing models shown in Table~\ref{table:fit-comparison}.
+For the data set of real HSI spectra described in Section~\ref{sec:experiments}, 80 GSM models were trained to explore the GSM hyperparameter space.
+The performance of the model was compared using the BIC, AIC and RMSE reconstruction with the results of the top $10$ performing models shown in Table~\ref{table:fit-comparison}.
 \begin{table}[H] 
-\caption{GSM hyperparameter optimization for the real HSI dataset: Multiple GSM models were trained to explore the impact of model hyperparameters.
-$N_v$ values from $3$ to $6$ were explored as suggested by the PCA decomposition of the dataset.
+\caption{GSM hyperparameter optimization for the real HSI data set: Multiple GSM models were trained to explore the impact of model hyperparameters.
+$N_v$ values from $3$ to $6$ were explored as suggested by the PCA decomposition of the data set.
 $\lambda_e$ was varied from $0.001$ to $1.0$ and $\lambda_w$ ranged from $1$ to $1000$.
@@
 \end{table}
-As the table indicates, a nonlinear GSM with $N_v=3$, $\lambda_e=0.01$, and $\lambda_w=1.0$ achieved minimum BIC, AIC, and reconstruction RMSE values.
-The lower value of $\lambda_w$ identified from these models reflects the presence of nonlinear mixing effects in the HSI data.
-While the values of model weights $W_{dm}$ for $m>N_v$ corresponding to nonlinear mixing were $0$ for the synthetic dataset, here these weights obtained a small but non-negligible median value of $0.0012$.
+As the table indicates, a non-linear GSM with $N_v=3$, $\lambda_e=0.01$, and $\lambda_w=1.0$ achieved minimum BIC, AIC, and reconstruction RMSE values.
+The lower value of $\lambda_w$ identified from these models reflects the presence of non-linear mixing effects in the HSI data.
+Although the values of model weights $W_{dm}$ for $m>N_v$ corresponding to non-linear mixing were $0$ for the synthetic data set, here these weights obtained a small but non-negligible median value of $0.0012$.
 
-Reflectance spectra generated by the trained GSM corresponding to maximum abundance values for each endmember are plotted in Figure~\ref{fig:robot-team-endmembers}.
+Reflectance spectra generated by the trained GSM corresponding to the maximum abundance values for each endmember are plotted in Figure~\ref{fig:robot-team-endmembers}.
 Based on their signatures, these endmembers are identified with water, vegetation (including near shore, filamentous blue-green algae) and the rhodmaine dye plume.
@@
 \end{adjustwidth}
-\caption{GSM applied to water spectra from real HSI dataset: \textbf{(a)} Spectra generated by the trained GSM for samples with maximum abundance for each endmember.
+\caption{GSM applied to water spectra from real HSI data set: \textbf{(a)} Spectra generated by the trained GSM for samples with maximum abundance for each endmember.
 Based on these spectral profiles, endmembers are identified with water, near-shore vegetation, and rhodamine dye.
@@
 
-To showcase the ability of the GSM to identify spatially localized contaminant sources, abundance maps for each individual endmember were generated by using the trained GTM to unmix all water pixels in the HSI.
+To showcase the ability of the GSM to identify spatially localized contaminant sources, abundance maps were generated for each individual end member using the trained GTM to unmix all water pixels in the HSI.
 In Figure~\ref{fig:endmember-abundance-dist} these abundance maps are compared for each endmember.
 The abundance map for the water endmember covers a majority of the center of the pond and decreases near the edge of the rhodamine plume where the dye and water mix.
-Vegetation in shallow water near the shore and the rhodamine dye plume in the western half of the pond are also clearly identified by their abundance maps.
-Given a HSI pixel resolution of $0.1 \times 0.1$ $\text{m}^2$, the spatial extent of vegetation is estimated to be $378.6$ $\text{m}^2$ while the plume initially extends across $255.7$ $\text{m}^2$.
+The vegetation in shallow water near the shore and the rhodamine dye plume in the western half of the pond are also clearly identified by their abundance maps.
+Given an HSI pixel resolution of $0.1 \times 0.1$ $\text{m}^2$, the spatial extent of vegetation is estimated to be $378.6$ $\text{m}^2$ while the plume initially extends across $255.7$ $\text{m}^2$.
 
@@
 
-Applying the trained GSM to unmix HSI from UAV flights provides an efficient way to map contaminant dispersion and transport.
-Figure~\ref{fig:plume-evo} demonstrates this by mapping the growing extent of the rhodamine dye between successive UAV flights.
-In just $15$ minutes, the plume expands from an initial area of $255.7$ $\text{m}^2$ to over $571.8$ $\text{m}^2$ as the dye mixes into the surrounding water.
+Applying the trained GSM to unmix HSI from UAV flights provides an efficient way to map the dispersion and transport of contaminant.
+Figure~\ref{fig:plume-evo} demonstrates this by mapping the growing extent of rhodamine dye between successive UAV flights.
+In just $15$ minutes, the plume expands from an initial area of $255.7$ $\text{m}^2$ to over $571.8$ $\text{m}^2$ as the dye mixes with the surrounding water.
 
@@
 
-The proliferation of hyperspectral imaging technologies in the remote sensing and environmental monitoring communities underscores the need for efficient algorithms which can make sense of high-dimensional HSI.
-Of particular interest are unsupervised algorithms which utilize all available wavelength bins to identify unique source profiles present realistic scenes.
-The GSM introduced in this paper is a novel method which simultaneously performs endmember extraction and spectral unmixing.
-Unlike popular endmember extraction algorithms such as VCA, PPI, and N-FINDR, the GSM does not assume the presence of pure pixels in HSI.
-Furthermore, the flexible structure of the mapping from the GSM latent space to the spectral data space allows the GSM to model nonlinear mixing effects, distinguishing it from other widely used linear models such as NMF.
-Being a probabilistic model also enables the GSM to quantify spectral variability via the precision parameter $\beta$ as indicated in Figure~\ref{fig:usgs-endmembers}.
+The proliferation of hyperspectral imaging technologies in the remote sensing and environmental monitoring communities underscores the need for efficient algorithms that can make sense of high-dimensional HSI.
+Of particular interest are unsupervised algorithms which utilize all available wavelength bins to identify unique source profiles that present realistic scenes.
+The GSM introduced in this paper is a novel method that simultaneously performs endmember extraction and spectral unmixing.
+Unlike popular endmember extraction algorithms, such as VCA, PPI, and N-FINDR, GSM does not assume the presence of pure pixels in HSI.
+Furthermore, the flexible structure of the mapping from the GSM latent space to the spectral data space allows the GSM to model non-linear mixing effects, distinguishing it from other widely used linear models such as NMF.
+Being a probabilistic model also enables the GSM to quantify the spectral variability through the precision parameter $\beta$ as indicated in Figure~\ref{fig:usgs-endmembers}.
 This additional capability is critical for analysis of realistic HSI where sensor noise, viewing geometry, and scene illumination all introduce uncertainty into extracted endmembers.
@@
 The growing popularity of deep learning methods has led many to explore applications of deep learning to the spectral unmixing problem.
-Variations on autoencoder architectures are a popular choice for both endmember extraction and unmixing.
+Variations in auto-encoder architectures are a popular choice for both endmember extraction and unmixing.
 For example, Borsoi et al.
-utilized a variational autoencoder to more accurately identify spectral variability in a LMM \cite{borsoi2019deep}.
+used a variational autoencoder to more accurately identify the spectral variability in a LMM \cite{borsoi2019deep}.
 Similarly, Palson et al.
@@
 However, the increased complexity of deep neural network approaches makes interpreting the latent space learned by encoder layers challenging and can dramatically increase training time and data volume requirements.
-In contrast, the latent space of the GSM is immediately interpretable due to the imposed simplex geometry while the form of $\psi$ leaves room to explore a variety of nonlinear mixing models.
-In this paper, $\psi$ was constructed to add nonlinear effects as additions or perturbations to an underlying linear mixing model with the $\lambda_w$ hyperparameter provided to control the degree of nonlinearity considered.
-In the original GTM formulation which inspired the GSM, the mapping $\psi$ strictly uses nonlinear features generated by Gaussian RBFs.
-Alternatively, one could easily construct $\psi$ for specific nonlinear models such as bilinear mixing by introducing features which depend on higher order combinations of the latent space (abundance) coordinates.
-It is important to note, though, that the EM algorithm formulated for GSM training is made possible due to the linear combination of activations $\phi(z)$ with model weights $\mathbf{W}$.
-For mixing models which instead depend nonlinearly on weights $\mathbf{W}$, an EM algorithm may require iterative nonlinear solves during each M step.
+In contrast, the latent space of the GSM is immediately interpretable due to the imposed simplex geometry, while the form of $\psi$ leaves room to explore a variety of non-linear mixing models.
+In this paper, $\psi$ was constructed to add non-linear effects as additions or perturbations to an underlying linear mixing model with the $\lambda_w$ hyperparameter provided to control the degree of non-linearity considered.
+In the original GTM formulation that inspired the GSM, the mapping $\psi$ strictly uses non-linear features generated by Gaussian RBFs.
+Alternatively, one could easily construct $\psi$ for specific non-linear models such as bilinear mixing by introducing features which depend on higher-order combinations of the latent space (abundance) coordinates.
+However, it is important to note that the EM algorithm formulated for GSM training is made possible due to the linear combination of activations $\phi(z)$ with model weights $\mathbf{W}$.
+For mixing models which instead depend non-linearly on weights $\mathbf{W}$, an EM algorithm may require iterative non-linear solves during each M step.
 
-The main limitation of the GSM is the curse-of-dimensionality encountered when generating a grid on the $(N_v-1)$-simplex for large numbers of endmembers, $N_v$.
+The main limitation of the GSM is the curse of dimensionality encountered when generating a grid on the $(N_v-1)$-simplex for large numbers of endmembers, $N_v$.
 This can be mitigated by instead randomly sampling points within the interior of the simplex using a uniform Dirichlet distribution to obtain a pre-determined number of nodes across the latent space.
-As the mixing coefficients $\pi_k$ are adapted during training, variability in inter-node spacing should not significantly impact model performance.
-Indeed, this was confirmed for the simulated dataset as summarized in Figure~\ref{fig:usgs-fits}.
-For considerably large datasets, the size of the responsibility matrix $\mathbf{R}$ can also lead to extended training times.
+As the mixing coefficients $\pi_k$ are adapted during training, the variability in the spacing of the internodes should not significantly affect the performance of the model.
+In fact, this was confirmed for the simulated data set as summarized in Figure~\ref{fig:usgs-fits}.
+For considerably large data sets, the size of the responsibility matrix $\mathbf{R}$ can also lead to extended training times.
 This can be addressed by augmenting the EM procedure to updated using mini-batches of training samples during each E-step as described by Bishop et al.
@@
 \cite{gtm-developments}.
-Rather than updating the full responsibility matrix, a subset of $\mathbf{R}$ corresponding to a single batch of training data can be evaluated with all other entries held constant.
-The GSM may also be extended in other ways, for example, by replacing the precision parameter $\beta$ with a covariance matrix to model wavelength dependent spectral variability common to many hyperspectral imaging platforms.
+Rather than updating the full responsibility matrix, a subset of $\mathbf{R}$ corresponding to a single batch of training data can be evaluated with all other entries kept constant.
+The GSM may also be extended in other ways, for example, by replacing the precision parameter $\beta$ with a covariance matrix to model wavelength-dependent spectral variability common to many hyperspectral imaging platforms.
 
-Based on the results from applying the GSM to real HSI as illustrated in Figure~\ref{fig:endmember-abundance-dist} and Figure~\ref{fig:plume-evo}, one clear application of the GSM is to contaminant identification and water quality assessment.
+Based on the results of applying the GSM to real HSI as illustrated in Figure~\ref{fig:endmember-abundance-dist} and Figure~\ref{fig:plume-evo}, one clear application of the GSM is to contaminant identification and water quality assessment.
 Modern UAV platforms make it possible to rapidly image bodies of water where direct access for in-situ data collection is restricted.
@@
 As a final consideration, we note that the problem of endmember extraction and spectral unmixing for HSI is identical to source apportionment in the context of air quality.
-Here the goal is to identify measurement profiles associated with sources of ambient air pollution based on measurements at a receptor site.
+Here, the goal is to identify measurement profiles associated with sources of ambient air pollution based on measurements at a receptor site.
 A popular model for source apportionment studies is Positive Matrix Factorization (PMF) introduced by Paatero and Tapper \cite{pmf-orig, ulbrich2009interpretation}.
 Just like NMF, PMF decomposes measurements into a linear combination of source profiles and relative abundances subject to non-negativity constraints.
-These linear receptor models assume that sources are not transformed during transport to the receptor, ignoring changes due to chemical reactions.
-Nonlinear mixing models like the GSM introduced here may therefore prove beneficial in this additional domain.
+These linear receptor models assume that the sources are not transformed during transport to the receptor, ignoring changes due to chemical reactions.
+Non-linear mixing models such as the GSM introduced here may, therefore, prove beneficial in this additional domain.
 
@@
 
-\textcolor{purple}{In this paper, we introduce the GSM as a novel method for unsupervised endmember extraction and spectral unmixing.
-The model was compared to three varieties of NMF on a synthetic dataset of linearly mixed spectra from the USGS spectral database.
+\textcolor{purple}{In this paper, we introduced the GSM as a novel method for unsupervised endmember extraction and spectral unmixing.
+The model was compared to three varieties of NMF on a synthetic data set of linearly mixed spectra from the USGS spectral database.
 Across all noise levels, the GSM outperformed NMF for both endmember accuracy and abundance estimation.
-When applied to real HSI collected over a north Texas pond, the GSM accurately models spectral mixing including nonlinear effects.
+When applied to real HSI collected over a north Texas pond, the GSM accurately models spectral mixing including non-linear effects.
 Rhodamine dye was released into the pond to simulate a localized contaminant source.
 Endmember abundances estimated by the GSM corresponding to the dye accurately tracked the evolution of the plume as it mixed into the surrounding water.
-These examples illustrate the power of combining nonlinear mixing models together with UAV-based hyperspectral imaging for environmental monitoring and water quality analysis.
+These examples illustrate the power of combining non-linear mixing models together with UAV-based hyperspectral imaging for environmental monitoring and water quality analysis.
 Future work will further develop the GSM to incorporate wavelength dependent spectral variability.
-Additionally the GSM may prove useful for air quality studies as a nonlinear source apportionment model.}
+Additionally the GSM may prove useful for air quality studies as a non-linear source apportionment model.}
 
@@
 \dataavailability{\textcolor{red}{We encourage all authors of articles published in MDPI journals to share their research data.
-In this section, please provide details regarding where data supporting reported results can be found, including links to publicly archived datasets analyzed or generated during the study.
+In this section, please provide details regarding where data supporting reported results can be found, including links to publicly archived data sets analyzed or generated during the study.
 Where no new data were created, or where data is unavailable due to privacy or ethical restrictions, a statement is still required.

### 2024-08-19 paper/paper.tex
fixed grammar mistakes

@@
 The spatial resolution of hyperspectral imagers generally results in pixels with mixed signals from multiple sources called endmembers.
-The task of unsupervised source identification using HSI data therefore involves two steps: endmember extraction, and abundance estimation.
+The task of unsupervised source identification using HSI data therefore involves two steps: endmember extraction and abundance estimation.
 Techniques such as vertex component analysis (VCA), the pixel purity index (PPI), and N-FINDR solve this first task by identifying HSI endmember spectra assuming the presence of some pure (unmixed) pixels \cite{vca-orig, ppi-orig, N-FINDR-orig}.
@@
 The UAV flights were performed near solar noon to maximize the amount of sunlight illuminating the water.
-For this Pond in North Texas, this corresponded to an average solar zenith angle of $56.7^{\circ}$ resulting in HSI with negligible sunglint effects.
+For this pond in North Texas, this corresponded to an average solar zenith angle of $56.7^{\circ}$ resulting in HSI with negligible sunglint effects.
 
@@
 Generating semantic segmentations of collected HSI as in panel b of Figure~\ref{fig:robot-team-endmembers} makes it easy to compare multiple HSI without resorting to pseudocolor images generated from a limited number of wavelength bands.
-Since the GSM models the distribution of all HSI spectra rather than individual parameters, it may also aide in-situ data collection by suggesting sampling points which maximize the area traversed in the GSM latent space rather than uniformly sampling across a wide spatial extent.
+Since the GSM models the distribution of all HSI spectra rather than individual parameters, it may also aid in-situ data collection by suggesting sampling points which maximize the area traversed in the GSM latent space rather than uniformly sampling across a wide spatial extent.
 Similar approaches have been developed to guide autonomous data collection vehicles by casting route planning as a prize collecting travelling salesman problem subject to resource constraints \cite{balas2007prize, suryan2020learning}.

### 2024-08-26 2c_robotteam-plots.jl
updated figure 8 to include abundance in label

@@ -336,7 +336,7 @@ heatmap!(ax, lon_min..lon_max, lat_min..lat_max, Zpred[:, :,1], colormap=cmap, c
 cb = Colorbar(fig[1,2],
               colormap=cmap, colorrange=clims,
               ticklabelsize=15,
-              label="Vegetation",
+              label="Vegetation Abundance",
               labelsize=16,
               minorticksvisible=true
               )
@@ -387,7 +387,7 @@ heatmap!(ax, lon_min..lon_max, lat_min..lat_max, Zpred[:, :,2], colormap=cmap, c
 cb = Colorbar(fig[1,2],
               colormap=cmap, colorrange=clims,
               ticklabelsize=15,
-              label="Water",
+              label="Water Abundance",
               labelsize=16,
               minorticksvisible=true,
               )

### 2024-08-26 paper/paper.tex

@@
 % Abstract (Do not insert blank lines, i.e. \\) 
-\abstract{\textcolor{violet}{In this paper, we introduce a new model for non-linear endmember extraction and spectral unmixing of hyperspectral imagery called Generative Simplex Mapping (GSM).
+\abstract{In this paper, we introduce a new model for non-linear endmember extraction and spectral unmixing of hyperspectral imagery called Generative Simplex Mapping (GSM).
 The model represents endmember mixing using a latent space with points sampled within a $(n-1)$-simplex corresponding to the abundance of $n$ unique sources.
@@
 The model is able to accurately identify spectral signatures corresponding to near-shore algae, water, and rhodmaine tracer dye introduced into the pond to simulate water contamination by a localized source.
-Abundance maps generated using the GSM accurately track evolution of the dye plume as it mixes into the surrounding water.}}
+Abundance maps generated using the GSM accurately track evolution of the dye plume as it mixes into the surrounding water.}
 
@@
 
-\textcolor{purple}{In this paper, we introduced the GSM as a novel method for unsupervised endmember extraction and spectral unmixing.
+In this paper, we introduced the GSM as a novel method for unsupervised endmember extraction and spectral unmixing.
 The model was compared to three varieties of NMF on a synthetic data set of linearly mixed spectra from the USGS spectral database.
@@
 Future work will further develop the GSM to incorporate wavelength dependent spectral variability.
-Additionally the GSM may prove useful for air quality studies as a non-linear source apportionment model.}
+Additionally the GSM may prove useful for air quality studies as a non-linear source apportionment model.
 
@@
 
-\funding{\textcolor{red}{This research was funded by the following grants: the Texas National Security Network Excellence Fund award for Environmental Sensing Security Sentinels; the SOFWERX award for Machine Learning for Robotic Teams and NSF Award OAC-2115094; support from the University of Texas at Dallas Office of Sponsored Programs, Dean of Natural Sciences and Mathematics, and Chair of the Physics Department is gratefully acknowledged; TRECIS CC* Cyberteam (NSF \#2019135); NSF OAC-2115094 Award; and EPA P3 grant number 84057001-0.}}
+\funding{This research was funded by the following grants: the Texas National Security Network Excellence Fund award for Environmental Sensing Security Sentinels; the SOFWERX award for Machine Learning for Robotic Teams and NSF Award OAC-2115094; support from the University of Texas at Dallas Office of Sponsored Programs, Dean of Natural Sciences and Mathematics, and Chair of the Physics Department is gratefully acknowledged; TRECIS CC* Cyberteam (NSF \#2019135); NSF OAC-2115094 Award; and EPA P3 grant number 84057001-0.}
 
 \institutionalreview{Not applicable.}
-
 \informedconsent{Not applicable.}
 
-\dataavailability{\textcolor{red}{We encourage all authors of articles published in MDPI journals to share their research data.
-In this section, please provide details regarding where data supporting reported results can be found, including links to publicly archived data sets analyzed or generated during the study.
-Where no new data were created, or where data is unavailable due to privacy or ethical restrictions, a statement is still required.
-Suggested Data Availability Statements are available in section ``MDPI Research Data Policies'' at \url{https://www.mdpi.com/ethics}.}}
+\dataavailability{The hyperspectral images supporting the conclusions of this article will be made available by the authors on request.
+An implementation of the GSM model in the Julia programming language is available at \url{https://github.com/john-waczak/GenerativeTopographicMapping.jl}, accessed on 21 August 2024.}
 
-\acknowledgments{\textcolor{red}{David Shaefer is gratefully acknowledged for his help coordinating UAV flights during HSI collection.
+\acknowledgments{David Shaefer is gratefully acknowledged for his help coordinating UAV flights during HSI collection.
 We thank Adam Aker, Lakitha Wijeratne, Shawhin Talebi, Ashen Fernando, Prabuddha Dewage, Mazhar Iqbal, Matthew Lary, and Gokul Balagopal for their help deploying the UAV.
-Don MacLaughlin, Scotty MacLaughlin, and the city of Plano, TX, are gratefully acknowledged for allowing us to deploy the autonomous robot team on their property.
-Christopher Simmons is gratefully acknowledged for his computational support.
-We thank  Antonio Mannino for his advice with regard to selecting the robotic boat's sensing suite.
-Annette Rogers is gratefully acknowledged for supporting the arrangement of insurance coverage.
-Steven Lyles is gratefully acknowledged for supporting the arrangement of a secure place for the robot team.}}
+Don MacLaughlin, Scotty MacLaughlin, and the city of Plano, TX, are gratefully acknowledged for allowing us to deploy the UAV on their property.
+Christopher Simmons is gratefully acknowledged for his computational support.}
 

### 2024-10-10 paper/paper.tex
updated citations to include doi's where possible for all sources
Updates from Overleaf

@@
 In this paper, we introduce a new variant of the GTM dubbed Generative Simplex Mapping (GSM), which can extract endmember spectra and unmix non-linear mixtures.
-By replacing the rectangular latent space of the GTM with a gridded $n$-simplex, the vertices of the GSM can be immediately interpreted as endmembers corresponding to $n+1$ unique spectral signatures.
+By replacing the rectangular latent space of the GTM with a gridded $(n-1)$-simplex, the vertices of the GSM can be immediately interpreted as endmembers corresponding to $n$ unique spectral signatures.
 The mapping from the latent space to the HSI data space models signal mixing while barycentric coordinates for the latent space simplex estimate relative endmember abundances.
@@
 
-In the standard LMM model, the mapping $\psi$ is given by $\psi(z;W) = \mathbf{W}z$ where the columns of $\mathbf{W}$ correspond to endmember spectra.
+In the standard LMM model, the mapping $\psi$ is given by $\psi(z;\mathbf{W}) = \mathbf{W}z$ where the columns of $\mathbf{W}$ correspond to endmember spectra.
 To model non-linear mixing, $z$ is replaced by the output of $M$-many activation functions such that $\psi(z;W) = \mathbf{W}\phi(z)$.
@@
 After initialization, the expectation and maximization steps are repeated in turn until $Q$ converges to a predetermined tolerance level.
-For large $N_v$ we note that generating a regular grid within the simplex becomes cumbersome as the number grid nodes scales as ${k + N_v - 2 \choose N_v - 1}$ for $k$ nodes per edge.
+For large $N_v$ we note that generating a regular grid within the simplex becomes cumbersome as the number of grid nodes scales as ${k + N_v - 2 \choose N_v - 1}$ for $k$ nodes per edge.
 An alternative approach is to randomly sample points within the simplex to obtain a total of $K$ nodes.
@@
 \begin{equation}
-    \theta(\rho_i, \hat{\rho}_i) =  \arcos\left( \dfrac{\langle \rho_i, \hat{\rho}_i \rangle}{\lVert \rho_i \rVert \cdot \lVert \hat{\rho}_i \rVert}\right)
+    \theta(\rho_i, \hat{\rho}_i) =  \arccos\left( \dfrac{\langle \rho_i, \hat{\rho}_i \rangle}{\lVert \rho_i \rVert \cdot \lVert \hat{\rho}_i \rVert}\right)
 \end{equation}

### 2024-10-10 paper/paper.tex
updated mathematical notation to reflect that used in the dissertation

@@
 \graphicspath{{figures/}}
+\usepackage{pdfpages}
 
@@
 
-In the original GTM formulation, the data vectors $x$ (reflectance spectra) are described by latent variables $z$ mapped into the data space by a non-linear function $\psi$.
+In the original GTM formulation, the data vectors $\mathbf{x}$ (reflectance spectra) are described by latent variables $\mathbf{z}$ mapped into the data space by a non-linear function $\psi$.
 The data space distribution is taken to be normal with the precision parameter $\beta$ to account for measurement noise and spectral variability.
@@
 \begin{equation}\label{eqn:data-space-distribution}
-    p(x \mid z, \mathbf{W}, \beta) = \left(\frac{\beta}{2\pi} \right)^{D/2}\exp\left( -\frac{\beta}{2}\lVert \psi(z; \mathbf{W}) - x \rVert^2 \right)
+    p(\mathbf{x} \mid \mathbf{z}, \mathbf{W}, \beta) = \left(\frac{\beta}{2\pi} \right)^{D/2}\exp\left( -\frac{\beta}{2}\lVert \psi(\mathbf{z}; \mathbf{W}) - \mathbf{x} \rVert^2 \right)
 \end{equation}
@@
 \begin{equation}\label{eqn:latent-distribution}
-    p(z) = \sum\limits_k^K \pi_k \, \delta(z - z_k) \quad\text{where}\quad  \sum_k^K\pi_k = 1.
+    p(\mathbf{z}) = \sum\limits_k^K \pi_k \, \delta(\mathbf{z} - \mathbf{z}_k) \quad\text{where}\quad  \sum_k^K\pi_k = 1.
 \end{equation}
@@
 \begin{equation}\label{eqn:llh}
-    \mathcal{L} = \sum\limits_n^N \ln \left( \sum\limits_k^K \pi_k \, p(x_n \mid z_k, \mathbf{W}, \beta) \right)
+    \mathcal{L} = \sum\limits_n^N \ln \left( \sum\limits_k^K \pi_k \, p(\mathbf{x}_n \mid \mathbf{z}_k, \mathbf{W}, \beta) \right)
 \end{equation}
@@
 
-In the standard LMM model, the mapping $\psi$ is given by $\psi(z;\mathbf{W}) = \mathbf{W}z$ where the columns of $\mathbf{W}$ correspond to endmember spectra.
-To model non-linear mixing, $z$ is replaced by the output of $M$-many activation functions such that $\psi(z;W) = \mathbf{W}\phi(z)$.
-The activations applied to each GSM node can then be collected to form a matrix with elements $\Phi_{km} = \phi_m(z_k)$.
-For $m \leq N_v$, we take $\Phi_{km} = z_k$ to model linear mixing.
+In the standard LMM model, the mapping $\psi$ is given by $\psi(\mathbf{z};\mathbf{W}) = \mathbf{W}\mathbf{z}$ where the columns of $\mathbf{W}$ correspond to endmember spectra.
+To model non-linear mixing, $\mathbf{z}$ is replaced by the output of $M$-many activation functions such that $\psi(\mathbf{z};\mathbf{W}) = \mathbf{W}\phi(\mathbf{z})$.
+The activations applied to each GSM node can then be collected to form a matrix with elements $\Phi_{km} = \phi_m(\mathbf{z}_k)$.
+For $m \leq N_v$, we take $\Phi_{km} = [\mathbf{z}_k]_m$ to model linear mixing.
 The remaining  $M-N_v$ activations are computed using radial basis functions (RBF) with centers $\mu_m$ distributed throughout the simplex (but not at the vertices) with
@@
     \Phi_{km} = \begin{cases}
-        \dfrac{s - \lVert z_k - \mu_m \rVert}{s}; & \lVert z_k - \mu_m \rVert \leq s \\
-        0; & \lVert z_k - \mu_m \rVert > s
+        \dfrac{s - \lVert \mathbf{z}_k - \mu_m \rVert}{s}; & \lVert \mathbf{z}_k - \mu_m \rVert \leq s \\
+        0; & \lVert \mathbf{z}_k - \mu_m \rVert > s
     \end{cases}
@@
 Here, $N_v=3$ has been chosen for illustrative purposes.
-Nodes are mapped into the data space via the map $\psi(z)$ utilizing $M$-many radially symmetric basis functions (red).
-Spectral variability is estimated via the precision parameter $\beta$ shown here in the data space as a light blue band around the spectrum given by $\psi(z)$.}
+Nodes are mapped into the data space via the map $\psi(\mathbf{z})$ utilizing $M$-many radially symmetric basis functions (red).
+Spectral variability is estimated via the precision parameter $\beta$ shown here in the data space as a light blue band around the spectrum given by $\psi(\mathbf{z})$.}
 \label{fig:gsm-diagram}
@@
 \begin{equation}\label{eq:responsibility}
-    R_{kn}  = p(z_k \mid x_n, \mathbf{W}, \beta) = \dfrac{\pi_k \, p(x_n \mid z_k, \mathbf{W}, \beta)}{\sum\limits_{k'}^K \pi_{k'} \, p(x_n \mid z_{k'}, \mathbf{W}, \beta)}.
+    R_{kn}  = p(\mathbf{z}_k \mid \mathbf{x}_n, \mathbf{W}, \beta) = \dfrac{\pi_k \, p(\mathbf{x}_n \mid \mathbf{z}_k, \mathbf{W}, \beta)}{\sum\limits_{k'}^K \pi_{k'} \, p(\mathbf{x}_n \mid \mathbf{z}_{k'}, \mathbf{W}, \beta)}.
 \end{equation}
@@
 \begin{equation}\label{eq:beta-update}
-    \frac{1}{\beta^{\text{new}}}  = \frac{1}{ND}\sum\limits_n^N\sum\limits_k^K R_{kn}\lVert \psi(z_k; \mathbf{W}) - x_n \rVert^2.
+    \frac{1}{\beta^{\text{new}}}  = \frac{1}{ND}\sum\limits_n^N\sum\limits_k^K R_{kn}\lVert \psi(\mathbf{z}_k; \mathbf{W}) - \mathbf{x}_n \rVert^2.
 \end{equation}
@@
 \begin{equation}
-    p(z_1,...z_{N_v}) = \frac{\Gamma(\sum_i^{N_v}\alpha_i)}{\prod_{i}^{N_v}\Gamma(\alpha_i)} \prod_{i}^{N_v} z_i^{\alpha_i-1}
+    p(\mathbf{z}) = \frac{\Gamma(\sum_i^{N_v}\alpha_i)}{\prod_{i}^{N_v}\Gamma(\alpha_i)} \prod_{i}^{N_v} [\mathbf{z}]_i^{\alpha_i-1}
 \end{equation}
@@
 \begin{equation}
-    \hat{z}_n = \sum_k^K R_{kn}z_k.
+    \hat{\mathbf{z}}_n = \sum_k^K R_{kn}\mathbf{z}_k.
 \end{equation}
@@
 A popular model for source apportionment studies is Positive Matrix Factorization (PMF) introduced by Paatero and Tapper \cite{pmf-orig, ulbrich2009interpretation}.
-Just like NMF, PMF decomposes measurements into a linear combination of source profiles and relative abundances subject to non-negativity constraints.
+Just like NMF, PMF decomposes measurements into a convex combination of source profiles and relative abundances subject to non-negativity constraints.
 These linear receptor models assume that the sources are not transformed during transport to the receptor, ignoring changes due to chemical reactions.
@@
 \section{Patents}
-
 \textcolor{red}{UPDATE ME!}

### 2024-10-14 paper/paper.tex
Updates from Overleaf

@@
 % Abstract (Do not insert blank lines, i.e. \\) 
-\abstract{In this paper, we introduce a new model for non-linear endmember extraction and spectral unmixing of hyperspectral imagery called Generative Simplex Mapping (GSM).
+\abstract{We introduce a new model for non-linear endmember extraction and spectral unmixing of hyperspectral imagery called Generative Simplex Mapping (GSM).
 The model represents endmember mixing using a latent space with points sampled within a $(n-1)$-simplex corresponding to the abundance of $n$ unique sources.
@@
 However, these supervised methods rely on \textit{a priori} knowledge of expected sources to identify appropriate reference sensors for data collection.
-In light of these limitations, unsupervised methods are needed that can expediently extract source signatures from high-dimensional HSI.
+In light of these limitations, unsupervised methods are needed which can reliably extract source signatures from high-dimensional HSI.
 
@@
 With the growing popularity of deep learning approaches in remote sensing, a variety of models based on autoencoder architectures have been introduced for unmixing HSI data \cite{non-negative-autoencoders,su2019daen,borsoi2019deep,palsson2020convolutional}.
-However, the complexity introduced by these models significantly impacts training time and decreases the interpretability of the model.
+However, the complexity introduced by these models significantly impacts training time and decreases the interpretability of the resulting model.
 An ideal approach should enable both endmember extraction and non-linear unmixing while accounting for spectral variability.
@@
 When choosing a two-dimensional latent space, the GTM can be used to visualize the distribution of HSI spectra while mapping the latent space nodes to the HSI data space provides endmembers \cite{robot-team-gtm}.
-Unfortunately, the rectangular latent space grid employed by the GTM does not directly translate into endmember abundances.
-Additionally, the expectation-maximization (EM) algorithm used to train the GTM does not guarantee non-negativity of GTM node spectra.
+Unfortunately, the rectangular latent space grid employed by the GTM does not directly translate into endmember abundances, and the expectation-maximization (EM) algorithm used to train the GTM does not guarantee non-negativity of GTM node spectra.
 
@@
 By replacing the rectangular latent space of the GTM with a gridded $(n-1)$-simplex, the vertices of the GSM can be immediately interpreted as endmembers corresponding to $n$ unique spectral signatures.
-The mapping from the latent space to the HSI data space models signal mixing while barycentric coordinates for the latent space simplex estimate relative endmember abundances.
-Furthermore, by taking inspiration from the multiplicative updates of NMF, the GSM algorithm maintains the non-negativity of resulting endmember spectra.
+The mapping from the latent space to the HSI data space models signal mixing while barycentric coordinates for the latent space simplex represent relative endmember abundances.
+Additionally, by taking inspiration from the multiplicative updates of NMF, the GSM algorithm maintains the non-negativity of resulting endmember spectra.
 If only linear mixing is present, the GSM algorithm drives non-linear contributions to $0$.
@@
 Section~\ref{sec:gsm} describes the proposed method.
-Section~\ref{sec:experiments} and Section~\ref{sec:results} describe experiments using real and simulated HSI data to evaluate the GSM model.
+Section~\ref{sec:experiments} and Section~\ref{sec:results} describe experiments using simulated and real HSI data to evaluate the GSM model.
 Section~\ref{sec:disucssion} discusses additional applications and extensions of the GSM to be explored in future work.
@@
 
-\section{Patents}
-\textcolor{red}{UPDATE ME!}
+%\section{Patents}
 

### 2024-11-05 paper/paper.tex
Addedd patent section
updated for first round of review

@@
 \abstract{We introduce a new model for non-linear endmember extraction and spectral unmixing of hyperspectral imagery called Generative Simplex Mapping (GSM).
-The model represents endmember mixing using a latent space with points sampled within a $(n-1)$-simplex corresponding to the abundance of $n$ unique sources.
-Points in this latent space are non-linearly mapped to reflectance spectra via a flexible function combining linear and non-linear mixing.
+The model represents endmember mixing using a latent space of points sampled within a $(n-1)$-simplex corresponding to $n$ unique sources.
+Barycentric coordinates within this simplex are naturally interpreted as relative endemember abundances satisfying both the abundance sum-to-one and abundance non-negativity constraints.
+Points in this latent space are mapped to reflectance spectra via a flexible function combining linear and non-linear mixing.
 Due to the probabilistic formulation of the GSM, spectral variability is also estimated by a precision parameter describing the distribution of observed spectra.
-Model parameters are determined using a generalized expectation-maximization algorithm.
-In the event of purely linear mixing, non-linear contributions are naturally driven to zero.
+Model parameters are determined using a generalized expectation-maximization algorithm which guarantees non-negativity for extracted endmembers.
+We first compare the GSM against three varieties of nonnegative matrix factorization (NMF) on a sythetic dataset of linearly mixed spectra from the USGS spectral database.
+Here the GSM performed favorably for both endmember accacy and abundance estimation with the all non-linear contributions driven to zero by the fitting procedure.
+In a second experiment, we apply the GTM to model non-linear mixing in real hyperspectral imagery captured over a pond in North Texas.
+The model accurately identified spectral signatures corresponding to near-shore algae, water, and rhodmaine tracer dye introduced into the pond to simulate water contamination by a localized source.
+Abundance maps generated using the GSM accurately track the evolution of the dye plume as it mixes into the surrounding water.}
+
+% In the event of purely linear mixing, non-linear contributions are naturally driven to zero.
 The GSM outperforms three varieties of non-negative matrix factorization for both endmember extraction accuracy and abundance estimation on a synthetic data set of linearly mixed spectra from the USGS spectral library.
@@
 Techniques such as vertex component analysis (VCA), the pixel purity index (PPI), and N-FINDR solve this first task by identifying HSI endmember spectra assuming the presence of some pure (unmixed) pixels \cite{vca-orig, ppi-orig, N-FINDR-orig}.
-By further assuming a linear mixing model (LMM) in which observed spectra are described by a linear combination of endmembers with non-negative abundances, HSI can be unmixed using a variety of techniques such as constrained least squares \cite{spectral-unmixing-orig, fcls-unmixing}.
+By further assuming a linear mixing model (LMM) in which observed spectra are described by a linear combination of endmembers with non-negative abundances, HSI can be unmixed using a variety of techniques such as constrained least squares and multi-objective optimization \cite{spectral-unmixing-orig, fcls-unmixing, gmogh}.
 Among these methods, Nonnegative Matrix Factorization (NMF) is a widely used approach which extracts endmember spectra and unmixes abundances simultaneously via matrix factorization \cite{nmf-orig, unmixing-nmf-review, unmixing-nmf-review-2}.
@@
 introduced the Generative Topographic Mapping (GTM), a probabilistic latent-variable model inspired by the SOM \cite{gtm-orig}.
-When choosing a two-dimensional latent space, the GTM can be used to visualize the distribution of HSI spectra while mapping the latent space nodes to the HSI data space provides endmembers \cite{robot-team-gtm}.
+When choosing a two-dimensional latent space, the GTM can be used to visualize the distribution of HSI spectra while mapping the latent space nods to the HSI data space provides endmembers \cite{robot-team-gtm}.
 Unfortunately, the rectangular latent space grid employed by the GTM does not directly translate into endmember abundances, and the expectation-maximization (EM) algorithm used to train the GTM does not guarantee non-negativity of GTM node spectra.
 
-In this paper, we introduce a new variant of the GTM dubbed Generative Simplex Mapping (GSM), which can extract endmember spectra and unmix non-linear mixtures.
+In this paper, we introduce a new variant of the GTM, dubbed Generative Simplex Mapping (GSM), which can extract endmember spectra and unmix non-linear mixtures.
 By replacing the rectangular latent space of the GTM with a gridded $(n-1)$-simplex, the vertices of the GSM can be immediately interpreted as endmembers corresponding to $n$ unique spectral signatures.
-The mapping from the latent space to the HSI data space models signal mixing while barycentric coordinates for the latent space simplex represent relative endmember abundances.
+The mapping from the latent space to the HSI data space accounts for linear and non-linear spectral mixing while barycentric coordinates for the latent space simplex represent relative endmember abundances.
 Additionally, by taking inspiration from the multiplicative updates of NMF, the GSM algorithm maintains the non-negativity of resulting endmember spectra.
@@
 Prior distributions included for GSM model weights yield hyperparameters which can be tuned to control the smoothness of the resulting spectra and the degree of non-linear mixing applied.
+In summary, the key innovations introduced by the GSM are:
+\begin{itemize}
+    \item the GSM can model linear \textit{and} nonlinear spectral mixing
+    \item the GSM does not assume the presence of pure pixels in the dataset 
+    \item the probabilistic formulation of the GSM accounts for spectral variability
+    \item the simplex used for the latent space structure of the GSM is directly interpretable and forces abundances to satisfy both the abundance sum-to-one and abundance non-negativity constraints
+    \item the fiting procedure introduced for the GSM maintains non-negativity of endmember spectra
+\end{itemize}
+
+%  \textcolor{red}{ In summary, the key innovations of the GSM are
+% \begin{itemize}
+%     \item Modeling of linear and non-linear spectral mixing
+%     \item The model is probabilistic, with spectral varaibility accounted for via a precision parameter
+%     \item Satisfaction of constraints
+%     \item No pure-pixel assumption
+% \end{itemize}    
+% }
 
@@
 
-In the original GTM formulation, the data vectors $\mathbf{x}$ (reflectance spectra) are described by latent variables $\mathbf{z}$ mapped into the data space by a non-linear function $\psi$.
+In the original GTM formulation, the data vectors $\mathbf{x}$ (reflectance spectra of length $D$) are described by latent variables $\mathbf{z}$ mapped into the data space by a non-linear function $\psi$.
 The data space distribution is taken to be normal with the precision parameter $\beta$ to account for measurement noise and spectral variability.
@@
 \end{equation}
-where $\mathbf{W}$ are model weights which parameterize the mapping $\psi$.
+where $\mathbf{z}=(z_1, z_2, ..., z_{N_v})$ corresponding to $N_v$-many sources and $\mathbf{W}$ is a $D \times M$ matrix of model weights which parameterize the mapping $\psi$.
 
@@
 The first is to replace the rectangular GTM grid with a gridded simplex with $N_v$ vertices.
-The barycentric coordinates for each node then describe the relative abundance of each endmember with each vertex corresponding to a pure endmember.
-The second change is to replace equal prior probabilities with adaptive mixing coefficients $\pi_k$ to allow the GSM to model HSI with nonuniform mixing distributions.
-Together, this leads to a latent space prior distribution given by
+Barycentric coordinates for each node then describe the relative abundance of each endmember with siplex vertices corresponding to pure sources.
+The second change is to augment the equal prior probabilities of the GTM latent space with adaptive mixing coefficients $\pi_k$ to allow for nonuniform sampling across the abundance simplex.
+This addition allows the GSM to flexibly address many possible mixing scenarios without an \textit{a priori} assumption for a specific mixing distribution.
+Together, these changes result in a prior distribution for the latent space given by
 \begin{equation}\label{eqn:latent-distribution}
@@
 \end{equation}
+where $\delta(\cdot)$ is the Dirac delta function and $\mathbf{z}_k$ are the positions of each node within the simplex.
 
@@
 where $s$ is the spacing between RBF centers.
+A visual representation of the non-linear activation functions in Equation~\ref{eq:act-function} is shown below in Figure~\ref{fig:act-func} for a 2-component GSM.
+\begin{figure}[!b]
+    \centering
+    \includegraphics[width=0.45\columnwidth]{methods/gsm/activation-function.pdf}
+    \caption{Illustration of the non-linear activation functions for a 2-component GSM.
+The simplex vertices are shown as green dots ($\mu_1$ and $\mu_2$) and three RBF centers are shown in red ($\mu_1$, $\mu_2$, and $\mu_3$).
+With no RBF centers at the vertices, we guarantee no non-linear contributions occur for pure endmember spectra.}
+    \label{fig:act-func}
+\end{figure}
 In this form, the first $N_v$ columns of $\mathbf{W}$ correspond to endmember spectra, while the remaining columns account for additional non-linear effects.
-For linear mixing, the GSM training algorithm should therefore drive $W_{dm}$ to $0$ for $m\geq N_v$.
-A visualization of the GSM model is shown in Figure~\ref{fig:gsm-diagram}.
+Equation~\ref{eq:act-function} is specifically chosen so that no non-linear contributions are possible for pure spectra at the vertices of the simplex.
+At all other points, the output of $\psi$ involves both linear and non-linear contributions.
+If only linear mixing is present, the GSM training algorithm should therefore drive $W_{dm}$ to $0$ for $m\geq N_v$.
+A visualization of the GSM model for three sources is shown in Figure~\ref{fig:gsm-diagram}.
 
-\begin{figure}[H]
+\begin{figure}
+    \centering
 \includegraphics[width=\columnwidth]{methods/gsm/gsm-diagram.png}
@@
     Q &= \sum_n^N\sum_k^K R_{kn} \left(\ln\pi_k + \frac{D}{2}\ln\left(\frac{\beta}{2\pi}\right) - \frac{\beta}{2}\sum_d^D\left(\sum_m^M W_{dm}\Phi_{km} - X_{nd}\right)^2\right) \\ 
-    &\qquad + \frac{N_vD}{2}\ln\left(\frac{\lambda_e}{2\pi}\right) - \frac{\lambda}{2}\sum_d^D \sum_{m=1}^{N_v} W_{dm}^2  \\ 
+    &\qquad + \frac{N_vD}{2}\ln\left(\frac{\lambda_e}{2\pi}\right) - \frac{\lambda_e}{2}\sum_d^D \sum_{m=1}^{N_v} W_{dm}^2  \\ 
     &\qquad + (M-N_v)D\ln\left(\dfrac{\lambda_w}{2}\right) - \lambda_w\sum_d^D\sum_{m=N_v+1}^{M} W_{dm}
@@
 
-To illustrate the effectiveness of the GSM, we first demonstrate its ability to model simple linear mixing by driving non-linear weights to zero during the fitting process.
+To illustrate the effectiveness of the GSM, we first demonstrate its ability to model linear mixing.
+This serves as an important limiting case since linearly mixed spectra should not lead to the spurous introduction of non-linear contributions by the GSM.
+The goal of this first test is therefore to demonstrate that the GSM drives non-linear weights to zero for linearly mixed data while providing a fair test to compare GSM to well-established linear mixing models.
 To this end, a synthetic data set comprising linear mixtures of three sample spectra from the U.S.
@@
 \begin{figure}[H]
-\includegraphics[width=\columnwidth]{paper/figures/methods/usgs/usgs-dataset.pdf}
+\includegraphics[width=\columnwidth]{methods/usgs/usgs-dataset.pdf}
 \caption{Synthetic data set formed from USGS spectra.
@@
 
-For comparison, NMF models were also trained on each data set as this method is both highly popular and does not include the pure-pixel assumption common to other technqieus like VCA and PPI.
-Countless variations on the original NMF method have been introduced into the literature with one review identifying more than $100$ distinct NMF variations \cite{unmixing-nmf-review}.
-For the purpose of evaluating the GSM, we considered the standard $\ell_2$ and KL-divergence formulations introduced by Lee and Seung \cite{nmf-algorithms}.
+For comparison, three NMF models were also trained on each data set as this method is both highly popular and does not include the pure-pixel assumption common to other techniques like VCA and PPI.
+Countless variations on the original NMF method have been introduced into the literature with one review identifying more than $100$ distinct NMF varieties \cite{unmixing-nmf-review}.
+For the purpose of comparing to the GSM, we considered the standard $\ell_2$ and KL-divergence formulations introduced by Lee and Seung \cite{nmf-algorithms}.
 We also included the robust $\ell_{2,1}$ NMF as described by Kong et al \cite{nmf-l21}.
+We note that the goal of this test is not to prove the GSM is superior to other models for linear mixing, but rather to demonstrate that the GSM can model linearly mixed data without introducing unnecessary complexity.
 
@@
 Here we again see that both versions of the GSM outperformed NMF.
-To justify that all models were fairly trained to convergence, the RMSE data reconstruction was also computed.
+To justify that all models were fairly trained to convergence, the data reconstruction RMSE was also computed.
 This metric uses the trained mixing model to compute the error between the original data set and the reconstructed spectra generated using the extracted endmembers and their abundances.
-For all GSM and NMF models, the RMSE data reconstruction converged to the level of random noise introduced into the data.
+For all GSM and NMF models, the data reconstruction RMSE converged to the level of random noise introduced into the data.
 These results reflect a fair comparison between the NMF and GSM models.
@@
 
-%\section{Patents}
+\section{Patents}
+A provisional US patent application for this work has been filed on 2024-10-10 entitled: \textit{Generative Simplex Mapping: Non-Linear Endmember Extraction And Spectral Unmixing For Hyperspectral Imagery}, application number 63/705,854.
 

### 2024-11-07 paper/paper.tex
updated paper for first round of review

@@
 We first compare the GSM against three varieties of nonnegative matrix factorization (NMF) on a sythetic dataset of linearly mixed spectra from the USGS spectral database.
-Here the GSM performed favorably for both endmember accacy and abundance estimation with the all non-linear contributions driven to zero by the fitting procedure.
+Here the GSM performed favorably for both endmember accacy and abundance estimation with all non-linear contributions driven to zero by the fitting procedure.
 In a second experiment, we apply the GTM to model non-linear mixing in real hyperspectral imagery captured over a pond in North Texas.
@@
 In the original GTM formulation, the data vectors $\mathbf{x}$ (reflectance spectra of length $D$) are described by latent variables $\mathbf{z}$ mapped into the data space by a non-linear function $\psi$.
-The data space distribution is taken to be normal with the precision parameter $\beta$ to account for measurement noise and spectral variability.
+The data space distribution is taken to be normal with a precision parameter $\beta$ to account for measurement noise and spectral variability.
 The GSM uses same structure, that is
@@
 where $\mathbf{z}=(z_1, z_2, ..., z_{N_v})$ corresponding to $N_v$-many sources and $\mathbf{W}$ is a $D \times M$ matrix of model weights which parameterize the mapping $\psi$.
+As Equation~\ref{eqn:data-space-distribution} indicates, the precision parameter corresponds to a standard deviation of $\sqrt{\beta^{-1}}$.
 
@@
 Finally, the precision parameter $\beta$ is initialized to the variance of the $(N_v+1)$-th principal component.
+The matrix $\mathbf{\Phi}$ is pre-computed and stored for reuse during each EM cycle together with the data matrix $\mathbf{X}$.
 After initialization, the expectation and maximization steps are repeated in turn until $Q$ converges to a predetermined tolerance level.
-For large $N_v$ we note that generating a regular grid within the simplex becomes cumbersome as the number of grid nodes scales as ${k + N_v - 2 \choose N_v - 1}$ for $k$ nodes per edge.
+For large $N_v$ we note that generating a regular grid within the simplex becomes cumbersome as the number of grid nodes scales as ${k + N_v - 2 \choose N_v - 1}$ for $k$ nodes per edge leading to a large responsibility matrix.
 An alternative approach is to randomly sample points within the simplex to obtain a total of $K$ nodes.
@@
 This serves as an important limiting case since linearly mixed spectra should not lead to the spurous introduction of non-linear contributions by the GSM.
-The goal of this first test is therefore to demonstrate that the GSM drives non-linear weights to zero for linearly mixed data while providing a fair test to compare GSM to well-established linear mixing models.
+The goal of this first test is therefore to demonstrate that the GSM drives non-linear weights to zero for linearly mixed data while providing a fair test to compare the GSM to a well-established linear mixing model.
+This ability clealy distinguishes the GSM from other non-nonlinear unmixing approaches such as autoencoders, which by their design, include non-linear mixing even when it is not present in the underlying data.
 To this end, a synthetic data set comprising linear mixtures of three sample spectra from the U.S.
@@
 Each HSI pixel included 462 wavelength bins ranging from $391$ to $1011$ nm.
-To evaluate the ability of GSM to identify potential contaminant sources, rhodamine dye, a commonly used tracer in hydrological studies, was released into a pond.
+To evaluate the ability of GSM to identify potential contaminant sources, rhodamine dye, a commonly used tracer in hydrological studies, was released into the pond.
 Two UAV flights spaced 15 minutes apart were used to capture the evolution of the plume as it dispersed into the surrounding water.
 
-The hyperspectral imager is in a pushbroom configuration so that HSI was captured one scan line at a time.
+The hyperspectral imager is in a pushbroom configuration so that each HSI was captured one scan line at a time.
 Data from an embedded GPS/INS unit enable direct georectification of captured imagery using the method described in \cite{muller2002program}.
@@
 The SNR of $20$ added to this example corresponds to zero-mean Gaussian noise with a standard deviation of $\sigma=0.0493$.
-After training, the GSM found $\sqrt{\beta^{-1}}=0.0495$ that accurately captures the introduced noise.
-This ability to assess the spectral variability of extracted endmembers is a key advantage of the GSM.
+After training, the GSM found $\sqrt{\beta^{-1}}=0.0495$, accurately identifying this introduced noise.
+The ability to assess the spectral variability of extracted endmembers is a key advantage of the GSM resulting from its probabilistic formulation.
 
@@
 This can be mitigated by instead randomly sampling points within the interior of the simplex using a uniform Dirichlet distribution to obtain a pre-determined number of nodes across the latent space.
-As the mixing coefficients $\pi_k$ are adapted during training, the variability in the spacing of the internodes should not significantly affect the performance of the model.
-In fact, this was confirmed for the simulated data set as summarized in Figure~\ref{fig:usgs-fits}.
-For considerably large data sets, the size of the responsibility matrix $\mathbf{R}$ can also lead to extended training times.
-This can be addressed by augmenting the EM procedure to updated using mini-batches of training samples during each E-step as described by Bishop et al.
+As the mixing coefficients $\pi_k$ are adapted during training, variability in spacing between nodes should not significantly affect the performance of the model.
+This was confirmed for the simulated data set as shown in Figure~\ref{fig:usgs-fits}.
+In terms of computational efficiency, each expectation step involves $\mathcal{O}(K \times N)$ operations to update the entries of the responsibility matrix leading to extended training times for considerably large data sets.
+This can be addressed by augmenting the EM procedure to use mini-batches of training samples as outlined by Bishop et al.
 for the GTM in ref.
 \cite{gtm-developments}.
-Rather than updating the full responsibility matrix, a subset of $\mathbf{R}$ corresponding to a single batch of training data can be evaluated with all other entries kept constant.
-The GSM may also be extended in other ways, for example, by replacing the precision parameter $\beta$ with a covariance matrix to model wavelength-dependent spectral variability common to many hyperspectral imaging platforms.
+Rather than updating the full responsibility matrix during each iteration, a subset of $\mathbf{R}$ corresponding to a single batch of training data can be evaluated with all other entries kept constant.
+The GSM may also be extended in other ways, for example, by replacing the precision parameter $\beta$ with a vector $\beta_{\labmda}$ of parameters to model wavelength-dependent spectral variability common to many hyperspectral imaging platforms.
 
